---
title: "Use of Code Files"
output:
  html_document: default
  pdf_document:
    includes:
      in_header: "preamble.tex"
    latex_engine: xelatex
---
```{css, echo=F}
.colbox {
  padding: 1em;
  background: white;
  color: black;
  border: 2px solid orange;
  border-radius: 10px;
}
.center {
  text-align: left;
}
}
```


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Packages

The following are, for each chapter, the packages required, with any
dependencies, to evaluate all code.
Additionally, knitr and Hmisc are required in order to process
the Rmd source file.
 
 The lattice package, which is also a dependency for latticeExtra, gets 
 some use in all chapters
 
* Ch1: latticeExtra (lattice is a dependency); DAAG car MASS AICcmodavg
 BayesFactor boot MPV ggplot2 tidyr statsr
* Ch2: DAAG MASS qra investr HistData BHH2 xtable BayesFactor boot zoo
 boot MCMCpack
* Ch3: DAAG car MASS AICcmodavg leaps BayesFactor splin es
* Ch4: DAAG effects mgcv splines scam MASS latticeExtra car WDI
 AICcmodavg ggplot2 kableExtra qgam
* Ch5: DAAG car mgcv colorspace HistData
* Ch6: DAAG ggsci latticeExtra ggplot2 mice car forecast mgcv tseries
* Ch7: DAAG lme4 afex MASS utils devtools qra glmmTMB DHARMa MEMSS
 forecast splines gamlss plotrix nlme
* Ch 8: DAAG latticeExtra plot rpart rpart.plot MASS ggplot2 car
 randomForest, bayesreg
*Ch9: DAAG MASS RColorBrewer teigen BiocManager DAAGbio hddplot lmtest
 splines cobalt mice datasets car micemd oz randomForest ggplot2
 latticeExtra mvtnorm teigen limma hddplot mgcv MatchIt sandwich
 gridExtra DAAGbio mlbench (exercise)
* Appendix A: DAAG dplyr tidyr tibble MASS gplots plotrix latticeExtra
 RColorBrewer

Should a package not be available on CRAN, it will in many instances
be possible to obtain the latest development version from a GitHub site.

# Alternative outputs and themes for the R markdown (`.Rmd`) files

Primarily, it is expected that users will run individual code chunks
one or more at a time as they work through the text.  In this mode
of working, needed R packages can be installed as required, as 
calculations proceed.

# Options and issues for processing a complete R qmd file

An alternatives is to process the whole file at one time, either by supplying the file name as argument to `knitr()`, or by clicking ‘knit’ in Rmarkdown with the file open.  Set the output option in the YAML header at the start of the file as required, to create HTML or PDF or Word output.  Additionally, the Markdown file that is created as an intermediary can be saved, if required.  An example of YAML settings is:

```
---
title: "Use of Code Files"
author: "John Maindonald"
date: "`r Sys.Date()`"
output: 
  html_document
---
```

Some available alternatives for the final three lines include:

```
output:
  word_document: default
  html_document: default
  pdf_document: default
---
```

```
output: rmarkdown::html_vignette
---
```


```{r}
library(knitr)
opts_chunk[['set']](cache.path='cache-', out.width="80%", fig.width=6, fig.height=6, fig.align='center', size="small", ps=10, strip.white = TRUE,
                    comment=NA, width=70, tidy.opts = list(replace.assign=FALSE))
```

```{r CodeControl, echo=FALSE}
xtras=FALSE
## xtras <- TRUE
library(knitr)
## opts_chunk[['set']](results="asis")
opts_chunk[['set']](eval=F)
## opts_chunk[['set']](eval=T)
```

## Code markup

The following shows an approach to placing graphs side by side
```{r, fig.width=4, fig.height=4,fig.show="hold", fig.align="default", out.width="48.5%"}
z <- seq(-4,4,length=101)
plot(z, dnorm(z), type="l", ylab="Normal density")
plot(z, dt(z, df=5), type="l", ylab="t-statistic density, 5 df")
```

## Using the `Hmisc::knitrSet()` function

This function creates a setup where 

* Space around the graph can be adjusted by setting the chunk 
options `bot`, `left`, `top`, and `rt`;
* Commonly used base graphics parameters can be supplied as chunk options.
These include `lwd`, `mgp`, `las`, `tcl`, `axes`, `xpd`, `mfg`.
* Default settings can be supplied for a number of chunk options.

See `?Hmisc::knitrSet` for details.  With R markdown files, the setting
`fig.align="default"` is required.  Either do not change from the
default globally, or accompany `fig.show="hold"` whenever used with
`fig.align="default"`, thus:

```{r, fig.width=4, fig.height=4,fig.show="hold", fig.align="default", out.width="48.5%"}
Hmisc::knitrSet(basename="tmp", lang='markdown', fig.path="figs/g", w=6, h=6)
z <- seq(-4,4,length=101)
plot(z, dnorm(z), type="l", ylab="Normal density")
plot(z, dt(z, df=5), type="l", ylab="t-statistic density, 5 df")
```

```{r, eval=T}
if(file.exists("/Users/johnm1/pkgs/PGRcode/inst/doc/")){
code <- knitr::knit_code$get()
txt <- paste0("\n## ", names(code),"\n", sapply(code, paste, collapse='\n'))
writeLines(txt, con="/Users/johnm1/pkgs/PGRcode/inst/doc/UsingCode.R")
}
```
