---
title: "Chapter 4: Exploiting the linear model framework"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Ch4-Exploiting}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

On options for working with the code see the vignettes
[Ch1-Learning](PGRcode/Ch1-Learning.html) and
[UsingCode](PGRcode/UsingCode.html).

```{r CodeControl, echo=FALSE}
options(rmarkdown.html_vignette.check_title = FALSE)
## xtras=FALSE
xtras <- F
library(knitr)
## opts_chunk[['set']](results="asis")
## opts_chunk[['set']](eval=T)
opts_chunk[['set']](eval=F)
```

##### Packages required (with dependencies)
DAAG effects mgcv splines scam MASS latticeExtra car WDI AICcmodavg ggplot2 kableExtra qgam

Additionally, Hmisc and knitr are required in order to process the Rmd source file. 

Note the use of the 'patchwork' package to make it easy
to place two ggplot2 plots side by side.

```{r setup, cache=FALSE}
Hmisc::knitrSet(basename="exploit", lang='markdown', fig.path="figs/g", w=7, h=7)
oldopt <- options(digits=4, width=70, scipen=999)
library(knitr)
## knitr::render_listings()
opts_chunk[['set']](cache.path='cache-', out.width="80%", fig.align="center", 
                    fig.show='hold', formatR.arrow=FALSE, ps=10, 
                    strip.white = TRUE, comment=NA, width=70, 
                    tidy.opts = list(replace.assign=FALSE))
```

### Section 4.1 Levels of a factor -- using indicator variables
#### Subsection 4.1.1: Example -- sugar weight

```{r D1_1a}
sugar <- DAAG::sugar  # Copy dataset 'sugar' into the workspace
## Ensure that "Control" is the first level
sugar[["trt"]] <- relevel(sugar[["trt"]], ref="Control")
options()[["contrasts"]]  # Check the default factor contrasts
## If your output does not agree with the above, then enter
## options(contrasts=c("contr.treatment", "contr.poly"))
```

```{r 4.1, w=3.25, h=2.4, echo=FALSE, out.width="50%"}
gph <- lattice::stripplot(trt ~ weight, pch=4, data=sugar,
  xlab=list("Weight (mg)", cex=0.9))
  bw10 <- list(fontsize=list(text=10, points=5))
update(gph, aspect=0.5, par.settings=bw10)
```

```{r D1_1c}
sugar.aov <- aov(weight ~ trt, data=sugar)
## To display the model matrix, enter: model.matrix(sugar.aov)
## Note the use of summary.lm(), not summary() or summary.aov()
round(signif(coef(summary.lm(sugar.aov)), 3), 4)
```

```{r D1_1d}
sem <- summary.lm(sugar.aov)$sigma/sqrt(3)  # 3 results/trt
# Alternatively, sem <- 6.33/sqrt(2)
qtukey(p=.95, nmeans=4, df=8) * sem
```

#### Subsection 4.1.2: Different choices for the model matrix when there are factors

```{r D1_2a}
contrasts(sugar$trt) <- 'contr.sum'
sugarSum.aov <- aov(weight ~ trt, data = sugar)
round(signif(coef(summary.lm(sugarSum.aov)), 3),4)
```

```{r D1_2b, echo=T}
dummy.coef(sugarSum.aov)
```

#####                Factor contrasts -- further details

```{r D1_2c}
contrasts(sugar$trt) <- "contr.sum"
```

```{r D1_2d}
fish <- factor(1:3, labels=c("Trout","Cod","Perch"))
```

```{r D1_2e}
contr.treatment(fish)
# Base is "Trout"
```

```{r D1_2f}
contr.SAS(fish)
# Base is "Perch"
```

```{r D1_2g}
contr.sum(fish)
# Base is mean of levels
```

#####  *Tests for main effects in the presence of interactions?

### Section 4.2 Block designs and balanced incomplete block designs
#### Subsection 4.2.1: Analysis of the rice data, allowing for block effects

```{r D2_1a}
rice <- DAAG::rice
ricebl.aov <- aov(ShootDryMass ~ Block + variety * fert, data=rice)
print(summary(ricebl.aov), digits=3)
```

```{r D2_1b}
round(signif(coef(summary.lm(ricebl.aov)), 3), 5)
with(summary.lm(ricebl.aov),
cat("Residual standard error: ", sigma, "on", df[2], "degrees of freedom"))
```

```{r D2_1c, size='normalsize', results='asis'}
## AOV calculations, ignoring block effects
rice.aov <- aov(ShootDryMass ~ variety * fert, data=rice)
summary.lm(rice.aov)$sigma
```

```{r D2_1d}
ricebl.aov <- aov(ShootDryMass ~ factor(Block) + variety * fert, data=rice)
```

```{r D2_1e}
model.tables(ricebl.aov, type="means", se=TRUE, cterms="variety:fert")
```

#### Subsection 4.2.2: A balanced incomplete block design

```{r D2_2a}
appletaste <- DAAG::appletaste
with(appletaste, table(product, panelist))
```

```{r D2_2b}
sapply(appletaste, is.factor)  # panelist & product are factors
appletaste.aov <- aov(aftertaste ~ product + panelist, data=appletaste)
summary(appletaste.aov)
```

```{r 4_2, w=6, h=2.0, left=-1, cex.lab=0.8, echo=FALSE, ps=9, mfrow=c(1,2), out.width="90%"}
termplot(appletaste.aov, partial=TRUE, col.res="black", fg='gray',
         cex=0.8, mgp=c(1.5,0.4,0))
 # Specify `partial=TRUE` to show partial residuals
```

```{r D2_2e}
as.data.frame(effects::Effect("product", appletaste.aov, confidence.level=0.95))
```

```{r D2_2f}
## NB that 'product' was first term in the model formula
## Thus, the 1st 4 coefficients have the information required
coef(summary.lm(appletaste.aov))[1:4, ]
```

### Section 4.3 Fitting multiple lines

```{r D3a}
## Fit various models to columns of data frame leaftemp (DAAG)
leaftemp <- DAAG::leaftemp
leaf.lm1 <- lm(tempDiff ~ 1 , data = leaftemp)
leaf.lm2 <- lm(tempDiff ~ vapPress, data = leaftemp)
leaf.lm3 <- lm(tempDiff ~ CO2level + vapPress, data = leaftemp)
leaf.lm4 <- lm(tempDiff ~ CO2level + vapPress +
               vapPress:CO2level, data = leaftemp)
```

```{r D3b}
anova(leaf.lm1, leaf.lm2, leaf.lm3, leaf.lm4)
```

```{r 4_3, echo=FALSE, w=6, h=2.8, left=-1, top=2.5, lwd=0.5, cex.lab=0.9, ps=8, mgp=c(1.8, 0.5,0), out.width="100%"}
options(contrasts = c("contr.treatment", "contr.poly"))
cex.eq <- 0.85
yran <- with(leaftemp, range(tempDiff))
yran[2] <- yran[2] + diff(yran) * 0.08
par(fig=c(0, 0.39, 0, 0.9))
plot(tempDiff~vapPress, xlab = "Vapour Pressure", data=leaftemp,
ylab = "Temperature difference", ylim = yran, fg="gray", tcl=-0.25,
pch = as.numeric(CO2level), cex=0.5, cex.axis=0.8, col="black")
mtext(side = 3, line = 1.5, "A: Single line", adj = 0, cex=1.15)
topleft <- par()$usr[c(1, 4)] + c(cex.eq, -cex.eq) * par()$cxy
chh <- par()$cxy[2]*0.5
ab1 <- leaf.lm2$coef
mtext(side=3,line=0.72,
paste("tempDiff =", round(ab1[1], 2), round(ab1[2], 2),
" x vapPress",sep = ""), adj=0, col="black", cex=cex.eq)
abline(ab1[1], ab1[2])
par(fig=c(0.3, 0.69, 0, 0.9), new=T)
plot(tempDiff~vapPress, data=leaftemp, xlab = "Vapour Pressure",
ylab = "", ylim = yran, pch = as.numeric(CO2level), tcl=-0.25,
cex=0.5, cex.axis=0.8, fg="gray")
mtext(side = 3, line = 1.5, "B: Parallel lines", adj = 0, cex=1.15)
a123 <- c(leaf.lm3$coef[1], sum(leaf.lm3$coef[1:2]),
sum(leaf.lm3$coef[c(1,3)]))
b1 <- leaf.lm3$coef[4]
mtext(side=3,line=.72,
paste("Intercepts are:", paste(round(a123,2), collapse=", ")),
adj = 0, col = "black", cex = cex.eq)
mtext(side=3,line=0, paste("Slope is:", round(b1, 2), sep = " "),
adj = 0, col = "black", cex = cex.eq)
ran <- with(leaftemp, sapply(split(vapPress, CO2level), range))
for(i in 1:3){
yi <- a123[i] + b1 * ran[,i]
lines(ran[,i], yi, lty = c(4,5,7)[i], lwd = 1, col = "black")
}
par(fig=c(0.6, 0.99, 0, 0.9), new=T)
plot(tempDiff~vapPress, data=leaftemp, xlab = "Vapour Pressure",
ylab = "",  ylim = yran,  pch = as.numeric(CO2level), tcl=-0.25,
cex=0.5, cex.axis=0.8, fg="gray")
mtext(side = 3, line = 1.5, "C: Separate lines", adj = 0, cex=1.15)
a123 <- c(leaf.lm4$coef[1], sum(leaf.lm4$coef[1:2]),
sum(leaf.lm4$coef[c(1,3)]))
b123 <- c(leaf.lm4$coef[4], sum(leaf.lm4$coef[4:5]),
sum(leaf.lm4$coef[c(4,6)]))
mtext(side=3,line=.72,
paste("Intercepts are:", paste(round(a123,2), collapse=", ")),
adj = 0, col = "black", cex = cex.eq)
mtext(side=3,line=0,
paste("Slopes are:", paste(round(b123,2), collapse=", ")),
adj=0, cex=cex.eq)
for(i in 1:3){
a <- a123[i]
yi <- a123[i] + b123[i] * ran[,i]
lines(ran[,i], yi, lty = c(4,5,7)[i], lwd = 1, col = "black")
}
par(fig=c(0, 1, 0, 1),new=T, mar=rep(0,4))
plot(0:1, 0:1, bty="n", axes=F, xlab="", ylab="", type="n")
legend(0.55, 0.985, legend=c("low","medium","high"), lty=c(4,5,7), col="black",
pch=1:3, xjust=0.5, yjust=1, bty="n", pt.cex=0.72, ncol=3,
text.width=0.1, cex=0.85)
```

```{r 4_4, echo=FALSE, w=7.25, h=1.65, mgp=c(1.8,0.5,0), top=1, left=-0.5, ps=9, mfrow=c(1,4), out.width="100%"}
plot(leaf.lm3, cook.levels=0.12, caption=c('A: Resids vs Fitted', 'B: Normal Q-Q',
'C: Scale-Location', '', 'D: Resids vs Leverage'), cex.caption=0.85, fg="gray")
```

```{r D3e}
print(coef(summary(leaf.lm3)), digits=2)
```

### Section 4.4 Methods for fitting smooth curves

#### Subsection 4.4.1: Polynomial Regression

```{r D4_1a, warning=FALSE}
seedrates <- DAAG::seedrates
form2 <- grain ~ rate + I(rate^2)
# Without the wrapper function I(), rate^2 would be interpreted
# as the model formula term rate:rate, and hence as rate.
quad.lm2 <- lm(form2, data = seedrates)
## Alternative, using gam()
## quad.gam <- mgcv::gam(form2, data = seedrates)
```

```{r D4_1b}
suppressPackageStartupMessages(library(ggplot2))
```

```{r 4_5, w=7.2, h=3.0, left=1, lwd=0.5, tcl=-0.35, echo=FALSE, out.width="100%"}
## Use ggplot2 functions to plot points, line, curve, & 95% CIs
## library(ggplot2)
gph <- ggplot(DAAG::seedrates, aes(rate,grain))+
  geom_point(aes(size=3), color='magenta')+xlim(c(25,185))
colors <- c("Linear"="blue", "Quadratic"="red")
ggdat <- ggplot_build(gph+geom_smooth(aes(rate,grain,color="Linear"),
  method=lm, formula=y~poly(x,2),fullrange=TRUE))$data[[2]]
gph1 <- gph+geom_smooth(aes(color="Linear"), method=lm, formula=y~x, fullrange=TRUE, fill='dodgerblue')
gph1 +  geom_line(data = ggdat, aes(x = x, y = y, color="Quadratic"),
                  linewidth=0.75)+
  geom_ribbon(data=ggdat, aes(x=x,y=y, ymin=ymin, ymax=ymax,
                              color="Quadratic"), linewidth=0.75,
  fill=NA, linetype=2, outline.type='both', show.legend=FALSE) +
  scale_color_manual(values=colors, aesthetics = "color")+
  theme(legend.position=c(.8,.78)) +
  coord_cartesian(expand=FALSE) + xlab("Seeding rate (kg/ha)") +
    ylab("Grains per head") + labs(color="Model") +
  guides(size='none',
         color = guide_legend(override.aes = list(fill="transparent") ) )
## detach("package:ggplot2")
```

```{r 4_5, eval=F}
```

```{r D4_1d, results="hold"}
quad.lm2 <- lm(grain ~ rate + I(rate^2), data = DAAG::seedrates)
print(coef(summary(quad.lm2)), digits=2)
cat("\nCorrelation matrix\n")
print(summary(quad.lm2, corr=TRUE)$correlation, digits=2)
```

#####      *An alternative formulation using orthogonal polynomials

```{r D4_1e}
seedratesP.lm2 <- lm(grain ~ poly(rate,2), data = seedrates)
print(coef(summary(seedratesP.lm2)), digits=2)
```

```{r D4_1f}
## Alternative, using mgcv::gam()
seedratesP.gam <- mgcv::gam(grain ~ poly(rate,2), data = seedrates)
```

```{r D4_2a}
logseed.lm <- lm(log(grain) ~ log(rate), data=DAAG::seedrates)
coef(summary(logseed.lm))
```

```{r 4_6, w=7, h=2.8, left=2, rt=2, fig.pos='ht', echo=FALSE, out.width="100%"}
## Use ggplot2 functions to plot points, line, curve, & 95% CIs
## library(ggplot2)
gph <- ggplot(DAAG::seedrates, aes(rate,grain)) +
  geom_point(size=3, color="magenta")+xlim(c(25,185))
colors <- c("Loglinear"="gray40", "Quadratic"="red")
ggdat <- ggplot_build(gph+geom_smooth(method=lm, formula=y~poly(x,2),
                                      fullrange=TRUE))$data[[2]]
ggln <- ggplot_build(gph+geom_smooth(method=lm,
                        formula=log(y)~log(x),fullrange=TRUE))$data[[2]]
## Assign to gphA rather than (as in text) plotting at this point
gphA <- gph +  geom_line(data = ggdat, aes(x = x, y = y, color="Quadratic"),
                 linewidth=0.75) +
geom_ribbon(data=ggdat, aes(x=x,y=y, ymin=ymin, ymax=ymax, color="Quadratic"),
            linewidth=0.75, fill=NA, linetype=2, outline.type='both',
            show.legend=FALSE) +
geom_line(data = ggln, aes(x = x, y = exp(y), color="Loglinear"),
          linewidth = 0.75) +
geom_ribbon(data=ggln, aes(x=x,y=exp(y), ymin=exp(ymin), ymax=exp(ymax),
            color="Loglinear"), fill=NA, linewidth=0.75, linetype=3,
            outline.type='both', show.legend=FALSE)+
  scale_color_manual(values=colors, aesthetics = "color")+
  coord_cartesian(expand=FALSE) +
  xlab("Seeding rate (kg/ha)") + ylab("Grains per head") +
  labs(title="A: Loglinear fit vs quadratic fit", color="Model") +
  guides(size='none',
         color = guide_legend(override.aes = list(fill="transparent") ) ) +
  theme(legend.position=c(.8,.78))
df <- data.frame(rate=rep(DAAG::seedrates$rate,2), res=c(resid(logseed.lm),
  log(DAAG::seedrates$grain)-log(fitted(quad.lm2))),
  Model=rep(c("Loglinear","Quadratic"),rep(nrow(DAAG::seedrates),2)))
## Assign to gphB rather than (as in text) plotting at this point
gphB <- ggplot(df, aes(x=rate, y=res, shape=Model,color=Model))+
geom_point(size=2.5) + scale_color_manual(values=colors) +
xlab("Seeding rate (kg/ha)") + ylab("Residuals on log scale") +
labs(title="B: Residuals") +
  guides(size='none',
         color = guide_legend(override.aes = list(fill="transparent") ) ) +
  theme(legend.position=c(.8,.78))
## Now take advantage of the magic of the 'patchwork' package
library(patchwork)
gphA+gphB
## detach("package:ggplot2")
```

```{r 4_6, eval=F}
```

```{r D4_2c}
aic <- AIC(quad.lm2, logseed.lm)
aic["logseed.lm",2] <- aic["logseed.lm",2] + sum(2*log(seedrates$grain))
round(aic,1)
```

```{r D4_3a}
seedrates<-DAAG::seedrates
quad.lm2 <- lm(grain ~ poly(rate,degree=2), data=seedrates)
ns.lm2 <- lm(grain ~ splines::ns(rate,df=2), data=seedrates)
tps.gam2 <- mgcv::gam(grain ~ s(rate, k=3, fx=T), data=seedrates)
```

```{r D4_3b}
mflist <- lapply(list(quad=quad.lm2, nsplines=ns.lm2, tps=tps.gam2), model.matrix)
mftab <- with(mflist, cbind(quad, nsplines, tps))
colnames(mftab) <- c("(Int)", "poly2.1", "poly2.2", "(Int)", "ns2.1", "ns2.2", "(Int)", "s3.1", "s3.2")
library(kableExtra)
linesep = c('', '', '', '\\addlinespace')
kbl(mftab, booktabs=TRUE, format='latex', toprule=FALSE,
format.args=list(justify="right", width=8)) |>
kable_styling(latex_options = c("scale_down",latex_options = "hold_position"), position='center') |>
add_header_above(c('poly(rate,2)' = 3, 'splines::ns(rate,df=2)' = 3, 's(rate, k=3, fx=T)' = 3),
align='c', monospace=rep(T,3))|>
add_header_above(c('lm: grain~' = 3, 'lm: grain~'=3, 'gam: grain~'=3),
                 align='c', monospace=rep(T,3), line=F)
```

#####          GAM models versus models fitted using lm()
#####            Alternative fits -- what is the best choice?
```{r D4_3c}
## Load required packages
suppressPackageStartupMessages(library(splines))
suppressPackageStartupMessages(library(mgcv))
```

```{r 4_7, echo=FALSE, mfrow=c(1,2), mgp=c(2,0.5,0), w=6, h=2.7, left=-2.5, top=2, lwd=0.5, out.width="100%"}
library(mgcv)
fruitohms <- within(DAAG::fruitohms, kohms <- ohms/1000)
## Panel A: 3, 4 and 5 d.f. tprs (cubic spline fits are almost identical)
plot(kohms ~ juice, data=fruitohms, ylim=c(0, max(kohms)*1.1), fg="gray")
ohms.tp3 <- gam(kohms~s(juice, bs="tp", k=3, fx=T), data=fruitohms)
ohms.tp4 <- gam(kohms~s(juice, bs="tp", k=4, fx=T), data=fruitohms)
ohms.tp5 <- gam(kohms~s(juice, k=5, fx=T), data=fruitohms)
lines(fitted(ohms.tp4) ~ juice, data=fruitohms, col=2)
lines(fitted(ohms.tp4) ~ juice, data=fruitohms, col=2, lty=2, lwd=2)
lines(fitted(ohms.tp5) ~ juice, data=fruitohms, col=2, lty=3, lwd=2)
legend("topright", title="Comparison of fits", lty=c(1,2,3), lwd=c(1,2,2), col=c(2,2,2),
cex=0.8, legend=c("s(juice, k=3, fx=TRUE))", "s(juice, k=4, fx=TRUE)",
                   "s(juice, k=5, fx=TRUE)"))
mtext(side=3, line=0.75, "A: 3 vs 4 vs 5 d.f. thin plate spline", adj=0, cex=1.25)
## Panel B: Penalized spline fits
plot(kohms ~ juice, data=fruitohms, ylim=c(0, max(kohms)*1.1),
     ylab=" ", fg="gray")
## Fits with automatic choice of smoothing parameter
ohms.tp <- gam(kohms~s(juice, bs="tp"), data=fruitohms)
ohms.tpML <- gam(kohms ~ s(juice, bs="tp"), data=fruitohms, method="ML")
ohms.tpBIC <- gam(kohms ~ s(juice, bs="tp"), data=fruitohms,
                  gamma=log(nrow(fruitohms))/2, method="ML")
lines(fitted(ohms.tp) ~ juice, data=fruitohms, col=2)
lines(fitted(ohms.tpML) ~ juice, data=fruitohms, col=2, lty=2, lwd=2)
lines(fitted(ohms.tpBIC) ~ juice, data=fruitohms, col=2, lty=3, lwd=2)
legend("topright", title="Penalized regression spline fits", cex=0.8,
legend=c('method="GCV.Cp"','method="ML"', 'gamma=log(n)/2'),
         lty=1:3, col=c(2,2,2), lwd=c(1,2,2))
mtext(side=3, line=0.75, "B: Penalized regression spline fits", adj=0, cex=1.25)
```

```{r D4_3e, echo=T}
ohms.tp <- gam(kohms~s(juice, bs="tp"), data=fruitohms)
ohms.cs <- gam(kohms~s(juice, bs="cs"), data=fruitohms)
range(fitted(ohms.tp)-fitted(ohms.cs))
```

```{r D4_3f, echo=T}
summary(ohms.tp)
```

```{r D4_3g}
summary(ohms.tpBIC)
```

#### Subsection 4.4.3: The contributions of basis curves to the fit

```{r 4_8, echo=FALSE, w=6.5, h=3.75, out.width="90%"}
## k=3 gives 3 basis terms; k=4 gives 4 basis terms
ohms.tp3 <- gam(kohms ~ s(juice, k=3), data=fruitohms)
ohms.tp4 <- gam(kohms ~ s(juice, k=4), data=fruitohms)
## Reorder to constant term first, then linear term, then ...
matohmstp3 <- model.matrix(ohms.tp3)[, c(1,3:2)]
matohmstp4 <- model.matrix(ohms.tp4)[, c(1,4:2)]
m <- dim(matohmstp4)[1]
longdf1 <- data.frame(juice=rep(fruitohms[, "juice"],4),
  basis2 = c(as.vector(matohmstp3),rep(NA,m)), basis3 = as.vector(matohmstp4),
  gp = factor(rep(c("Intercept", paste("tpSpline",1:3, sep="")), rep(m,4))))
gph1 <- lattice::xyplot(basis3 ~ juice | gp, data=longdf1, layout=c(1,4),
  ylab="Basis terms", strip=FALSE,
  panel=function(x,y,subscripts){
    lattice::llines(smooth.spline(x,y))
    y2 <- longdf1$basis2[subscripts]
    if(!any(is.na(y2))) lattice::llines(smooth.spline(x,y2),lty=1)},
  outer=TRUE)
  b2 <- coef(lm(kohms ~ 0+matohmstp3, data=fruitohms))
  b3 <- coef(lm(kohms ~ 0+matohmstp4, data=fruitohms))
  spline2 <- as.vector(sweep(matohmstp3, 2, b2, "*"))
  spline3 <- as.vector(sweep(matohmstp4, 2, b3, "*"))
  longdf2 <- data.frame(juice=rep(fruitohms[, "juice"],4),
  spline2 = c(spline2, rep(NA,m)), spline3=spline3,
    gp = factor(rep(c("Intercept", paste("spline",1:3, sep="")), rep(m,4))) )
    ## yran <- range(c(spline2, spline3))
  yran <- c(-6,8.5)
gph2 <- lattice::xyplot(spline3 ~ juice | gp, data=longdf2, layout=c(1,4),
                        scales=list(y=list(at=c(-4, 0, 4,8))), ylim=yran,
  ylab="Add these contributions (kilo-ohms)", strip=FALSE,
  panel=function(x,y,subscripts){
    lattice::llines(smooth.spline(x,y))
    y2 <- longdf2$spline2[subscripts]
    if(!any(is.na(y2)))
      lattice::llines(smooth.spline(x,y2), lty=2, lwd=2, col="red",
                      alpha=0.5)}, outer=TRUE)
maint <- c(A="A: Basis functions, tprs", B="B: Contribution to fitted curve")
bw10 <- list(fontsize=list(text=10, points=5), lty=c(2,2,1,1))
striplabels <- list(
  A=c("Constant","Basis 1","Basis 2","Basis 3"),
  B=c("Constant","Add 1","Add 2","Add 3"))
stripfun <- function(factor.levels)
  lattice::strip.custom(strip.names=TRUE, var.name="", sep=expression(""),
                        factor.levels=factor.levels)
print(update(gph1, par.settings=bw10,
strip.left=stripfun(factor.levels=striplabels[["A"]]),
  main=list(maint["A"], y=-0.5)), position=c(0,0,.5,1))
print(update(gph2, par.settings=bw10,
  strip.left=stripfun(factor.levels=striplabels[["B"]]),
  main=list(maint["B"], y=-0.5)), position=c(.5,0,1,1), newpage=FALSE)
```

#### Subsection 4.4.4: Checks on the fitted model

```{r 4_9, w=6, h=5, mfrow=c(2,2), tcl=-0.5, left=-2, top=1, mar=c(2.5,2.5,2.1,.5), col=adjustcolor('black', 0.4), fg="gray", echo=FALSE, ps=9, out.width="100%"}
out <- capture.output(gam.check(ohms.tpBIC, tcl=-0.5))
```

```{r D4_4c}
## Printed output from `gam.check(ohms.tpBIC)`
cat(out, sep="\n")
```

#### Subsection 4.3.5: Monotone curves

```{r 4_10, echo=FALSE, mfrow=c(1,3), mgp=c(2,0.5,0), w=7.2, h=2.4, left=-2.0, top=1, lwd=0.5, out.width="100%", fig.pos="t"}
ohms.scam <- scam::scam(kohms ~ s(juice,bs="mpd"), data=fruitohms)
plot(ohms.scam, resid=T, pch=1, shift=mean(predict(ohms.scam)),
     xlab="Apparent juice content (%)", ylab="Resistance (kohms)",
     fg="gray", shade=T, shade.col=adjustcolor('blue',0.4))
mtext(side=3, line=0.5, "A: Fitted monotone curve", cex=1.25, adj=0)
qqnorm(resid(ohms.scam), fg="gray", main="")
qqline(resid(ohms.scam), col=2)
mtext(side=3, line=0.5, "B: Normal Q-Q plot of residuals", cex=1.25, adj=0)
s2 <- smooth.spline(fruitohms$juice, resid(ohms.scam))
s2p <- update(s2, penalty=log(nrow(fruitohms))/2)
plot(s2$x, s2$yin, fg="gray", xlab="Apparent juice content", ylab="Residual")
lines(s2$x, s2$y, col=2)
lines(s2p$x, s2p$y, col=2, lty=2, lwd=2)
mtext(side=3, line=0.5, "C: Residuals vs ... juice content", cex=1.25, adj=0)
```

```{r D4_5b}
ohms.scam <- scam::scam(kohms ~ s(juice,bs="mpd"), data=fruitohms)
summary(ohms.scam)
```

```{r D4_5}
AIC(ohms.scam, ohms.tp)
```

```{r D4_5d}
BIC(ohms.scam, ohms.tp)
```

#### Subsection 4.4.6: Different smooths for different levels of a factor

```{r D4_6a}
whiteside <- MASS::whiteside
gas.gam <- gam(Gas ~ Insul+s(Temp, by=Insul), data=whiteside)
```

```{r 4_11, echo=FALSE, warning=FALSE, w=6.5, h=3.5, bot=-1.0, left=1.5, cache=FALSE}
library(mgcv)
suppressPackageStartupMessages(library(latticeExtra, quietly=TRUE))
whiteside <- MASS::whiteside
gas.gam <- gam(Gas ~ Insul+s(Temp, by=Insul), data=whiteside)
lims <- range(whiteside$Temp)
lims <- lims+diff(range(lims))*c(-0.04, 0.04)
df2 <- expand.grid(Temp=seq(from=lims[1], to=lims[2], length.out=20),
Insul=factor(unique(levels(whiteside$Insul))))
df2[, c('fit','se')] <- predict(gas.gam, se=TRUE, newdata=df2)
df2 <- within(df2, {low <- fit-2*se; high<-fit+2*se})
parset <- DAAG::DAAGtheme(color=FALSE, pch=c(1,4), cex=1.5)
gph <- xyplot(Gas ~ Temp, data=whiteside, groups=Insul,
  xlab = list("Temperature (degrees Celsius)", cex=1.25),
  ylab = list("Gas usage (1000s of cubic ft", cex=1.25),
  xaxs="i", yaxs="i", aspect = 1,
  auto.key=list(points=TRUE, lines=TRUE, space="right", cex=1.25),
  scales=list(cex=1.0, tck=0.5, xlim = lims, ylim = c(-.3, 1)),
  par.settings=parset,
  main="Gas usage vs temperature, before/after insulation")

my.panel.bands <- function(x, y, upper, lower, fill, col,
subscripts, ..., font, fontface)
{
upper <- upper[subscripts]
lower <- lower[subscripts]
panel.polygon(c(x, rev(x)), c(upper, rev(lower)),
col = fill, border = FALSE,
...)
}
panel2 <- function(x, y, ...){
panel.superpose(x, y, panel.groups = my.panel.bands, type='l',...)
panel.xyplot(x, y, type='l', lwd=2, ...)
}

gph2 <- xyplot(fit~Temp, data=df2, groups=Insul, upper = df2$high, lower = df2$low,
               panel = panel2, fill=adjustcolor('gray40', 0.4), subscripts=T,
               scales=list(cex=1.0, tck=0.5, xlim=lims), par.settings=parset)

update(gph, xlim=lims)+latticeExtra::as.layer(gph2)
```

```{r D4_6c}
summary(gas.gam)
```

```{r D4_6d, eval=FALSE}
Box.test(resid(gas.gam)[whiteside$Insul=='Before'], lag=1)
Box.test(resid(gas.gam)[whiteside$Insul=='After'], lag=1)
```

#### Subsection 4.4.7: The remarkable reach of mgcv and related packages
#####              Departures from independence assumptions
#### Subsection 4.4.8: Multiple spline smoothing terms --- dewpoint data

```{r 4_12, echo=FALSE, w=4.5, h=2.0, left=-1, fig.pos='h', ps=9, mfrow=c(1,2), out.width="68%"}
## GAM model -- `dewpoint` data
dewpoint <- DAAG::dewpoint
ds.gam <- gam(dewpt ~ s(mintemp) + s(maxtemp), data=dewpoint)
plot(ds.gam, resid=TRUE, pch=".", se=2, cex=2, fg="gray")
```

```{r 4_12, eval=F}
```

#####        Using residuals as a check for non-additive effects

```{r 4_13, echo=FALSE, w=5, h=2.5, out.width="80%"}
library(lattice)
## Residuals vs maxtemp, for different mintemp ranges
mintempRange <- equal.count(dewpoint$mintemp, number=3)
ds.xy <- xyplot(residuals(ds.gam) ~ maxtemp|mintempRange, data=dewpoint,
                layout=c(3,1), scales=list(tck=0.5), aspect=1, cex=0.65,
                par.strip.text=list(cex=0.75), type=c("p","smooth"),
                xlab="Maximum temperature", ylab="Residual")
ds.xy
```

```{r 4_13, eval=F}
```

#####                         *A smooth surface

```{r D5_1c,  w=6, h=2.7, mfrow=c(1,2), out.width="100%"}
## Fit surface
ds.tp <- gam(dewpt ~ s(mintemp, maxtemp), data=DAAG::dewpoint)
vis.gam(ds.tp, plot.type="contour")   # gives a contour plot of the
# fitted regression surface
vis.gam(ds.gam, plot.type="contour")  # cf, model with 2 smooth terms
```

#### Subsection 4.4.9:  Atlantic hurricanes that made landfall in the US

```{r D5_2a}
hurricNamed <- DAAG::hurricNamed
hurricS.gam <- gam(car::yjPower(deaths, lambda=-0.2) ~
  s(log(BaseDam2014)) + s(LF.PressureMB),
  data=hurricNamed, method="ML")
anova(hurricS.gam)
```

```{r 4_14, echo=FALSE, w=6, h=1.95, bot=1, left=-1, top=1.5, fig.pos='b', ps=9, mgp=c(2,0.5,0), mfrow=c(1,3), out.width="100%"}
plot(hurricS.gam, resid=TRUE, pch=16, cex=0.5, select=1, fg="gray")
mtext(side=3, line=1, "A: Term in log(BaseDam2014)", cex=1.0, adj=0, at=-3.75)
plot(hurricS.gam, resid=TRUE, pch=16, cex=0.5, select=2, fg="gray")
mtext(side=3, line=1, "B: Term in LF.PressureMB", cex=1.0, adj=0, at=878)
qqnorm(resid(hurricS.gam), main="", fg="gray")
mtext(side=3, line=1, "C: Q-Q plot of residuals", cex=1.0, adj=0, at=-4.25)
```

```{r 4_14, eval=F}
```

#####  An explanatory variable with an overly long-tailed distribution

```{r D5_2c"100%"}
hurricSlog1.gam <- gam(log(deaths+1) ~ s(log(BaseDam2014)), data=hurricNamed)
hurricSlog2.gam <- gam(log(deaths+1) ~ s(BaseDam2014), data=hurricNamed)
```

```{r 4_15, echo=FALSE, w=6.5, h=3, left=-1, top=1.5, ps=10, fig.pos='b', mfrow=c(1,2), out.width="80%"}
plot(hurricSlog1.gam, resid=TRUE, pch=16, cex=0.5, adj=0, fg="gray")
mtext(side=3, "A: Use log(BaseDam2014)", cex=1.4, adj=0, line=1, at=-3.15)
plot(hurricSlog2.gam, resid=TRUE, pch=16, cex=0.5, fg="gray")
mtext(side=3, "B: Use BaseDam2014", cex=1.4, adj=0, line=1, at=-28500)
```

```{r 4_15, eval=F}
```

#### Subsection 4.4.10: Other smoothing methods

### Section 4.5 Quantile regression

```{r D5_3a}
## If necessary, install the 'WDI' package & download data
if(!file.exists("wdi.RData")){
  if(!is.element("WDI", installed.packages()[,1]) )install.packages("WDI")
inds <- c('SP.DYN.TFRT.IN','SP.DYN.LE00.IN', 'SP.POP.TOTL')
indnams <- c("FertilityRate", "LifeExpectancy", "population")
wdi2020 <- WDI::WDI(country="all", indicator=inds, start=2020, end=2020,
                    extra=TRUE)
wdi2020 <- na.omit(droplevels(subset(wdi2020, !region %in% "Aggregates")))
wdi <- setNames(wdi2020[order(wdi2020[, inds[1]]),inds], indnams)
save(wdi, file="wdi.RData")
}
```

#####       2020 World Bank data on fertility and life expectancy

```{r D5_3b}
load("wdi.RData")  # Needs `wdi.RData` in working directory; see footnote
library(qgam)
wdi[, "ppop"] <- with(wdi, population/sum(population))
wdi[,"logFert"] <- log(wdi[,"FertilityRate"])
form <- LifeExpectancy ~ s(logFert)
## Panel A model
fit.qgam <- qgam(form, data=wdi, qu=.5)
## Panel B: Multiple (10%, 90% quantiles; unweighted, then weighted
fit19.mqgam <- mqgam(form, data=wdi, qu=c(.1,.9))
wtd19.mqgam <- mqgam(form, data=wdi, qu=c(.1,.9),
                      argGam=list(weights=wdi[["ppop"]]))
```

```{r D5_3c}
hat50 <- cbind(LifeExpectancy=wdi[, "LifeExpectancy"], logFert=wdi[,"logFert"],
                as.data.frame(predict(fit.qgam, se=T)))
hat50 <- within(hat50, {lo <- fit-2*se.fit; hi <- fit+2*se.fit})
hat19 <- as.data.frame(matrix(nrow=nrow(wdi), ncol=4))
for(i in 1:2){hat19[[i]] <- qdo(fit19.mqgam, c(.1,.9)[i], predict)
              hat19[[i+2]] <- qdo(wtd19.mqgam, c(.1,.9)[i], predict) }
  ## NB, can replace `predict` by `plot`, or `summary`
colnames(hat19) <- c(paste0(rep(c('q','qwt'),c(2,2)), rep(c('10','90'),2)))
hat19 <- cbind(hat19, logFert=wdi[,"logFert"])
```

```{r 4_16, echo=FALSE, w=7.2, h=2.4, out.width="100%", warning=FALSE,  message=FALSE, mfrow=c(1,3)}
## Panel A: Fit with SE limits, 50% quantile
gphA <- xyplot(lo+fit+hi~logFert, data=hat50, lty=c(2,1,2),lwd=1.5,type='l') +
  latticeExtra::as.layer(xyplot(LifeExpectancy~logFert,
                                data=hat50, pch='.', cex=2))
## Panel B: Multiple quantiles; unweighted and weighted fits
gph19 <- xyplot(q10+q90+qwt10+qwt90 ~ logFert, type="l",
                data=hat19, lty=rep(1:2,c(2,2)),lwd=1.5)
gphB <- xyplot(LifeExpectancy ~ logFert, data=wdi) + as.layer(gph19)
update(c("A: 50% curve, 2 SE limits"=gphA, "B: 0.1, 0.9 quantiles"=gphB,
         x.same=T, y.same=T), between=list(x=0.5),
       xlab="Fertility Rate", ylab="Life Expectancy",
       scales=list(x=list(at=log(2^((0:5)/2)), labels=round(2^((0:5)/2),1)),
                   alternating=F),
       par.settings=DAAG::DAAGtheme(color=F, col='gray50', cex=2, pch='.'))
```

```{r 4_16, eval=F}
```

```{r D5_3e}
## Plots for the individual quantiles can be obtained thus:
## ## Panel A
plot(fit.qgam, shift=mean(predict(fit.qgam)))
## Panel B, 10% quantile
fitm10 <- qdo(fit19.mqgam, qu=0.1)
plot(fitm10, resid=T, shift=mean(predict(fitm10)),
     ylim=range(wdi$LifeExpectancy), cex=2)
wfitm10 <- qdo(wtd19.mqgam, qu=0.1)
plot(wfitm10, resid=T, shift=mean(predict(wfitm10)),
     ylim=range(wdi$LifeExpectancy), cex=2)
```

### Section 4.6: Further reading and remarks

### Exercises (4.7)

```{r D7a1}
roller.lm <- lm(depression~weight, data=DAAG::roller)
roller.lm2 <- lm(depression~weight+I(weight^2), data=DAAG::roller)
```

```{r D7a2, w=4, h=4, out.width="40%"}
toycars <- DAAG::toycars
lattice::xyplot(distance ~ angle, groups=factor(car), type=c('p','r'),
                data=toycars, auto.key=list(columns=3))
```

```{r D7b1}
parLines.lm <- lm(distance ~ 0+factor(car)+angle, data=toycars)
sepLines.lm <- lm(distance ~ factor(car)/angle, data=toycars)
```

```{r D7b2}
sepPol3.lm <- lm(distance ~ factor(car)/angle+poly(angle,3)[,2:3], data=toycars)
```

```{r D7b3}
sapply(list(parLines.lm, sepLines.lm, sepPol3.lm), AICcmodavg::AICc)
```

```{r <<D7b4}
setNames(sapply(list(parLines.lm, sepLines.lm, sepPol3.lm),
  function(x)summary(x)$adj.r.squared), c("parLines","sepLines","sepPol3"))
```

```{r D7c}
seedrates.lm <- lm(grain ~ rate + I(rate^2), data=seedrates)
seedrates.pol <- lm(grain ~ poly(rate,2), data=seedrates)
```

```{r D7d1}
geo.gam <- gam(thickness ~ s(distance), data=DAAG::geophones)
```

```{r D7d2, w=6, h=2.7, mfrow=c(1,2), out.width="95%"}
plot(DAAG::geophones$distance, acf(resid(geo.gam), lag.max=55)$acf)
Box.test(resid(geo.gam), lag=10)
Box.test(resid(geo.gam), lag=20)
Box.test(resid(geo.gam), lag=20, type="Ljung")
```

```{r D7e, w=4, h=4, out.width="40%"}
library(mgcv)
xy <- data.frame(x=1:200, y=arima.sim(list(ar=0.75), n=200))
df.gam <- gam(y ~ s(x), data=xy)
plot(df.gam, residuals=TRUE)
```


