---
title: "Chapter 7: Multilevel models, and repeated measures"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Ch7: Multilevel models . . .}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

On options for working with the code see the vignettes
[Ch1-Learning](PGRcode/Ch1-Learning.html) and
[UsingCode](PGRcode/UsingCode.html).

```{r CodeControl, echo=FALSE}
## xtras=TRUE
xtras <- F
library(knitr)
## opts_chunk[['set']](results="asis")
## opts_chunk[['set']](eval=T)
opts_chunk[['set']](eval=F)
options(rmarkdown.html_vignette.check_title = FALSE)
```

##### Packages required (plus any dependencies)
DAAG lme4 afex MASS utils devtools qra glmmTMB DHARMa MEMSS forecast splines gamlss plotrix nlme

Additionally, knitr and Hmisc are required in order to process the Rmd source file.

```{r setup, cache=FALSE}
Hmisc::knitrSet(basename="mva", lang='markdown', fig.path="figs/g", w=7, h=7)
oldopt <- options(digits=4, formatR.arrow=FALSE, width=70, scipen=999)
library(knitr)
## knitr::render_listings()
opts_chunk[['set']](cache.path='cache-', out.width="80%", fig.align="center", 
                    fig.show='hold', size="small", ps=10, strip.white = TRUE,
                    comment=NA, width=70, tidy.opts = list(replace.assign=FALSE))
```

### Section 7.1 Corn yield data --- analysis using `aov()`

#####                  Corn yield measurements example

```{r 7_1, echo=FALSE, w=4,h=2.5, las=0, out.width="65%"}
ant111b <- within(DAAG::ant111b, Site <- reorder(site, harvwt, FUN=mean))
gph <- lattice::stripplot(Site ~ harvwt, data=ant111b,
                          xlab="Harvest weight of corn")
update(gph, par.settings=DAAG::DAAGtheme(color=FALSE), scales=list(tck=0.5))
```

```{r 7_1, eval=F}
```

```{r G1a}
ant111b <- DAAG::ant111b
ant111b.aov <- aov(harvwt ~ 1 + Error(site), data=ant111b)
```

```{r G1b}
summary(ant111b.aov)
```

#####                   Interpreting the mean squares
#####                    Details of the calculations
#####         Practical use of the analysis of variance results
#####                  Random effects vs. fixed effects
#####            Nested factors -- a variety of applications
#### Subsection 7.1.1: A More Formal Approach
#####       Relations between variance components and mean squares
#####               Interpretation of variance components
#####                      Intra-class correlation

### Section 7.2 Analysis using `lme4::lmer()`

```{r G2a, message=FALSE}
library(lme4)
ant111b.lmer <- lmer(harvwt ~ 1 + (1 | site), data=ant111b)
```

```{r G2b}
## Note that there is no degrees of freedom information.
print(ant111b.lmer, ranef.comp="Variance")
```

#####             The processing of  output from lmer()

```{r G2c}
coef(summary(ant111b.lmer))
```

#####             Fitted values and residuals in lmer()

```{r G2d}
s2W <- 0.578; s2L <- 2.37; n <- 4
sitemeans <- with(ant111b, sapply(split(harvwt, site), mean))
grandmean <- mean(sitemeans)
shrinkage <- (n*s2L)/(n*s2L+s2W)
## Check  that fitted values equal BLUPs, and compare with site means
BLUP <- grandmean + shrinkage*(sitemeans - grandmean)
BLUP <- fitted(ant111b.lmer)[match(names(sitemeans), ant111b$site)]
BLUP <- grandmean + ranef(ant111b.lmer)$site[[1]]
```

```{r G2e}
rbind(BLUP=BLUP, sitemeans=sitemeans)
```

#####  *Uncertainty in the parameter estimates --- profile likelihood and alternatives

```{r G2f, message=FALSE}
prof.lmer <- profile(ant111b.lmer)
CI95 <- confint(prof.lmer, level=0.95)
rbind("sigmaL^2"=CI95[1,]^2, "sigma^2"=CI95[2,]^2)
```

```{r G2g}
CI95[3,]
```

```{r 7_2, w=5.25, h=2.1, echo=FALSE, out.width="85%"}
library(lattice)
gph <- xyplot(prof.lmer, conf=c(50, 80, 95, 99)/100,
              aspect=0.8, between=list(x=0.35))
update(gph, scales=list(tck=0.5), ylab="Normal deviate")
```

```{r 7_2, eval=F}
```

#####         Modeling more than two levels of random variation

### Section 7.3 Survey data, with clustering

```{r 7_3, w=4.0, h=2.5, echo=FALSE, out.width="65%"}
## Means of like (data frame science: DAAG), by class
science <- DAAG::science
classmeans <- with(science, aggregate(like, by=list(PrivPub, Class), mean))
# NB: Class identifies classes independently of schools
#     class identifies classes within schools
names(classmeans) <- c("PrivPub", "Class", "avlike")
gph <- bwplot(~avlike|PrivPub, layout=c(1,2), xlab="Average score",
              panel=function(x,y,...){panel.bwplot(x,y,...)
              panel.rug(x,y,...)}, data=classmeans)
update(gph, scales=list(tcl=0.4))
```

```{r 7_3, eval=F}
```

#### Subsection 7.3.1: Alternative models

```{r G3_1a}
science <- DAAG::science
science.lmer <- lmer(like ~ sex + PrivPub + (1 | school) +
                     (1 | school:class), data = science,
                     na.action=na.exclude)
```

```{r G3_1b}
print(VarCorr(science.lmer), comp="Variance", digits=2)
```

```{r G3_1c}
print(coef(summary(science.lmer)), digits=2)
```

```{r G3_1d}
summary(science.lmer)$ngrps
```

```{r G3_1e}
science1.lmer <- lmer(like ~ sex + PrivPub + (1 | school:class),
                      data = DAAG::science, na.action=na.exclude)
```

```{r G3_1f}
print(VarCorr(science1.lmer), comp="Variance", digits=3)
print(coef(summary(science1.lmer)), digits=2)
```

```{r G3_1g}
opt <- options(contrasts=c("contr.sum","contr.poly"))
  # Change is otherwise made as and if required for individual factors
  # prior to fitting model, and a warning message is generated.
afex::mixed(like ~ sex + PrivPub + (1 | school:class), method="KR", type=2,
            data = na.omit(science),  sig_symbols=rep("",4), progress=FALSE)
options(opt)       # Reset to previous contrasts setting
```

#####              More detailed examination of the output

```{r G3_1h}
## Use profile likelihood
pp <- profile(science1.lmer, which="theta_")
# which="theta_": all random parameters
# which="beta_": fixed effect parameters
var95 <- confint(pp, level=0.95)^2
# Square to get variances in place of SDs
rownames(var95) <- c("sigma_Class^2", "sigma^2")
signif(var95, 3)
```

```{r G3_1i}
## Fit model and generate quantities that will be plotted
science1.lmer <- lmer(like ~ sex + PrivPub + (1 | school:class),
data = science, na.action=na.omit)
## Panel A: random site effects vs number in class
ranf <- ranef(obj = science1.lmer, drop=TRUE)[["school:class"]]
flist <- science1.lmer@flist[["school:class"]]
privpub <- science[match(names(ranf), flist), "PrivPub"]
num <- unclass(table(flist)); numlabs <- pretty(num)
## Panel B: Within class variance estimates vs numbers
res <- residuals(science1.lmer)
vars <- tapply(res, INDEX=list(flist), FUN=var)*(num-1)/(num-2)
## Panel C: Normal probability of random site effects (`ranf`)
## Panel D: Normal probability of residuals (`res`)
```

```{r 7_4, w=6, h=4.4, bot=0.5, left=-1, top=0.5, ps=10, mgp=c(1.75,0.5,0), mfrow=c(2,2), echo=FALSE, out.width="100%"}
opar <- par(oma=c(0,0,1.5,0))
## Panel A: Plot effect estimates vs number
xlab12 <- "# in class (square root scale)"
plot(sqrt(num), ranf, xaxt="n", pch=c(1,3)[as.numeric(privpub)], cex=0.8,
     xlab=xlab12, ylab="Estimate of class effect", fg="gray")
lines(lowess(sqrt(num[privpub=="private"]),
ranf[privpub=="private"], f=1.1), lty=2)
lines(lowess(sqrt(num[privpub=="public"]),
ranf[privpub=="public"], f=1.1), lty=3)
axis(1, at=sqrt(numlabs), labels=paste(numlabs), lwd=0, lwd.ticks=1)
## Panel B: Within class variance estimates vs numbers
plot(sqrt(num), vars, pch=c(1,3)[unclass(privpub)], cex=0.8,
     xlab=xlab12, ylab="Within-class variance", fg="gray")
lines(lowess(sqrt(num[privpub=="private"]),
      as.vector(vars)[privpub=="private"], f=1.1), lty=2)
lines(lowess(sqrt(num[privpub=="public"]),
as.vector(vars)[privpub=="public"], f=1.1), lty=3)
## Panel C: Normal quantile-quantile plot of site effects
qqnorm(ranf, ylab="Ordered site effects", cex=0.8, main="",
       col="gray40", fg="gray")
## Panel D: Normal quantile-quantile plot of residuals
qqnorm(res, ylab="Ordered w/i class residuals", cex=0.8, main="",
       col="gray40", fg="gray")
par(fig = c(0, 1, 0, 1), oma = c(0, 0, 0, 0), mar = c(0, 0, 0, 0),
new = TRUE)
plot(0, 0, type = "n", bty = "n", xaxt = "n", yaxt = "n")
legend(x="top", legend=c("Private     ", "Public"), pch=c(1,3),
       lwd=c(1,1), lty=2:3, cex=1.25,
       xjust=0.5, yjust=0.8, horiz=TRUE, merge=FALSE, bty="n")
par(opar)
```

```{r 7_4, eval=F}
```

#### Subsection 7.3.2: Instructive, though faulty, analyses
#####                Ignoring class as the random effect

```{r G3_2a}
science2.lmer <- lmer(like ~ sex + PrivPub + (1 | school),
data = science, na.action=na.exclude)
print(coef(summary(science2.lmer)), digits=3)
```

```{r G3_2b}
## NB: Output is misleading
print(VarCorr(science2.lmer), comp="Variance", digits=3)
```

#####             Ignoring the random structure in the data

```{r G3_2c}
## Faulty analysis, using lm
science.lm <- lm(like ~ sex + PrivPub, data=science)
round(coef(summary(science.lm)), digits=4)
```

#### Subsection 7.3.3: Predictive accuracy

### Section 7.4 A multilevel experimental design

```{r 7_5, w=6, h=4.0, mgp=c(0,.5,.5), echo=FALSE, out.width="100%"}
par(mar=rep(0.25,4))
MASS::eqscplot(c(0,13),c(4.0,13),type="n",xlab="",ylab="", asp=1, axes=F)
eps <- 0.1
suby <- 12
vines<-function(x=1,y=1,subp=0, suby=12){
lines(c(y,y,y+1,y+1,y), suby-c(x,x+1,x+1,x,x),lwd=0.5)
points(c(y+.2,y+.2,y+.8,y+.8),suby-c(x+.2,x+.8,x+.8,x+.2),pch=3,cex=0.65)
text(y+.5,suby-(x+.5),paste(subp))
}
k<-0
for(i in c(1,3,5,7)){k<-k+1; vines(1,i,c(3,1,0,2)[k])}
k<-0
for(i in c(1,3,5,7)){k<-k+1; vines(4,i,c(2,1,0,3)[k])}
k <- 0
for(i in c(1,4,4,1)){k<-k+1
j<-c(9,9,11,11)[k]
vines(i,j,c(3,2,1,0)[k])
}
lines(c(2*eps,2.85,NA,10.15,13-2*eps), suby-c(3,3,NA,3,3),lty=2)
lines(c(0,2.85,NA,10.15,13),suby-c(0,0,NA,0,0),lty=2)
lines(c(0,4.5,NA,8.5,13),suby-c(8,8,NA,8,8),lty=2)
lines(rep(0,5),suby-c(0,1.25,NA,6.75,8),lty=2)
lines(rep(13,5),suby-c(0,1.25,NA,6.75,8),lty=2)
lines(c(9,9,12,12,9)+c(-eps,-eps,eps,eps,-eps),
      suby-(c(1,5,5,1,1)+c(-eps,eps,eps,-eps,-eps)), lwd=1)
lines(c(1,1,8,8,1)+c(-eps,-eps,eps,eps,-eps),
      suby-c(c(1,2,2,1,1)+c(-eps,eps,eps,-eps,-eps)), lwd=1)
lines(c(1,1,8,8,1)+c(-eps,-eps,eps,eps,-eps),
      suby-c(c(1,2,2,1,1)+3+c(-eps,eps,eps,-eps,-eps)), lwd=1)
text(6.5,suby,"6 meters height artifical shelter belt")
text(0,suby-4,"9 meters height shelter belt", srt=90)
text(13,suby-4,"19 meters height shelter belt", srt=-90)
text(6.5,suby-8,"Willow shelter belt")
text(0.5,suby-6.5,"0 Unshaded   \n1 Shaded Aug-Dec   \n2 Dec-Feb   \n3 Feb-May", adj=0)
text(6.5,suby-3,"16 meters height willow shelter belt")
offset <- c(4.75, 4.75-sqrt(3)*0.5)/6
arrows(x0=0,y0=12.1, x1=0+offset[1], y1=12.1+offset[2], length=0.15)
text(0.17, 12.55,"N")
```

```{r 7_5, eval=F}
```

#### Subsection 7.4.1: The analysis of variance (anova) table

```{r G4_1a}
## Analysis of variance: data frame kiwishade (DAAG)
kiwishade <- DAAG::kiwishade
kiwishade.aov <- aov(yield ~ shade + Error(block/shade),
data=kiwishade)
summary(kiwishade.aov)
```

#### Subsection 7.4.2: Expected values of mean squares

```{r G4_2a}
model.tables(kiwishade.aov, type="means", cterms="shade")
```

```{r G4_2b, size='normalsize'}
## Calculate treatment means
with(kiwishade, sapply(split(yield, shade), mean))
```

#### Subsection 7.4.3: \kern-4pt* The analysis of variance sums of squares breakdown

```{r G4_3a, size='normalsize'}
## For each plot, calculate mean, and differences from the mean
vine <- paste("vine", rep(1:4, 12), sep="")
vine1rows <- seq(from=1, to=45, by=4)
kiwivines <- unstack(kiwishade, yield ~ vine)
kiwimeans <- apply(kiwivines, 1, mean)
kiwivines <- cbind(kiwishade[vine1rows,  c("block","shade")],
Mean=kiwimeans, kiwivines-kiwimeans)
kiwivines <- with(kiwivines, kiwivines[order(block, shade), ])
mean(kiwimeans)      # the grand mean
```

```{r 7_6, w=6, h=2.8, lwd=1, echo=FALSE, results="hide"}
kiwishade <- DAAG::kiwishade
kiwimeans <- with(kiwishade, aggregate(yield,by=list(block,shade),mean))
names(kiwimeans) <- c("block","shade","meanyield")
plotmeans.lm <- lm(meanyield~block+shade, data=kiwimeans)
effects <- predict(plotmeans.lm, type="terms")
kiwishade.lm <- lm(yield ~ block*shade, data=kiwishade)
y <- c(effects[,"block"]/sqrt(2) * sqrt(16),
effects[,"shade"]/sqrt(3) * sqrt(12),
residuals(plotmeans.lm)/sqrt(6) * sqrt(4),
residuals(kiwishade.lm)/sqrt(12))
n <- rep(4:1, c(12, 12, 12, 48))
gps <- rep(c("block effect\n(ms=86.2)", "shade\n(464.8)",
"plot\n(20.9)", "vine\n(12.2)"), c(12, 12, 12, 48))
gps <- factor(gps, levels = rev(c("block effect\n(ms=86.2)",
"shade\n(464.8)", "plot\n(20.9)", "vine\n(12.2)")))
gps.sd <- sapply(split(y,gps), sd)
gps.av <- sapply(split(y,gps), mean)
plot(range(y), range(n)+c(-0.5, 0.5), xlab="", ylab="", yaxt="n", type="n", fg="gray")
text(y, n+0.15, "|", cex=0.8, col=adjustcolor("black", alpha.f=0.5))
un <- 1:4
sapply(un, function(j){lines(gps.av[j]+c(-gps.sd[j], gps.sd[j]),
rep(j-0.15,2), col="gray")
lines(rep(gps.av[j],2)-gps.sd[j],
j-0.15+c(-.06, .06), col="gray")
lines(rep(gps.av[j],2)+gps.sd[j],
j-0.15+c(-.06, .06), col="gray")
})
mtext(side=1,line=2.25, cex=0.9,
text="Variation in Yield (kg)\n(Add to grand mean of yield = 96.53)")
par(las=2)
axis(2, at=1:4, labels=levels(gps), lwd=0, lwd.ticks=1)
```

```{r 7_6, eval=F}
```

#### Subsection 7.4.4: The variance components
#### Subsection 7.4.5: The mixed model analysis

```{r G4_5a}
kiwishade.lmer <- lmer(yield ~ shade + (1|block) + (1|block:plot),
data=kiwishade)
# block:shade is an alternative to block:plot
```

```{r G4_5b}
print(kiwishade.lmer, ranef.comp="Variance", digits=3)
```

#####                  Residuals and estimated effects

```{r 7_7, w=6, h=6, echo=FALSE, out.width="100%"}
## Panel A
parsetA <- modifyList(DAAG::DAAGtheme(color=FALSE),
                      list(layout.heights=list(key.top=2.25, axis.top=.75)))
if (!exists("kiwishade.lmer"))
kiwishade.lmer <- lme4::lmer(yield ~ shade + (1|block/shade), data=kiwishade)
pk1 <- xyplot(residuals(kiwishade.lmer) ~ fitted(kiwishade.lmer)|block,
              groups=shade, layout=c(3,1), par.strip.text=list(cex=1.0),
              data=kiwishade, grid=TRUE,
              xlab="Fitted values (Treatment + block + plot effects)",
              ylab="Residuals",
  main=list(label="A: Standardized residuals after fitting block and plot effects",
          cex=1.05, x=0.01, y=0, font=1, just="left"),
             auto.key=list(space='top', points=TRUE, columns=4))
print(update(pk1, scales=list(x=list(alternating=TRUE), tck=0.35),
             par.settings=parsetA), position=c(0,.52,1,1))
## Panel B
ploteff <- ranef(kiwishade.lmer, drop=TRUE)[[1]]
nam <- names(sort(ploteff, decreasing=TRUE)[1:4])
parsetB <- modifyList(DAAG::DAAGtheme(color=FALSE),
                      list(layout.heights=list(axis.top=0.85)))
pk2 <- qqmath(ploteff, ylab="Plot effect estimates",
              key=list(x=0, y=0.98, corner=c(0,1),
              text=list(labels=nam, cex=0.65)),
              main=list(label="B: Normal Q-Q plot of plot effects",
                        cex=1.05, x=0.01, y=0.25, font=1, just="left"),
              xlab="Normal quantiles")
print(update(pk2, scales=list(tck=0.35), par.settings=parsetB),
      newpage=FALSE, position=c(0,0,.5,.5))
## Panel C
simvals <- simulate(kiwishade.lmer, nsim=3, seed=23)
simranef <- function(y)ranef(lme4::refit(kiwishade.lmer, y))[[1]]
simeff <- apply(simvals, 2, function(y) scale(simranef(y)))
simeff <- as.data.frame(simeff)
pk3 <- qqmath(~ sim_1+sim_2+sim_3, data=simeff,
              ylab="Simulated plot effects\n(2 sets, standardized)",
              Qs=list(tck=0.35), aspect=1,
               main=list(label="C: 3 simulations -- normal Q-Q plots",
                         x=0.01, y=0.25, cex=1.05, font=1, just="left"),
               xlab="Normal quantiles")
print(update(pk3, scales=list(tck=0.35), par.settings=parsetB),
      newpage=FALSE, position=c(0.48,0,1,.5))
```

```{r 7_7, eval=F}
```

```{r G4_5e}
with(kiwishade, sapply(split(yield, shade), mean))
```

#### Subsection 7.4.6: Predictive accuracy

### Section 7.5 Within and between subject effects

```{r 7_8, w=6.0, h=5.0, out.width="100%", echo=FALSE}
library(lattice)
tinting <- within(DAAG::tinting, targtint <- paste(target,toupper(tint),sep=':'))
gphA <- bwplot(targtint~log(csoa) | sex*agegp, data=tinting, layout=c(4,1), between=list(x=0.25))
mainA <- list("A: Boxplots for `log(csoa)`, by combinations of `target` and `tint`",
              font=1, x=0.125, y=0.125, cex=1.0, just='left')
gphB <- bwplot(targtint~log(it) | sex*agegp, data=tinting, layout=c(4,1), between=list(x=0.25))
mainB <- list("B: Boxplots for `log(it)`, by combinations of `target` and `tint`",
              font=1, x=0.125, y=0.125, cex=1.0, just='left')
print(update(gphA, main=mainA, xlab=""), position=c(0, 0.48, 1, 1.0), more=T)
print(update(gphB, main=mainB, xlab=""), position=c(0,0,1,0.52))
```

```{r 7_8, eval=F}
```

#####                       Model fitting criteria
#### Subsection 7.5.1: Model selection

```{r G5_1a}
## Capitalize tinting$agegp
levels(tinting$agegp) <- R.utils::capitalize(levels(tinting$agegp))
## Fit all interactions: data frame tinting (DAAG)
it3.lmer <- lmer(log(it) ~ tint*target*agegp*sex + (1 | id),
                 data=DAAG::tinting, REML=FALSE)
```

```{r G5_1b}
it2.lmer <- lmer(log(it) ~ (tint+target+agegp+sex)^2 + (1 | id),
                 data=DAAG::tinting, REML=FALSE)
```

```{r G5_1c}
it1.lmer <- lmer(log(it)~(tint+target+agegp+sex) + (1 | id),
                 data=DAAG::tinting, REML=FALSE)
```

```{r G5_1d}
anova(it1.lmer, it2.lmer, it3.lmer)
```

```{r G5_1e, eval=xtras, w=6, h=5, mfrow=c(2,2), out.width="100%"}
## Code that gives the first four such plots, for the observed data
interaction.plot(tinting$tint, tinting$agegp, log(tinting$it))
interaction.plot(tinting$target, tinting$sex, log(tinting$it))
interaction.plot(tinting$tint, tinting$target, log(tinting$it))
interaction.plot(tinting$tint, tinting$sex, log(tinting$it))
```

#### Subsection 7.5.2: Estimates of model parameters

```{r G5_2a}
it2.reml <- update(it2.lmer, REML=TRUE)
print(coef(summary(it2.reml)), digits=2)
# NB: The final column, giving degrees of freedom, is not in the
# summary output. It is our addition.
```

```{r G5_2b, size='normalsize'}
subs <- with(tinting, match(unique(id), id))
with(tinting, table(sex[subs], agegp[subs]))
```

### Section 7.6 A mixed model with a betabinomial error
#### Subsection 7.6.1: The betabinomial distribution

```{r G6_1a}
## devtools::install_github("jhmaindonald/qra")  # Use if not found on CRAN
```

#####                              Notation
#####                           Source of data

```{r G6_1b}
HawCon <- within(as.data.frame(qra::HawCon), {
  trtGp <- gsub("Fly", "", paste0(CN,LifestageTrt))
  trtGp <- factor(trtGp, levels=sort(unique(trtGp))[c(1,5,2,6,3,7,4,8)])
  trtGpRep <- paste0(CN,LifestageTrt,":",RepNumber)
  scTime <- scale(TrtTime)  # Centering and scaling can help model fit
  })
```

```{r G6_1c, message=FALSE, warning=FALSE}
## Load packages that will be used
suppressMessages(library(lme4)); suppressMessages(library(glmmTMB))
```

```{r G6_1d, message=FALSE, warning=FALSE}
library(splines)
form <- cbind(Dead,Live)~0+trtGp/TrtTime+(1|trtGpRep)
## Add the quadratic term from a degree 2 orthogonal polynomial
form2s <- update(form, . ~ . + scale(scTime^2))
  ## Scale "corrections" to reduce risk of potential numerical problems
HCbb.cll <- glmmTMB(form, dispformula=~trtGp+ns(scTime,2),
                    family=glmmTMB::betabinomial(link="cloglog"), data=HawCon)
HCbb2s.cll <- update(HCbb.cll, formula=form2s)
HCbb.logit <- glmmTMB(form, dispformula=~trtGp+ns(scTime,2),
                      family=glmmTMB::betabinomial(link="logit"), data=HawCon)
HCbb2s.logit <- update(HCbb.logit, formula=form2s)
```

```{r G6_1e, message=FALSE, warning=FALSE}
summary(HCbb2s.cll)$coefficients$cond["scale(scTime^2)",]    ## CLL
summary(HCbb2s.logit)$coefficients$cond["scale(scTime^2)",]  ## Logit
```

```{r G6_1f}
AICtab <- AIC(BB=HCbb.cll,HCbb2s.cll,HCbb.logit,HCbb2s.logit)
AICtab[['Details']] <- c(paste0('BB: Compl. log-log', c('', ', added curve')),
                          paste0('BB: Logit', c('', ', added curve')))
AICtab[order(AICtab[["AIC"]]), ]
```

```{r 7_9, fig.width=8.5, fig.height=4.25, fig.show='hold', top=2, out.width="100%",  fig.align='center', message=FALSE, echo=FALSE}
dat <- expand.grid(trtGp=factor(levels(HawCon$trtGp), levels=levels(HawCon$trtGp)),
TrtTime=pretty(range(HawCon$TrtTime),15), trtGpRep=NA)
ab <- qra::getScaleCoef(HawCon$scTime)
dat$scTime <- with(dat,(TrtTime-ab[1])/ab[2])
hatClog <- predict(HCbb2s.cll, newdata=dat)
hatClog <- predict(HCbb2s.cll, se=T, newdata=dat)
hatLgt <- predict(HCbb2s.logit, newdata=dat)
hatLgt <- predict(HCbb2s.logit, se=T, newdata=dat)
dat2 <- cbind(rbind(dat,dat), link=rep(c('clog','logit'), rep(nrow(dat),2)))
dat2 <- within(dat2, {fit<-c(hatClog$fit,hatLgt$fit);
se.fit<-c(hatClog$se.fit,hatLgt$se.fit);
link=rep(c('clog','logit'), rep(nrow(dat),2))})
dat2 <- within(dat2, {lwr<-fit-2*se.fit; upr<-fit+2*se.fit})
library(lattice)
my.panel.bands <- function(x, y, upper, lower, fill, col,
subscripts, ..., font, fontface)
{
upper <- upper[subscripts]
lower <- lower[subscripts]
panel.lines(x,lower, ...)
panel.lines(x,upper, ...)
}
panel2 <- function(x, y, ...){
panel.superpose(x, y, panel.groups = my.panel.bands, type='l', lty=3, alpha=0.25,...)
panel.xyplot(x, y, type='l', lwd=2, cex=0.6, ...)
}
parset <- simpleTheme(col=rep(4:1,rep(2,4)), lty=rep(1:2, 4), lwd=2)
## c('solid','1141')
p <- c(.02,.2,.5,.8, .99, .999968)
cloglog <- make.link('cloglog')$linkfun
logit <- make.link('logit')$linkfun
fitpos <- list(cloglog(p), logit(p))
lab <- paste(p)
lim <- list(cloglog(c(0.02, 0.99998)), logit(c(0.02, 0.99998)))
lim <- lapply(lim, function(x)c(x[1],x[1]+diff(x)*1.2))

gph <- xyplot(fit~TrtTime|link, outer=TRUE, data=dat2, groups=trtGp,
              upper = dat2$upr, lower = dat2$lwr, panel = panel2,
              xlab="Days in coolstorage", ylab="Fitted value",
              auto.key=list(text=levels(HawCon$trtGp), columns=4,
                            points=FALSE, lines=TRUE),
              par.settings=parset, layout=c(2,1),
              scales=list(x=list(at=c(2,6,10,14)),
              y=list(relation='free',
              at=fitpos, labels=lab, limits=lim), alternating=c(1,1)))
update(gph, strip=strip.custom(factor.levels=c("A: Complementary log-log link",
"B: Logit link")))
```

```{r 7_9, eval=F}
```

```{r 7_10, fig.width=6.5, fig.height=2.65, top=2, out.width="100%",  fig.align='center', fig.show="hold", echo=FALSE}
parset <- DAAG::DAAGtheme(color=TRUE, col.points=rep(1:4,rep(2,4)),
                          pch=rep(1:4,2), lwd=2)
HawCon$rho2clog <- qra::getRho(HCbb.cll)
HawCon$dispClog <- with(HawCon, 1+(Total-1)*rho2clog)
par(oma=c(0,0,2,0))
titles=c(expression("A: "*rho*", cloglog link"),expression("B: "*rho*", logit link"),
expression("C: Dispersion "*Phi*" (Panel A)"))
library(lattice)
HawCon$rho2logit <- qra::getRho(HCbb.logit)
xyplot(rho2clog+rho2logit+dispClog ~ TrtTime, groups=trtGp, data=HawCon,
       outer=TRUE, between=list(x=0.25), par.settings=parset,
   scales=list(x=list(alternating=FALSE,at=c(4,8,12), labels=paste(c(4,8,12))),
               y=list(relation='free',tick.number=4)),
   auto.key=list(columns=4, between.columns=2, between=1),
   xlab="Days in coolstorage", ylab="Parameter Estimate",
   strip=strip.custom(factor.levels=titles))

```

```{r 7_10, eval=F}
```

#### Subsection 7.6.2: Diagnostic checks

```{r 7_11, fig.width=8, fig.height=8, left=-1.5, top=3.5, bot=1.5, out.width="100%", mgp=c(2.65,.5,0), fig.align='center', mfrow=c(2,2), message=FALSE, echo=FALSE}
par(oma=c(0,0,2,0.5))
## Code for plots, excluding main titles
set.seed(29)
simRef <- DHARMa::simulateResiduals(HCbb.cll, n=250, seed=FALSE)
DHARMa::plotQQunif(simRef)
title(main="A: Q-Q plot, quantile residuals", adj=0, line=2.5, 
      font.main=1, cex.main=1.25)
DHARMa::plotResiduals(simRef, form=HawCon$trtGp)
axis(1, at=c(2,4,6,8), labels=levels(HawCon$trtGp)[c(2,4,6,8)],
     gap.axis=0, line=1, lwd=0)
title(main="B: Boxplots comparing treatment groups", adj=0, line=2.5,
      font.main=1, cex.main=1.25)
DHARMa::plotResiduals(simRef)
title(main="C: Plot against model predictions", adj=0, font.main=1, line=3.25,
       cex.main=1.25)
DHARMa::plotResiduals(simRef, form=HawCon$Total)
title(main="D: Plot against total number", adj=0, font.main=1, line=3.25,
       cex.main=1.25)
```

```{r 7_11, eval=F}
```

#### Subsection 7.6.3: Lethal time estimates and confidence intervals

```{r G6_3a}
shorten <- function(nam, leaveout=c('trtGp','Fly',':')){
for(txt in leaveout){
nam <- gsub(txt,'', nam, fixed=TRUE)
}
nam
}
```

```{r G6_3b, echo=-3}
LTbb.cll <- qra::extractLT(p=0.99, obj=HCbb.cll, link="cloglog",
                        a=1:8, b=9:16, eps=0, df.t=NULL)[,-2]
rownames(LTbb.cll) <- shorten(rownames(LTbb.cll))
```

```{r G6_3c}
LTbb.logit <- qra::extractLT(p=0.99, obj=HCbb.logit, link="logit",
                             a=1:8, b=9:16, eps=0, offset=0,
df.t=NULL)[,-2]
rownames(LTbb.logit) <- shorten(rownames(LTbb.logit))
```

```{r 7_12, echo=FALSE, fig.width=5, fig.asp=0.6, warning=FALSE, fig.align='center', message=FALSE, out.width="65%", fig.show="hold"}
library(plotrix)
gpNam <- rownames(LTbb.cll)
ordEst <- order(LTbb.cll[,1])
col5 <- c("blue","lightslateblue","blueviolet",'gray50','gray80')
plotCI(1:8-0.1, y=LTbb.cll[ordEst,1], ui=LTbb.cll[ordEst,3],
  li=LTbb.cll[ordEst,2], lwd=2, col=col5[1], xaxt="n",
  fg="gray", xlab="", ylab="LT99 Estimate (days)",
  xlim=c(0.8,8.2), ylim=c(0,19.5))
plotCI(1:8+0.1, y=LTbb.logit[ordEst,1], ui=LTbb.logit[ordEst,3],
  li=LTbb.logit[ordEst,2], lwd=2, col=col5[2], xaxt="n", add=TRUE)
axis(1, labels=FALSE, tck=0.02, col.ticks="gray40")
text(x = 1:length(gpNam)+0.1,
  ## Move labels to just below bottom of chart.
  y = par("usr")[3] - 0.8,
  labels = gpNam[ordEst], ## Use names from the data list.
    xpd = NA,        ## Change the clipping region
  srt = 45,          ## Rotate the labels by 45 degrees.
  adj = 0.95)        ## Adjust the labels to almost 100% right-justified
grid()
legend("topleft", legend=c("BB-clog", "BB-logit"),
       inset=c(0.01,0.01), lty=c(1,1), col=col5[1:2],
text.col=col5[1:2], bty="n",y.intersp=0.85)

```

```{r 7_12, eval=F}
```

```{r G6_3e, echo=1:4}
HCbin.cll <- glmmTMB(cbind(Dead,Live)~0+trtGp/TrtTime+(scTime|trtGp),
                     family=binomial(link="cloglog"), data=HawCon)
LTbin.cll <- qra::extractLT(p=0.99, obj=HCbin.cll,
                            a=1:8, b=9:16, eps=0, df.t=NULL)[,-2]
rownames(LTbin.cll) <- shorten(rownames(LTbin.cll))
```

#####               A warning against over-interpretation

### Section 7.7 Observation level random effects --- the `moths` dataset

```{r G7a}
moths <- DAAG::moths
moths$transect <- 1:41  # Each row is from a different transect
moths$habitat <- relevel(moths$habitat, ref="Lowerside")
A.glmer <-  glmer(A~habitat+sqrt(meters)+(1|transect),
family=poisson(link=sqrt), data=moths)
print(summary(A.glmer), show.resid=FALSE, correlation=FALSE, digits=3)
```

```{r G7b}
suppressPackageStartupMessages(library(gamlss))
A1quasi.glm <- glm(A~habitat, data=moths, family=quasipoisson(link=sqrt))
Asqrt.lss <- gamlss(A ~ habitat + sqrt(meters), trace=FALSE,
                    family = NBI(mu.link='sqrt'), data = moths)
A1.glmer <- glmer(A~habitat+(1|transect), data=moths, family=poisson(link=sqrt))
```

```{r G7c}
Cglm <- coef(summary(A1quasi.glm))
Cglmer <- coef(summary(A1.glmer))
fitAll <- cbind("quasi-Coef"=Cglm[-1,1], "quasi-SE"=Cglm[-1,2],
  "NBI-Coef"=coef(Asqrt.lss)[2:8], "NBI-SE"=c(0.94,.41,.43,.51,.39,.53,.75),
  "glmer-Coef"=Cglmer[-1,1], "glmer-SE"=Cglmer[-1,2])
rownames(fitAll) <- substring(rownames(fitAll),8)
round(fitAll, 2)  # NB, all SEs are for the difference from 'Lowerside'

```

```{r G7d}
detach("package:glmmTMB", character.only=TRUE)
```

### Section 7.8 Repeated measures in time
#####              The theory of repeated measures modeling
#####                       *Correlation structure
#####         Different approaches to repeated measures analysis
#### Subsection 7.8.1: Example -- random variation between profiles

```{r G8_1a}
humanpower1 <- DAAG::humanpower1
```

```{r 7_13, echo=FALSE, w=4, h=4, out.width="50%"}
## Plot points and fitted lines
xyplot(o2 ~ wattsPerKg, groups=id, data=humanpower1,
       par.settings=DAAG::DAAGtheme(color=F), scales=list(tck=0.5),
       panel=function(x,y,subscripts,groups,...){
                yhat <- fitted(lm(y ~ groups*x))
                panel.superpose(x, y, subscripts, groups, pch=1:5, cex=1.2)
                panel.superpose(x, yhat, subscripts, groups, type="b", cex=0.5)
                },
       auto.key=list(columns=5, lines=T),
       xlab="Watts per kilogram",
       ylab=expression("Oxygen intake ("*ml.min^{-1}*.kg^{-1}*")"))
```

```{r 7_13, eval=F}
```

#####               Separate lines for different athletes

```{r G8_1c}
## Calculate intercepts and slopes; plot Slopes vs Intercepts
## Uses the function lmList() from the lme4 package
humanpower1 <- DAAG::humanpower1
hp.lmList <- lmList(o2 ~ wattsPerKg | id, data=humanpower1)
coefs <- coef(hp.lmList)
round(coefs,3)
c("Correlation between intercept and slope"=cor(coefs[,1],coefs[,2]))
```

#####                    A random coefficients model

```{r G8_1d}
hp.lmer <- lmer(o2 ~ wattsPerKg + (wattsPerKg | id), data=humanpower1)
```

```{r G8_1e}
hp.lmer <- lmer(o2 ~ wattsPerKg + (0+wattsPerKg | id), data=humanpower1)
print(summary(hp.lmer), digits=3)
```

```{r G8_1f}
sort(coef(lmList(o2 ~ wattsPerKg | id, data=humanpower1))[,1])
```

#### Subsection 7.8.2: Orthodontic measurements on children

```{r G8_2a}
Orthodont <- MEMSS::Orthodont
Orthodont$logdist <- log(Orthodont$distance)
```

```{r 7_14, echo=FALSE, w=5.5, h=3.3, cex.lab=0.9, out.width="100%"}
## Plot showing pattern of change for each of the 25 individuals
gph <- xyplot(distance ~ age | Subject, groups=Sex, data=Orthodont,
              scales=list(x=list(rot=90, tick.number=3), y=list(log=2), tck=0.5), type=c("p","r"),
              layout=c(11,3))
update(gph, xlab=list("Age", cex=1.4), ylab=list("Distance", cex=1.4),
       par.settings=DAAG::DAAGtheme(color=FALSE, fontsize=list(text=7, points=4)))
```

```{r 7_14, eval=F}
```

#####                    Preliminary data exploration

```{r G8_2d}
## Use lmList() to find the slopes
ab <- cbind(coef(lmList(distance ~ age | Subject, Orthodont)),
            coef(lmList(logdist ~ age|Subject, data=Orthodont)))
names(ab) <- c("a", "b","alog","blog")
## Obtain intercept at x=mean(x)=11, for each subject.
## (For each subject, this is independent of the slope)
ab <- within(ab, {ybar = a + b*11; b=b; ylogbar = alog + blog * 11;
                  blog=blog; sex = substring(rownames(ab), 1 ,1)
                  })
bySex <- sapply(split(ab, ab$sex), function(z)range(z$b))
extremes <- with(ab, ybar %in% range(ybar) | b %in% bySex)

```

```{r 7_15, w=7.2, h=3.6, bot=1, left=-1, top=1, rt=1, ps=9, mfrow=c(1,2), mgp=c(2,0.5,0), echo=FALSE, out.width="100%"}
fundiff <- function(x)range(x)+diff(range(x))*c(-0.015, 0.04)
lims <- sapply(subset(ab, select=c('ybar',"b","ylogbar","blog")), fundiff)
plot(b ~ ybar, col=c(F="gray40", M="black")[sex], data=ab,
     fg="gray", xlim=lims[,"ybar"], ylim=lims[,"b"],
     pch=c(F=1, M=3)[sex], xlab="Mean distance", ylab="Slope")
with(subset(ab, extremes), text(b ~ ybar,
    labels=rownames(ab)[extremes], pos=4, xpd=TRUE))
mtext(side=3, line=0.75,"A: Profiles for distances", adj=0)
# Type 'qqnorm(ab$b)' to see the extent of M13's outlyingness
## For Panel B, repeat with logdist replacing distance
plot(blog ~ ylogbar, col=c(F="gray40", M="black")[sex], data=ab, fg="gray",
     pch=c(F=1, M=3)[sex], xlim=lims[,"ylogbar"], ylim=lims[,"blog"],
     xlab="Mean log distance", ylab="Slope")
with(subset(ab, extremes),
     text(blog ~ ylogbar, labels=rownames(ab)[extremes], pos=4, xpd=TRUE))
mtext(side=3, line=0.75,"B: Logarithms of distances", adj=0)
```

```{r 7_15, eval=F}
```

```{r G8_2f, size='normalsize'}
## Compare male slopes with female slopes
extreme.males <- match(c("M04","M13"), rownames(ab))
with(ab[-extreme.males,],
t.test(blog[sex=="F"], blog[sex=="M"], var.equal=TRUE))
# Specify var.equal=TRUE, to allow comparison with anova output
```

#####                    A random coefficients model

```{r G8_2h}
keep <- !(Orthodont$Subject%in%c("M04","M13"))
Orthodont$scAge <- with(Orthodont, age-11)  ## Center values of age
orthdiffx.lmer <- lmer(logdist ~ Sex * scAge + (scAge | Subject),
                       data=Orthodont, subset=keep)
```

```{r G8_2i}
rePCA(orthdiffx.lmer)
```

```{r G8_2j, echo=2:3}
opt <- options(digits=3, scipen=2)
orthdiff.lmer <- lmer(logdist ~ Sex * scAge + (1 | Subject),
                      data=Orthodont, subset=keep)
summary(orthdiff.lmer)
options(opt)
```

```{r G8_2k}
Orthodont2 <- droplevels(subset(Orthodont, keep))
opt <- options(contrasts=c("contr.sum","contr.poly"))
orthdiff.mixed <- afex::mixed(logdist ~ Sex * scAge + (1 | Subject), type=2,
                              method='S', data=Orthodont2)
  ## NB `type` refers to type of test, NOT `error` type.
options(opt)       # Reset to previous contrasts setting
orthdiff.mixed
```

```{r G8_2l}
contrasts(Orthodont2[['Subject']]) <- 'contr.sum'
contrasts(Orthodont2[['Sex']]) <- 'contr.sum'
```

```{r G8_2m}
orthdiffa.lmer <- update(orthdiff.lmer, formula=. ~ . -Sex:scAge)
AIC(orthdiffa.lmer,orthdiff.lmer)
```

#####                Correlation between successive times

```{r G8_2n}
res <- resid(orthdiff.lmer)
Subject <- factor(Orthodont$Subject[keep])
orth.arma <- sapply(split(res, Subject),
                    function(x)forecast::auto.arima(x)$arma[c(1,6,2)])
orthsum <- apply(orth.arma,2,sum)
orth.arma[, orthsum>0]
```

#####             Fitting a sequential correlation structure

```{r G8_2o}
library(nlme)
keep <- !(Orthodont$Subject%in%c("M04","M13"))
Orthodont2 <- droplevels(subset(Orthodont,keep))
orthdiff.lme <- lme(logdist ~ Sex * scAge, random = ~1|Subject,
                    cor=corCAR1(0.1, form=~1|Subject), data=Orthodont2)
## For AR1 models `phi` is the sequential correlation estimate
orthdiff.lme$modelStruct$corStruct
```

#####             *The variance for the difference in slopes

### Section 7.9: Further notes on multilevel models
#### Subsection 7.9.1: Sources of variation -- complication or focus of interest?
#### Subsection 7.9.2: Predictions from models with a complex error structure
#####  Consequences from assuming an overly simplistic error structure
#### Subsection 7.9.3: An historical perspective on multilevel models
#### Subsection 7.9.4: Meta-analysis
#### Subsection 7.9.5: Functional data analysis
#### Subsection 7.9.6: Error structure in explanatory variables

### Exercises (7.12)

```{r G12a}
n.omit <- 2
take <- rep(TRUE, 48)
take[sample(1:48,2)] <- FALSE
kiwishade.lmer <- lmer(yield ~ shade + (1|block) + (1|block:plot),
                       data = kiwishade,subset=take)
vcov <- VarCorr(kiwishade.lmer)
print(vcov, comp="Variance")

```

```{r G12b}
cult.lmer <- lmer(ct ~ Cultivar + Dose + factor(year) +
                       (-1 + Dose | gp), data = DAAG::sorption, REML=TRUE)
cultdose.lmer <- lmer(ct ~ Cultivar/Dose + factor(year) +
                           (-1 + Dose | gp), data = DAAG::sorption, REML=TRUE)
```

