---
title: "Chapter 9: Multivariate data exploration and discrimination"
output: rmarkdown::html_vignette
options: 
  params: 
    rmarkdown.html_vignette.check_title: false
vignette: >
  %\VignetteIndexEntry{Ch9: Multivariate data exploration . . .}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

On options for working with the code see the vignettes
[Ch1-Learning](PGRcode/Ch1-Learning.html) and
[UsingCode](PGRcode/UsingCode.html).

```{r CodeControl, echo=FALSE}
options(rmarkdown.html_vignette.check_title = FALSE)
## xtras=TRUE
xtras=FALSE
library(knitr)
## opts_chunk[['set']](results="asis")
opts_chunk[['set']](eval=F)
## opts_chunk[['set']](eval=T)
```

##### Packages required (plus any dependencies)

Packages used are: DAAG MASS RColorBrewer teigen BiocManager DAAGbio 
hddplot lmtest splines cobalt mice datasets car micemd oz randomForest
ggplot2 latticeExtra mvtnorm teigen limma hddplot mgcv MatchIt sandwich
gridExtra DAAGbio mlbench (exercise).

Additionally, knitr and Hmisc are required in order to process the Rmd source file.

```{r setup, cache=FALSE}
Hmisc::knitrSet(basename="mva", lang='markdown', fig.path="figs/g", w=7, h=7)
oldopt <- options(digits=4, formatR.arrow=FALSE, width=70, scipen=999)
library(knitr)
opts_chunk[['set']](cache.path='cache-', out.width="80%", fig.align="center", 
                    fig.show='hold', ps=10, strip.white = TRUE,
                    comment=NA, width=70, tidy.opts = list(replace.assign=FALSE))
```

```{r latticePlus, echo=FALSE}
suppressPackageStartupMessages(library(latticeExtra))
```

### Section 9.1: Multivariate exploratory data analysis
```{r I1a}
## Make the lattice package and the possum dataset available
library(latticeExtra)
possum <- DAAG::possum
```

#### Subsection 9.1.1: Scatterplot matrices
```{r 9_1, w=6.8, h=4.25, lwd=0.5, echo=FALSE, out.width="100%"}
## Colors distinguish sexes; symbols distinguish sites
sitenames <- row.names(DAAG::possumsites)[c(1,2,4:6,3,7)]
key <- list(points = list(pch=0:6), text=list(sitenames),
            columns=4, between=1, between.columns=2)
colr <- c("red","blue")
vnames <- c("tail\nlength","foot\nlength", "conch\nlength")
gphA <- with(possum, splom(~ possum[, 9:11], pch=(0:6)[site], col=colr[sex],
            xlab="",  varnames=vnames, key=key, axis.line.tck=0.6))
gphB <- with(possum, cloud(earconch~taill+footlgth, data=possum, 
  col=colr[sex], key=key, pch = (0:6)[site], 
  zlab=list("earconch", rot=90), zoom=0.925))
update(c("A: Scatterplot matrix"=gphA, "B: Cloud plot"=gphB),
       between=list(x=1))
```

```{r 9_1, eval=FALSE}
```

#### Subsection 9.1.2: Principal components analysis
#####                     Preliminary data scrutiny
```{r I1_2a}
## Ratios of largest to smallest values: possum[, 6:14] (DAAG)
possum <- DAAG::possum
sapply(na.omit(possum[, 6:14]), function(x)round(max(x)/min(x),2))
```

```{r I1_2b}
## Principal components calculations: possum[, 6:14] (DAAG)
here <- complete.cases(possum[, 6:14])
possum.prc <- prcomp(log(possum[here, 6:14]))
scores <- cbind(predict(possum.prc), possum[here, c('sex', 'site')])
```

```{r 9_2, w=4.8,h=2.9, cex.lab=1.15, echo=FALSE, out.width="75%"}
## For parset, key and colr; see code for Fig 9.1
pchr <- c(3,4,0,8,2,10,1)
parset <- list(fontsize=list(text=10, points=6), cex=0.75, pch=pchr, alpha=0.8)
key <- modifyList(key, list(columns=1, space="right"))
gph <- with(scores, xyplot(PC2 ~ PC1, aspect="iso", key = key,
            col = colr[sex], pch = (0:6)[site]))
update(gph, scales=list(tck=0.5), par.settings=parset,
       xlab="1st Principal Component", ylab="2nd Principal Component")
```

```{r 9_2, eval=FALSE}
```

```{r I1_2d}
print(summary(possum.prc),digits=2)
cat("\nRotations (otherwise called Loadings)\n")
print(possum.prc$rotation, digits=2)
## By default, blanks are shown for loadings < 0.1 in magnitude
```

#####           The stability of the principal components plot
```{r 9_3, w=5.5, h=2.4, cache=FALSE, warning=FALSE, echo=FALSE, out.width="100%"}
suppressPackageStartupMessages(library(ggplot2))
theme_set(theme_gray(base_size=8))
## Bootstrap principal components calculations: possum (DAAG)
## Sample from rows where there are no missing values
rowsfrom <- (1:nrow(possum))[complete.cases(possum[, 6:14])]
logpossum6to14 <- log(possum[rowsfrom, 6:14])
sexPop <- possum[rowsfrom, c("sex","Pop")]
n <- length(rowsfrom); ntimes <- 3
bootscores <- data.frame(scores1=numeric(ntimes*n), scores2=numeric(ntimes*n))
for (i in 1:ntimes){
  samprows <- sample(1:n, n, replace=TRUE)
  bootscores[n*(i-1)+(1:n), 1:2] <-
  prcomp(logpossum6to14[samprows, ])$x[, 1:2]
}
bootscores[, c("sex","Pop")] <- sexPop[samprows, ]
bootscores$sampleID <- factor(rep(1:ntimes, rep(n,ntimes)))
gph <- quickplot(x=scores1, y=scores2, colour=sex, size=I(1.0),
  asp=1, shape=Pop, facets=.~sampleID, data=bootscores) + 
  scale_shape_discrete(solid=F)
gph + scale_colour_manual(values=c("m"="blue","f"="red"))  +
  xlab("First Principal Component") + ylab("Second Principal Component")
```

```{r 9_3, eval=FALSE}
```

#### Subsection 9.1.3: Multi-dimensional scaling
#####                         Distance measures
#####                             Ordination
```{r 9_4, w=7.25, h=3.6, echo=FALSE, out.width="100%"}
## Code that will display individual graphs
d.possum <- dist(possum[,6:14])  # Euclidean distance matrix
MASS::sammon(d.possum, k=2, trace=FALSE)$points |> as.data.frame() |>
  setNames(paste0("ord",1:2)) |> cbind(Pop=DAAG::possum$Pop) -> sammon.possum
MASS::isoMDS(d.possum, k=2, trace=FALSE) |> as.data.frame() |> 
  setNames(paste0("ord",1:2)) |> cbind(Pop=DAAG::possum$Pop) -> mds.possum
gph1 <- xyplot(ord2~ord1, groups=Pop, aspect="iso", data=sammon.possum)
gph2 <- xyplot(ord2~ord1, groups=Pop, aspect="iso", data=mds.possum)
update(c(gph1, gph2, layout=c(2,1)), 
       par.settings=simpleTheme(pch=c(1,3)),
       between=list(x=0.5), auto.key=list(columns=2),
       strip=strip.custom(factor.levels=c("A: Sammon","B: ISOmds")))
```

```{r 9_4, eval=FALSE}
```

#####                            Binary data

### Section 9.2: Principal component scores in regression
```{r I2a}
## Principal components: data frame socsupport (DAAG)
socsupport <- DAAG::socsupport
ss.pr1 <- prcomp(as.matrix(na.omit(socsupport[, 9:19])), retx=TRUE, scale=TRUE)
```

```{r 9_5, echo=FALSE, w=4.5, h=4.5, out.width="55%"}
oldpar <- par(fg='gray40',col.axis='gray20',lwd=0.5,col.lab='gray20')
pairs(ss.pr1$x[, 1:3], col='gray40', gap=0.2)
par(oldpar)
```
```{r 9_5, eval=F}
```

```{r I2c}
summary(sort(ss.pr1$rotation[,1]))
## Note the very large maximum value
which.max(ss.pr1$x[,1])
## Try also boxplot(ss.pr1$x[,1])
## ss.pr1$x["36",1]  ## Check that this returns 42
```

```{r I2d}
use <- complete.cases(socsupport[, 9:19])
use[36] <- FALSE
ss.pr <- prcomp(as.matrix(socsupport[use, 9:19]))
```

```{r I2e}
## Output from summary()
print(summary(ss.pr), digits=1)  # Compare contributions
```

```{r I2f}
comp <- as.data.frame(ss.pr$x[,1:6])
ss.lm <- lm(socsupport[use, "BDI"] ~ ., data=comp)
signif(round(coef(summary(ss.lm)),5), digits=3)
```

```{r I2g}
print(ss.pr$rotation[, 1], digits=2)
```

```{r 9_6, w=4, h=4.2, echo=FALSE,  out.width="42%"}
## Plot BDI against first principal component score
gph <- xyplot(BDI ~ ss.pr$x[ ,1], groups=gender, data=socsupport[use,],
par.settings=simpleTheme(pch=1:2), auto.key=list(columns=2))
bw9 <- list(pch=c(1,3), list(text=9, points=5))
update(gph, scales=list(tck=0.5), par.settings=bw9,
xlab ="1st principal component")
```

```{r 9_6, eval=F}
```

### Section 9.3: Cluster analysis
#### Subsection 9.3.1: Hierarchical Clustering
```{r I3_1a}
library(mvtnorm)
makeClust <- function(n=6, d1=4, d2=4, sigs=c(1, 1, 1, 1), seed=NULL){
  if(!is.null(seed))set.seed(seed)
  g1 <- rmvnorm(n, mean = c(-d1,d2), sigma=sigs[1]*diag(2))
  g2 <- rmvnorm(n, mean = c(d1,d2), sigma=sigs[2]*diag(2))
  g3 <- rmvnorm(n, mean = c(-d1,-d2), sigma=sigs[3]*diag(2))
  g4 <- rmvnorm(n, mean = c(d1,-d2), sigma=sigs[4]*diag(2))
  rbind(g1,g2,g3,g4)
}
```

```{r 9_7, echo=FALSE, w=6, h=1.9, mfrow=c(1,3), top=1.5}
## Code for the plots
datA <- makeClust(seed=35151)
datB <- makeClust(d2=16, seed=35151)
datC <- makeClust(d1=2,d2=2, seed=35151)
plot(datA, xlab="X1", ylab="X2", fg="gray")  
title(main="A: 4blobsA", adj=0, line=0.5, font.main=1)
## Repeat previous two lines for datB and datC
plot(datB, xlab="X1", ylab="X2", fg="gray")  
title(main="B: 4blobsB", adj=0, line=0.5, font.main=1)
plot(datC, xlab="X1", ylab="X2", fg="gray")  
title(main="C: 4blobsC", adj=0, line=0.5, font.main=1)
```

```{r 9_7, eval=F}
```

```{r, eval=xtras}
## Possible alternative
config <- c('Equidistant blobs', 'Pulled vertically', 'Closer centers')
dat123 <- cbind(as.data.frame(rbind(datA, datB, datC)), 
                              gp=factor(rep(1:3, rep(6*4,3)), labels=config))
xyplot(V2 ~ V1 | gp, data=dat123, scales=list(relation='free'),
       strip=strip.custom(factor.levels=config), between=list(x=0.5),
       par.settings=DAAG::DAAGtheme(color=F))
```

```{r 9_8, w=9, h=2.4, lwd=1, top=2, bot=-0.5, echo=FALSE, out.width='100%', message=FALSE, echo=FALSE}
## Code for single linkage plots: `?plot.hclust` gives help for the plot method
clusres_sing <- hclust(dist(datA), method="single")
par(fig=c(0,0.75,0,1))
plot(clusres_sing, sub="", xlab="", ylab="Distance joined", adj=0.5, 
     main="", fg="gray")
mtext('A: Single linkage cluster dendrogram, for 4blobsA layout', side=3, adj=0, 
      font=1, line=1, cex=1.15)
par(fig=c(0.72,1,0,1), new=TRUE)
membs <- cutree(clusres_sing, 4)
col4= RColorBrewer::brewer.pal(4,'Set1')
plot(datA, xlab="X1", ylab="X2", col=col4[membs], fg='gray', pch=membs+1)
mtext('B: 4blobsA, by color', side=3, adj=1.0, font=1, cex=1.15, line=1)
## To see plots from 'average' and 'complete' linkage methods,do:
# plot(hclust(dist(datB), method="average"))
# plot(hclust(dist(datC), method="complete"))
```

```{r 9_8, eval=F}
```

```{r 9_9, w=8, h=4.5, mgp=c(2.75,.75,0), mfrow=c(2,1), left=1, top=2, lwd=1, out.width='90%', message=FALSE, echo=FALSE}
## Dendrograms from data where blobs were pulled vertically
## Follow each use of `hclust()` with a `plot()` command
sclusres_sing <- hclust(dist(datB), method="single")
plot(sclusres_sing, sub="", xlab="", ylab="Distance Joined", main="")
title(main='A: Single linkage, blobs pulled vertically (4blobsB)', 
      adj=0, font.main=1)
sclusres_sing_s <- hclust(dist(scale(datA)), method="single")
plot(sclusres_sing_s, sub="", xlab="", ylab="Distance Joined", main="")
title(main='B: Single linkage, (4blobsB, rescaled to variance 1)', 
      adj=0, font.main=1)
# sclusres_avg_s <- hclust(dist(scale(datB)), method="average")
# #plot(sclusres_avg_s, sub="", xlab="", ylab="")
# sclusres_comp_s <- hclust(dist(scale(datB)), method="complete")
# #plot(sclusres_comp_s, sub="", xlab="", ylab="")
```

```{r 9_9, eval=F}
```

```{r 9_10, w=8, h=6.5, left=1, top=2, bot=-1, lwd=1, mfrow=c(3,1), out.width='90%', message=FALSE, echo=FALSE}
## Code. Follow each use of `hclust()` with a `plot()` command
clusres_sing2 <- hclust(dist(datC), method="single")
plot(clusres_sing2, sub="", xlab="", ylab="", cex=1.25, cex.main=1.65,
     main="A: Single linkage, closer clusters (4blobsC)", adj=0, font.main=1)
clusres_avg2 <- hclust(dist(datC), method="average")
plot(clusres_avg2, sub="", xlab="", ylab="", cex=1.25, cex.main=1.65,
     main="B: Average linkage, closer clusters (4blobsC)", adj=0, font.main=1)
clusres_comp2 <- hclust(dist(datC), method="complete")
plot(clusres_comp2, sub="", xlab="", ylab="", cex=1.25, cex.main=1.65, 
     main="C: Complete linkage, closer clusters (4blobsC)", adj=0, font.main=1)
```

```{r 9_10, eval=F}
```

#### Subsection 9.3.2: $k$-Means Clustering
```{r 9_11, echo=FALSE, w=6, h=2.7, mfrow=c(1,2), out.width="95%"}
set.seed(35151)
kdat <- makeClust(n=100, d1=5, d2=5, sigs=c(.5, .5, 6, 6))
plot(kdat, xlab="X1", ylab="X2", fg="gray")
kmres <- kmeans(kdat, 4, nstart=30)
plot(kdat, col=rainbow(4)[kmres$cluster], pch=kmres$cluster+1, 
     xlab="X1", ylab="X2", fg="gray")
```

```{r 9_11, eval=FALSE}
```

#####         Comments on $k$-means and hierarchical clustering
#### Subsection 9.3.3: Mixture model-based clustering

```{r 9_12, w=7, h=2.1, echo=F, top=1.5, mfrow=c(1,3), mgp=c(2.8,0.5,0), out.width='100%'}
## Code
plotMix2 <- function(taus=c(.5, .5), means=c(10,15), sds=c(3,1), xlims=c(0,20)){
  curve(taus[1]*dnorm(x, mean=means[1], sd=sds[1]) + 
          taus[2]*dnorm(x, mean=means[2], sd=sds[2]), 
        from=xlims[1], to=xlims[2], ylab="Density", fg="gray")
  curve(taus[1]*dnorm(x, mean=means[1], sd=sds[1]), 
        from=xlims[1], to=xlims[2], col="red", lty=2, add=TRUE, fg="gray")
  curve(taus[2]*dnorm(x, mean=means[2], sd=sds[2]), 
        from=xlims[1], to=xlims[2], col="blue", lty=3, add=TRUE, fg="gray")
}
plotMix2(taus=c(.2, .8))
plotMix2(taus=c(.5, .5))
plotMix2(taus=c(.9, .1))
```

```{r 9_12, eval=F}
```

```{r I3_3c}
library(teigen)
possml <- na.omit(DAAG::possum[,c(3,9:11)])
set.seed(513451)
gaus_fit <- teigen(possml[,2:4], models="UUUU", gauss=TRUE, verbose=FALSE, 
                    scale=FALSE)
```

```{r 9_13, w=5, h=2.0, top=1, left=2, echo=FALSE, out.width='60%'}
## BIC values are plotted against number of groups
gaus_fit$allbic
plot(gaus_fit$allbic, type="b", ylab="", xlab="Number of Groups", fg="gray")
mtext(side=2, line=3.5, "BIC", las=0)
axis(1, at=1:9, fg="gray")
```

```{r 9_13, eval=FALSE}
```

```{r I3_3d}
table(possml$Pop, gaus_fit$classification)
```

```{r 9_14, top=1.5, eval=FALSE, echo=FALSE}
par(fig=c(0, 0.5, 0.5, 1))
plot(gaus_fit, what="contour", xmarg=1, ymarg=2, draw.legend=FALSE, fg="gray")
## See ?teigen::plot.teigen for details of the plot command used here.
par(fig=c(0, 0.5, 0, 0.5), new=TRUE)
plot(gaus_fit, what="contour", xmarg=1, ymarg=3, draw.legend=FALSE, fg="gray")
par(fig=c(0.5, 1, 0, 0.5), new=TRUE)
plot(gaus_fit, what="contour", xmarg=2, ymarg=3, draw.legend=FALSE, fg="gray")
par(fig=c(0,1,0,1))
```

```{r 9_14, w=7.2, h=6.24, top=1.5, bot=0.75, mgp=c(2.25, 0.5, 0), echo=F, mfrow=c(2,2), out.width='80%'}
```

```{r 9_14, eval=FALSE}
```

#####             Issues of high parametrization and scaling
#### Subsection 9.3.4: Relationship between $k$-means and mixture models

### Section 9.4: Discriminant analysis
#### Subsection 9.4.1: Example -- plant architecture

```{r 9_15, echo=FALSE, w=7.2, h=3.5, left=-0.5, top=2, rt=3, las=0, mgp=c(1.8,0.5,0), tcl=-0.3, cex.lab=0.9, echo=FALSE, ps=10, mfrow=c(1,2), out.width="90%"}
xaxlab <- "Leaf width (mm)"
yaxlab <- "Leaf length (mm)"
blk <- adjustcolor("black", alpha.f=0.6)
leafshape17 <- DAAG::leafshape17
plot(bladelen ~ bladewid, data=leafshape17, pch=c(1,3)[arch+1],
cex=0.75, col=blk, xlab = xaxlab, ylab = yaxlab, fg="gray")
mtext(side=3,line=1.8,"A", adj=-0.2)
plot(bladelen ~ bladewid, data=leafshape17, log="xy", asp=1, col=blk,
     pch=c(1,3)[arch+1], cex=0.75, xlab = paste0(xaxlab,", log scale"),
     ylab=paste0(yaxlab,", log scale"), fg="gray", xaxt="n")
xpos <- c(2.5,5,10,20)
ypos <- c(10,20,40)
lxpos<-log10(xpos)
lypos<-log10(ypos)
axis(4, at = ypos, label = paste(round(lypos,1)), lwd=0, lwd.ticks=1)
axis(1, at = xpos, lwd=0, lwd.ticks=1)
axis(3, at = xpos, label = paste(round(lxpos,1)), lwd=0, lwd.ticks=1)
mtext(side=3,line=1.8,"B", adj=-0.2)
mtext(side = 4, line = 1.8, "log10(Leaf length)")
mtext(side = 3, line = 1.8, "log10(Leaf width)")
```

```{r I4_1b, eval=xtras, w=5, h=5, out.width="40%"}
leafshape17 <- DAAG::leafshape17
plot(bladelen ~ bladewid, data=leafshape17, pch=c(1,3)[arch+1])
## For panel B, specify log="xy" in the call to plot()
```

#####      Logistic regression, versus linear discriminant analysis
#### Subsection 9.4.2: Logistic regression
```{r I4_2a}
## Fit logistic regression model
leafshape17 <- DAAG::leafshape17
leaf17.glm <- glm(arch ~ logwid + loglen, family=binomial(link=logit),
data=leafshape17)
print(DAAG::sumry(leaf17.glm)$coef, digits=2)
```

#####                        Predictive accuracy
```{r I4_2b}
set.seed(29)
leaf17.cv <- DAAG::CVbinary(leaf17.glm)
tCV <- table(DAAG::leafshape17[["arch"]], round(leaf17.cv$cvhat))
rownames(tCV) <- colnames(tCV) <- c("0=Plagiotropic","1=Orthotropic")
cbind(tCV, "Proportion correct"=c(tCV[1,1], tCV[2,2])/(tCV[,1]+tCV[,2]))
```

```{r I4_2c}
round(unlist(leaf17.cv[c("acc.training","acc.cv")]),3)
```

#### Subsection 9.4.3: Linear discriminant analysis
```{r I4_3a}
suppressPackageStartupMessages(library(MASS))
## Discriminant analysis; data frame leafshape17 (DAAG)
leaf17.lda <- lda(arch ~ logwid+loglen, data=DAAG::leafshape17)
print(leaf17.lda)
```

#####                 Assessments of predictive accuracy
```{r I4_3b}
set.seed(29)
leaf17cv.lda <- lda(arch ~ logwid+loglen, data=leafshape17, CV=TRUE)
## the list element 'class' gives the predicted class
## The list element 'posterior' holds posterior probabilities
tab <- table(leafshape17$arch, leaf17cv.lda$class)
rownames(tab) <- colnames(tab) <- c("0=Plagiotropic","1=Orthotropic")
cbind(tab, "Proportion correct"=c(tCV[1,1], tCV[2,2])/(tCV[,1]+tCV[,2]))
cbind(tab, c(tab[1,1], class.acc=tab[2,2])/(tab[,1]+tab[,2]))
cat("Overall proportion correct =", sum(tab[row(tab)==col(tab)])/sum(tab), "\n")
```

#####   The function qda(), and other alternatives to lda()
#### Subsection 9.4.4: An example with more than two groups
```{r I4_4a}
## Linear discriminant calculations for possum data
possum <- DAAG::possum
possum.lda <- lda(site ~ hdlngth + skullw + totlngth + taill + footlgth +
                  earconch + eye + chest + belly, data=na.omit(possum))
# na.omit() omits any rows that have one or more missing values
```


```{r 9_16, echo=FALSE, w=5.5, h=5.5, left=-1, top=2, las=0, mgp=c(1.8,0.5,0), tcl=-0.3, cex.lab=0.9, echo=FALSE, mfrow=c(1,2), out.width="68%"}
PGtheme <- DAAG::DAAGtheme(color=TRUE, cex=0.6, pch=c(3,4,0,8,2,10,1), alpha=0.8)
possum <- DAAG::possum[!is.na(DAAG::possum$footlgth), ]
sitenames <- row.names(DAAG::possumsites)[c(1,2,4:6,3,7)]
possum.lda <- lda(site~hdlngth+skullw+totlngth+taill+
                  footlgth+earconch+eye+chest+belly, data=possum)
scores <- predict(possum.lda)$x[,1:3]
possum.splom <-
  splom(~ scores, groups = possum$site, varnames=c("LD1","LD2","LD3"), 
        par.settings=PGtheme, xlab=NULL)
update(possum.splom, auto.key=list(text=sitenames,  
                                  columns=4, between=1, between.columns=2))
```


```{r I4_4c, eval=FALSE}
plot(possum.lda, dimen=3, col=1:9)
# Scatterplot matrix - scores on 1st 3 canonical variates
# See `?plot.lda` for details of the generic lda plot function
```

```{r I4_4d}
## Linear discriminant calculations for possum data
print(possum.lda, digits=3)
```


### Section 9.5: *High-dimensional data --- RNA-Seq gene expression
#####        Setup for installing and using Bioconductor packages
```{r I5_1a, eval=FALSE}
## For latest details, see: https://www.bioconductor.org/install/
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install()
BiocManager::install('limma','multtest')
```

#####                *Brief note on mRNA technical issues
#### Subsection 9.5.1: Data and design matrix setup
```{r I5_1b}
counts <- DAAGbio::plantStressCounts
colSums(counts)
```

```{r I5_1c}
## Require at least 3 counts per million that are > 1
keep <- rowSums(counts)>=3
counts <- counts[keep,]
```

```{r I5_1d}
treatment <- factor(rep(c("CTL", "L", "D"), rep(3,3)))
design <- model.matrix(~0+treatment)
colnames(design) <- levels(treatment)
```

#####                  A two-dimensional representation
```{r I5_1e, eval=FALSE}
library(limma)
v <- voom(counts, design, plot=TRUE)
```

```{r 9_17, w=7.2, h=3.5, left=0.5, top=1, bot=0.5, ps=9, mgp=c(2.25,0.5,0), mfrow=c(1,2), echo=FALSE, out.width="90%"}
par(oma=c(0,0,1,0))
library(limma)
v <- voom(counts, design, plot=TRUE)
firstchar <- substring(colnames(counts),1,1)
plotMDS(counts, labels=paste0(firstchar, rep(1:3,3)), cex=0.8)
box(col="gray")
mtext(side=3, line=0.4, adj=0, "MDS summary plot")
mtext(side=3, line=-0.25, adj=0.105, "A", outer=TRUE)
mtext(side=3, line=-0.25, adj=0.605, "B", outer=TRUE)
```

```{r 9_17, eval=FALSE}
```

#####                         Fitting the model
```{r I5_1f}
fit <- lmFit(v, design)
```

```{r I5_1g}
contrs <- c("D-CTL", "L-CTL", "L-D")
contr.matrix <- makeContrasts(contrasts=contrs,
levels=levels(treatment))
fit2 <- contrasts.fit(fit, contr.matrix)
efit2 <- eBayes(fit2)
```

#### Subsection 9.5.2: From $p$-values to false discovery rate (FDR)
```{r I5_2a}
## First contrast only; Drought-CTL
print(round(topTable(efit2, coef=1, number=4),15), digits=3)
```

```{r I5_2b, eval=xtras}
round(sort(p.adjust(p=efit2$p.value[,1], method="BH"))[1:4], 15) # Not run
```

```{r I5_2c}
round(topTable(efit2, number=4), 15)
```

```{r I5_2d}
head(decideTests(fit2),5)
```

```{r I5_2e}
summary(decideTests(fit2))
## Try also
## summary(decideTests(fit2, p.value=0.001))
```


### Section 9.6: High dimensional data from expression arrays
#### Subsection 9.6.1: Molecular classification of cancer --- an older technology
#####    Breakdown of ALL B-type data, with one observation excluded
```{r I6_1a}
library(hddplot)
data(golubInfo)
with(golubInfo, table(cancer, tissue.mf))
```

```{r I6_1b, eval=xtras}
## Identify allB samples that are BM:f or BM:m or PB:m
subsetB <- with(golubInfo,
cancer=="allB" & tissue.mf%in%c("BM:f","BM:m","PB:m"))
## Separate off the relevant columns of the matrix Golub
## NB: variables (rows) by cases (columns)
GolubB <- with(golubInfo, Golub[, subsetB])
## Form vector that identifies these as BM:f or BM:m or PB:m
tissue.mfB <- with(golubInfo, tissue.mf[subsetB, drop=TRUE])
## Change the level names to leave out the colons
levels(tissue.mfB) <- list("b_f"="BM:f", "b_m"="BM:m", "PBm"="PB:m")
```

#### Subsection 9.6.2: Classifications and associated graphs
#####                   Preliminary data manipulation
```{r I6_2a, eval=xtras, w=7.5, h=3.5, mfrow=c(1,2), out.width="100%"}
## Display distributions for the first 20 observations
boxplot(data.frame(GolubB[, 1:20]))  # First 20 columns (observations)
## Random selection of 20 rows (features)
boxplot(data.frame(GolubB[sample(1:7129, 20), ]))
```

#####                           Flawed graphs
```{r 9_18, echo=FALSE, w=7.2, h=3.75, left=-1.5, top=2, cex.lab=0.9, ps=10, mfrow=c(1,2), cache=FALSE, results='hide', out.width="80%"}
colr <- c("red","blue","gray40", "magenta")
tissue.mf <- golubInfo[, "tissue.mf"]
cancer <- golubInfo[, "cancer"]
G.PBf <- Golub[, tissue.mf=="PB:f" & cancer=="allB", drop=FALSE]
set.seed(41)
rGolubB <- matrix(rnorm(prod(dim(GolubB))), nrow=dim(GolubB)[1])
rownames(rGolubB) <- rownames(Golub)
rG.PBf <- matrix(rnorm(prod(dim(G.PBf))), nrow=dim(G.PBf)[1])
plot2 <- function(x = GolubB, cl=tissue.mfB, x.omit=Golub.PBf, cl.omit="PBf", 
                  ncol = length(cl), nfeatures=12, device = "", seed = 37,
                  pretext="", colr=1:3, levnames = NULL,
                  ylab="2nd discriminant function"){
  cl <- factor(cl)
  if(!is.null(levnames))levels(cl) <- levnames
  ord15 <- orderFeatures(x, cl=cl)[1:nfeatures]
  dfB <- t(x[ord15, ])
  dfB.lda <-  lda(dfB, grouping=cl)
  scores <- predict(dfB.lda, dimen=2)$x
  df.PBf <- data.frame(t(x.omit[ord15, drop=FALSE]))
  colnames(df.PBf) <- colnames(dfB)
  scores.other <- predict(dfB.lda, newdata=df.PBf)$x
  scoreplot(list(scores=scores, cl=cl, other=scores.other, cl.other=cl.omit,                       nfeatures=nfeatures), prefix.title=pretext, adj.title=0, 
                 fg="gray", params=list(other=list(pch=4, cex=1.5)),
            xlab="1st discriminant function", ylab=ylab)
}
plot2(x = GolubB, cl = tissue.mfB, x.omit=G.PBf, cl.omit="PBf",
      nfeatures=15, device = "", seed = 37, ylab="2nd discriminant function",
      colr=colr, pretext="A: ALL B-cell:")
plot2(x = rGolubB, cl = tissue.mfB, x.omit=rG.PBf, cl.omit="Gp 4", 
     device = "", seed = 37, colr=colr, levnames = c("Gp 1", "Gp 2", "Gp 3"),
     pretext="B: Random data:", ylab="")
```

```{r 9_18, eval=F}
```

```{r I6_2c, eval=FALSE, size='normalsize'}
## Uses orderFeatures() (hddplot); see below
ord15 <- orderFeatures(GolubB, cl=tissue.mfB)[1:15]
```

```{r I6_2d, eval=FALSE, size='normalsize'}
## Panel A: Take 1st 15 features & transpose to observations by features
dfB15 <- data.frame(t(GolubB[ord15, ]))
dfB15.lda <-  MASS::lda(dfB15, grouping=tissue.mfB)
scores <- predict(dfB15.lda, dimen=2)$x
## Scores for the single PB:f observation
chooseCols <- with(golubInfo, tissue.mf=="PB:f"& cancer=="allB")
df.PBf <- data.frame(t(Golub[ord15, chooseCols, drop=FALSE]))
scores.PBf <- predict(dfB15.lda, newdata=df.PBf, dimen=2)$x
## Warning! The plot that now follows may be misleading!
## Use hddplot::scoreplot()
scoreplot(list(scores=scores, cl=tissue.mfB, other=scores.PBf, cl.other="PB:f"),
          fg="gray")
```

```{r I6_2e, eval=FALSE, size='normalsize'}
## Panel B: Repeat plot, now with random normal data
simscores <- simulateScores(nrow=7129, cl=rep(1:3, c(19,10,2)),
cl.other=4, nfeatures=15, seed=41)
# Returns list elements: scores, cl, scores.other & cl.other
scoreplot(simscores)
```

#### Subsection 9.6.3: The mean-variance relationship
```{r 9_19, w=7.2, h=3.5, left=0.5, top=1, bot=0.5, ps=9, mgp=c(2.25,0.5,0), mfrow=c(1,2), echo=FALSE, out.width="90%"}
par(oma=c(0,0,1,0))
designG <- model.matrix(~0+tissue.mfB)
colnames(designG) <- levels(tissue.mfB)
vG <- vooma(GolubB, designG, plot=TRUE)         # Panel A
plotMDS(vG, pch=unclass(tissue.mfB), cex=0.8)   # Panel B
leglabs <- c("BM:female","BM:male","PB:female")
legend(x="bottomright", bty="n", legend=leglabs, pch=1:3)
mtext(side=3, line=0.4, adj=0, "MDS summary plot")
mtext(side=3, line=-0.275, adj=0.085, "A", outer=TRUE)
mtext(side=3, line=-0.275, adj=0.585, "B", outer=TRUE)
```

```{r 9_19, eval=FALSE}
```

#####    Cross-validation to determine the optimum number of features
#####   Cross-validation for a range of choices of number of features
```{r I6_3b, cache=FALSE}
##  Cross-validation to determine the optimum number of features
## 10-fold (x4). Warning messages are omitted.
## Accuracy measure will be: tissue.mfB.cv$acc.cv
tissue.mfB.cv <- cvdisc(GolubB, cl=tissue.mfB, nfeatures=1:23,
nfold=c(10,4), print.progress=FALSE)
## Defective measures will be in acc.resub (resubstitution)
## and acc.sel1 (select features prior to cross-validation)
tissue.mfB.badcv <- defectiveCVdisc(GolubB, cl=tissue.mfB,
foldids=tissue.mfB.cv$folds, nfeatures=1:23)
##
## Calculations for random normal data:
set.seed(43)
rGolubB <- matrix(rnorm(prod(dim(GolubB))), nrow=nrow(GolubB))
rtissue.mfB.cv <- cvdisc(rGolubB, cl=tissue.mfB, nfeatures=1:23,
nfold=c(10,4), print.progress=FALSE)
rtissue.mfB.badcv <- defectiveCVdisc(rGolubB, cl=tissue.mfB,
nfeatures=1:23,
foldids=rtissue.mfB.cv$folds)
```


```{r 9_20, echo=FALSE, w=7.2, h=3.5, left=-1.5, top=1, cex.lab=0.9, echo=FALSE, ps=9, mfrow=c(1,2), out.width="100%"}
cv1 <- tissue.mfB.cv
badcv1 <- tissue.mfB.badcv
cv2 <- rtissue.mfB.cv
badcv2 <- rtissue.mfB.badcv
plot.acc <- function(cv=cv1, badcv=badcv1, nseq=NULL, badnseq=NULL, AB="",
                    ylab="Predictive accuracy", add.legend=TRUE, fg='gray'){
maxg <- min(c(length(badcv$acc.resub), length(cv$acc.cv)))
if(is.null(nseq))nseq <- 1:maxg
plot(nseq, badcv$acc.resub[1:maxg], ylim=c(0,1), type="n", fg=fg,
yaxs="i", xlab="Number of features selected", ylab=ylab)
par(xpd=T)
points(nseq, badcv$acc.resub[1:maxg], col=2, type="b", lty=2, pch=0, cex=0.8)
par(xpd=FALSE)
points(nseq, badcv$acc.sel1[1:maxg], col="gray40", pch=3, cex=0.8)
lines(lowess(nseq, badcv$acc.sel1[1:maxg], f=.325, iter=0),
col="gray40", lty=2)
points(nseq, cv$acc.cv[1:maxg], col="blue", pch=1, cex=0.8)
lines(lowess(nseq, cv$acc.cv[1:maxg], f=.325, iter=0),
col="blue",lwd=2)
xy <- par()$usr[c(1,3)]
if(add.legend)
legend(xy[1], xy[2], xjust=0, yjust=0,
legend=c("Resubstitution accuracy",
"Defective cross-validation",
"Cross-validation - select at each fold"),
lty=c(1,2,1), lwd=c(1,1,2), pch=c(0,3,1),
col=c("red","gray40","blue"), cex=0.8)
mtext(side=3,line=0.35, AB, adj=0)
}
cv1 <- tissue.mfB.cv
badcv1 <- rtissue.mfB.badcv
plot.acc(cv1, badcv1, AB="A: Golub data (as for Figure 9.97A)")
plot.acc(cv2, badcv2, ylab="", AB="B: Random data", add.legend=FALSE)
```


#####                          Which features?
```{r I6_3d}
genelist <- matrix(tissue.mfB.cv$genelist[1:3, ,], nrow=3)
tab <- table(genelist, row(genelist))
ord <- order(tab[,1], tab[,2], tab[,3], decreasing=TRUE)
tab[ord,]
```

#### Subsection 9.6.4: Graphs derived from the cross-validation process

```{r 9_21, echo=FALSE, w=3.5, h=3.6, left=-1, top=2, ps=10, tcl=-0.25, lwd=0.75, echo=FALSE, results="hide", out.width="40%"}
## Uses tissue.mfB.acc from above
tissue.mfB.scores <-
cvscores(cvlist = tissue.mfB.cv, nfeatures = 3, cl.other = NULL)
scoreplot(scorelist = tissue.mfB.scores, cl.circle=NULL,
prefix="B-cell subset -", fg='gray')
```


```{r 9_21, eval=FALSE}
```

#####                  The key role of cross-validation
#### Subsection 9.6.5: Estimating contrasts, and calculating False Discovery Rates
```{r I6_5a}
fitG <- lmFit(vG, designG)
contrs <- c("b_f-b_m", "b_f-PBm", "b_m-PBm")
contr.matrix <- makeContrasts(contrasts=contrs,
levels=levels(tissue.mfB))
fit2 <- contrasts.fit(fitG, contr.matrix)
fit2 <- eBayes(fit2)
```

#####           From $p$-values to false discovery rate (FDR)
```{r I6_5b}
print(topTable(fit2, number=5), digits=2)
```

```{r I6_5c}
summary(decideTests(fit2))
## Try also
## summary(decideTests(fit2, p.value=0.001))
```

#####                      Distributional extremes

### Section 9.7: Causal inference from observational data --- balance and matching
#### Subsection 9.7.1: Tools for the task
```{r I7_1a}
library(DAAG)
## Columns 4:7 are factors; columns 9:10 (re75 & re78) are continuous
propmat <- matrix(0, ncol=6, nrow=8)
dimnames(propmat) <- list(c("psid1", "psid2", "psid3", "cps1", "cps2", "cps3",
"nsw-ctl", "nsw-trt"), names(nswdemo)[c(4:7, 9:10)])
for(k in 1:8){
  dframe <- switch(k, psid1, psid2, psid3, cps1, cps2, cps3,
  subset(nswdemo, trt==0), subset(nswdemo, trt==1))
  propmat[k,] <- c(sapply(dframe[,4:7], function(x){
                   z <- table(x); z[2]/sum(z)}),
                   sapply(dframe[,9:10], function(x)sum(x>0)/sum(!is.na(x))))
}
```

```{r 9_22, w=7.2, h=5.0, left=-1, rt=-0.5, bot=1, ps=9, echo=FALSE, message=FALSE, warning=FALSE, out.width="100%"}
PGtheme <- DAAG::DAAGtheme(color=TRUE)
library(DAAG)
if(!require(grid))return("Package 'grid' is not installed -- cannot proceed")
dsetnames <- c("nsw-ctl", "nsw-trt", "psid1", "psid2", "psid3",
               "cps1", "cps2", "cps3")
colrs <- c("gray","black", PGtheme$superpose.line$col[1:3])
lty <- c(1,2,1,1,1)
lwd <- c(1,0.75,0.75,0.75,0.75)
denplot <-
  function(sel=c(1:2,6:8), yvar="re75", offset=30, ylim=c(0,1.75),
    from=NULL, at=c(.5,1,1.5), labels=paste(at), bw="nrd0",
    ylab="Density", takelog=TRUE, col.axis="black"){
      nzre <- unlist(lapply(list(subset(nswdemo, trt==0),
                                 subset(nswdemo, trt==1),
                                 psid1, psid2, psid3, cps1, cps2, cps3)[sel],
                            function(x){z <- x[,yvar]; z[z>0]}))
num <- unlist(lapply(list(subset(nswdemo, trt==0), subset(nswdemo, trt==1),
                          psid1, psid2, psid3, cps1, cps2, cps3),
                     function(x){z <- x[,yvar]; sum(z>0)}))
xy <- data.frame(nzre=nzre, fac = factor(rep(dsetnames[sel], num[sel]),
                 levels=dsetnames[sel]))
if(takelog) {
y <- log(xy$nzre+offset)
xlab <- paste("log(", yvar, "+", offset, ")", sep="")} else 
  {
  y <- xy$nzre
  xlab <- yvar
}
densityplot(~ y, groups=fac, data=xy, bw=bw, from=from,
  scales=list(y=list(at=at, labels=labels, col=col.axis), tck=0.25),
  plot.points=FALSE, col=colrs[1:5], lwd=lwd, lty=lty, 
  key=list(x=0.01, y=0.99, text=list(dsetnames[sel[3:5]]), col=colrs[3:5],
           cex=0.75, lines=list(lwd=rep(1.5,3)), between=1),
  par.settings=list(col=colrs, lty=lty, cex=0.75, lwd=lwd, 
                    fontsize=list(text=9, points=5)),
  fg="gray", ylim=ylim, ylab=ylab, xlab=xlab)
}
## Plot base graph; overlay with lattice graphs on same page
par(fig=c(0,1,0,1), mar=c(0,0,0,0))
plot(0:1,0:1, axes=FALSE, type="n", bty="n", xlab="", ylab="")
legend(x="top",legend=dsetnames[1:2], lty=1:2, lwd=c(1,0.75),
       col=colrs[1:2], bty="n", ncol=2, yjust=0.75)
print(denplot(), position=c(0, 0, 0.32, 0.505), newpage=FALSE)
print(denplot(1:5, ylab=" ", col.axis="white"),
      position=c(0.21, 0, .53, 0.505), newpage=FALSE)
print(denplot(ylab=" ", yvar="re78", col.axis="white"),
      position=c(0.47, 0, 0.79, 0.505), newpage=FALSE)
print(denplot(1:5, ylab=" ", yvar="re78", col.axis="white"),
      position=c(0.68, 0, 1, 0.505), newpage=FALSE)
## Age
print(denplot(yvar="age", takelog=FALSE, ylim=c(0,0.1), from=16,
      at=c(.02,.04,.06,.08), labels=c(".02",".04",".06",".08")),
      position=c(0, 0.475, 0.32, .98), newpage=FALSE)
print(denplot(1:5, yvar="age", takelog=FALSE, ylim=c(0,0.1), from=16,
      at=c(.02,.04,.06,.08), labels=c(".02",".04",".06",".08"),
      ylab=" ", col.axis="white"),
      position=c(0.21, 0.475, .53, .98), newpage=FALSE)
## educ
print(denplot(1:5, yvar="educ", takelog=FALSE, ylim=c(0,0.5), bw=0.5,
      at=c(.1,.2,.3,.4), ylab=" "),
      position=c(0.47, 0.475, .79, .98), newpage=FALSE)
print(denplot(yvar="educ", takelog=FALSE, ylim=c(0,0.75), bw=0.5,
      at=c(.1,.2,.3,.4), ylab=" ", col.axis="white"),
      position=c(0.68, 0.475, 1, .98), newpage=FALSE)
```

```{r 9_22, eval=F}
```

```{r I7_1c}
addControl <-
function(control, offset=30){
  nam <- deparse(substitute(control))
  if(nam=="nswdemo")nsw0 <- nswdemo else
    nsw0 <- rbind(control, subset(DAAG::nswdemo, trt==1))
  nsw0$z75 <- factor(nsw0$re75==0, labels=c("0",">0"))
  nsw0$ethnicid <- factor(with(nsw0, ifelse(black==1, "black",
    ifelse(hisp==1, "hisp", "other"))), levels=c("other","black","hisp"))
nsw0 <- nsw0[, -match(c("black","hisp"), names(nsw0))]
nsw0
}
```

```{r I7_1d}
## Create dataset that will be used for later analyses
nsw <- addControl(psid1)
nsw <- within(nsw, {re75log <- log(re75+30);
                    re78log <- log(re78+30);
                    trt <- factor(trt, labels=c("Control","Treat"))})
## A treated values only dataset will be required below
trtdat <- subset(nsw, trt=="Treat")
trtdat$pres74 <- factor(!is.na(trtdat$re74), labels=c("<NA>","pres"))
table(trtdat$pres74)
```


```{r 9_23, echo=FALSE, w=7.2, h=3.2, warning=FALSE, message=FALSE, out.width="90%"}
suppressPackageStartupMessages(library(mgcv))
missing.gam <- gam(pres74 ~ s(age) + s(educ) + ethnicid + marr + nodeg +
  z75 + s(re75log), data=trtdat, family='binomial')
bwplot(pres74~predict(missing.gam), data=trtdat, xlab="Propensity score",
  panel=function(x,y,...)
    {panel.bwplot(x,y); panel.points(x,as.numeric(y)-0.5,pch="|")})
```


```{r I7_1f}
with(trtdat, table(pres74,z75))
```

#### Subsection 9.7.2: Regression comparisons
#####                      Regression calculations
```{r I7_2a, echo=FALSE}
library(DAAG)
dimnam <- list(c("psid1","psid2","psid3","cps1","cps2","cps3","nswdemo"),
               c("Est.","exp(Est.)","CIlower","CIupper"))
tab <- matrix(nrow=7, ncol=4, dimnames=dimnam)
for (i in 1:7){
  joindata <- addControl(control=eval(as.name(dimnam[[1]][i])))
  nsw.gam <- gam(log(re78+30)~ trt + ethnicid + z75 + nodeg + s(age) +
                 s(educ) + log(re75+30), data=joindata)
  trtvec <- summary(nsw.gam)$p.table["trt", 1:2]
  tab[i,] <- c(trtvec[1], exp(trtvec[1]), exp(c(trtvec[1]+trtvec[2]*c(-1.96,1.96   ))))
}
tab
```

```{r I7_2b}
nsw.gam <- gam(log(re78+30)~ trt + ethnicid + z75 + nodeg + s(age) +
               s(educ) + log(re75+30), data=nsw)
```

#### Subsection 9.7.3: The use of scores to replace covariates
#### Subsection 9.7.4: Two-dimensional representation using randomForest proximities
```{r I7_4a}
suppressPackageStartupMessages(library(randomForest))
form <- trt ~ age + educ + ethnicid + marr + nodeg + z75 + re75log
nsw.rf <- randomForest(form, data=nsw, sampsize=c(297,297))
p.rf <- predict(nsw.rf,type="prob")[,2]
sc.rf <- log((p.rf+0.001)/(1-p.rf+0.001))
```


```{r 9_24, echo=FALSE, w=5, h=2.5, warning=FALSE, message=FALSE, out.width="90%"}
## Add to p.rf approx half of difference of max(p.rf) from 1
library(MatchIt)
form <- trt ~ age + educ + ethnicid + marr + nodeg + z75 + re75log
match.rf <- matchit(form, ratio=2, data=nsw, distance=sc.rf)
## ratio=2 matches 2 control units to each treatment unit
dat2RF <- match.data(match.rf, distance="PropScore")  ## Reduced size dataset
nsw2.rf <- randomForest(form, sampsize=c(297,297), data=dat2RF, proximity=T)
dmat2 <- 1 - nsw2.rf$proximity+0.004
pts <- cmdscale(dmat2)
rf2.mds <- MASS::isoMDS(dmat2, pts, trace=FALSE)
isoScores <- rf2.mds$points
p2.rf <- predict(nsw2.rf,type="prob")[,2]
cutpts <- c(0, round(quantile(p2.rf, c(1/3, 2/3)), 2), 1)
cutvals <- cut(p2.rf, breaks = cutpts, include.lowest = TRUE)
gph1 <- xyplot(isoScores[, 2] ~ isoScores[, 1] | cutvals,
              groups = dat2RF$trt, aspect = 1,
              scale = list(tck = 0.6), xlab = "Co-ordinate 1",
              ylab = "Co-ordinate 2", auto.key = list(columns = 2))
col8 <- DAAG::DAAGtheme(color=TRUE, fontsize=list(text=8, points=5), 
                        pch = 3:4, alpha=0.75)
print(update(gph1, par.settings=col8, layout = c(3, 1)))
```


```{r I7_4c}
omitn <- match(c("PropScore","weights","subclass"), names(dat2RF), nomatch=0)
matchISO.rf <-matchit(trt ~ age + educ + ethnicid + marr + nodeg + z75 +
                      re75log, ratio=1, data=dat2RF[,-omitn], distance=isoScores[,1])
## summary(match.rf,un=F,improvement=F)
## summary(match.rf, un=F, interactions=T, improvement=F)$sum.matched[,1:4]
## In the first place, look only at the first 4 columns
```

```{r I7_4d}
dat1RF <- match.data(matchISO.rf, distance="PropScore")
dat1RF.lm <- lm(re78log ~ trt, data = dat1RF, weights = weights)
library(sandwich)   # Allows use of `vcovCL()` from the `sandwich` package
lmtest::coeftest(dat1RF.lm, vcov. = vcovCL, cluster = ~subclass)
## Check for increase in number with non-zero earnings
dat1RF.glm <- glm(I(re78>0) ~ trt, data = dat1RF, weights = weights, 
                  family=binomial)
lmtest::coeftest(dat1RF.glm, vcov. = vcovCL, cluster = ~subclass)
```

#####               Derivation and investigation of scores
```{r I7_4e}
library(mgcv)
formG <- trt ~ ethnicid + marr+ z75 + s(age) + s(educ) + s(re75log)
nsw.gam <- gam(formG, family=binomial(link="logit"), data=nsw)
pred <- predict(nsw.gam, type='response')
table(nsw$trt, round(pred))
## Alternative
library(splines)  ## Fit normal cubic splines using splines::ns()
formNS <- trt ~ ethnicid + marr+ z75 + ns(age,2) +
ns(educ) + ns(re75log,3)
nsw.glm <- glm(formNS, family=binomial(link="logit"), data=nsw)
pred <- predict(nsw.glm, type='response')
table(nsw$trt, round(pred))
cbind(AIC(nsw.glm,nsw.gam), BIC(nsw.glm, nsw.gam))
```

```{r I7_4f, eval=FALSE}
## Include factor by factor and variable interactions with ethnicid
## and marr (Result not shown)
formGx <- trt ~ (ethnicid+marr+z75)^2 + s(age, by=ethnicid)+
                s(educ, by=ethnicid) + s(re75log,by=ethnicid)+
                s(age, by=marr)+ s(educ, by=marr) + s(re75log,by=marr)
nswx.gam <- gam(formula = formGx, data = nsw, family=binomial(link = "logit"))
predx <- predict(nswx.gam, type='response')
table(nsw$trt, round(predx))
AIC(nsw.glm,nsw.gam,nswx.gam)
```

```{r I7_4g}
library(MatchIt)
## Use data frame that omits re74. Otherwise matchit() will generate NAs
## where they occur in re74, even though re74 is not in the model formula.
nswG <- nsw[, c("trt","age","educ","ethnicid", "marr","nodeg","z75",
                "re75log","re78log","re78")]
formG <- trt ~ ethnicid + marr+ z75 + s(age) + s(educ) + s(re75log)
match.gam <- matchit(formula = formG, data = nswG, method = "nearest",
                     distance = "gam", link = "logit", reestimate=TRUE)
datG <- match.data(match.gam, distance="PropScore")
## Summary information
match.gam
## summary(match.gam,un=F,improvement=F)
## summary(match.gam, un=F, interactions=T, improvement=F)$sum.matched[,1:4]
## In the first place, look only at the first 4 columns
```


```{r 9_25, w=4.5, h=2.1, left=-1, rt=-0.5, top=1.5, ps=9, echo=FALSE, message=FALSE, warning=FALSE, out.width="80%"}
opar <- par(cex=0.75, bty="n")
plot(match.gam, type='jitter', interactive=F, 
     col=adjustcolor('blue',.4), pch=3)
box(fg="gray")
par(opar)
```



```{r 9_26, echo=FALSE, w=6.5, h=2.7, message=FALSE, warning=FALSE, fig.show="hold", results="hide", out.width="100%"}
suppressPackageStartupMessages(library(gridExtra))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(cobalt))
gg1<- cobalt::love.plot(match.gam, position="bottom", grid=TRUE,
                        star.char="",stars='raw') +
  ggtitle("A: Differences from balance") +
  theme(plot.title = element_text(hjust=0, vjust=0.5, size=11),
        plot.margin=unit(c(9,15,0,9), 'pt'))
sub <- match(with(subset(datG, trt=="Control"),subclass),
             with(subset(datG, trt=="Treat"),subclass))
datGpaired <- cbind(subset(datG, trt=="Treat"),
                    with(subset(datG, trt=="Control")[sub,],
                    cbind("Cre78log"=re78log,"CPropScore"=PropScore)))
gg2 <- ggplot(datGpaired)+
  geom_point(aes(PropScore,I(re78log-Cre78log)), size=1)+
  geom_smooth(aes(PropScore,I(re78log-Cre78log)), method = "gam", 
              formula = y ~ s(x, bs = "cs")) +
  xlab("Propensity score for treated")+
  ylab("Treatment vs control differences") +
  ggtitle("B: Treatment vs control differences") +
  theme(plot.title = element_text(hjust=0, vjust=0.5, size=11),
        plot.margin=unit(c(9,9,0,15), 'pt'))
grid.arrange(gg1, gg2, ncol=2) 
```


```{r 9_26, eval=FALSE}
```

```{r I7_4k}
library(sandwich)
datG.lm <- lm(re78log ~ trt, data = datG, weights = weights)
## With 1:1 matching, the weights argument is not really needed
## Print first two coefficients only. 
lmtest::coeftest(datG.lm, vcov. = vcovCL, cluster = ~subclass)[1:2,]
## Check number whose income was greater than 0
datG.glm <- glm(I(re78>0) ~ trt, data = datG, weights = weights, family=binomial)
lmtest::coeftest(datG.glm, vcov. = vcovCL, cluster = ~subclass)[1:2,]
```

#####                  Alternative matching approaches
#### Subsection 9.7.5: Coarsened exact matching
```{r I7_5a}
form <- trt ~ age + educ + ethnicid + marr + nodeg + z75 + re75log
match5.cem <- matchit(formula=form, data=nswG, method="cem", cutpoints=5)
datcem5 <- match.data(match5.cem)
match6.cem <- matchit(formula=form, data=nswG, method="cem", cutpoints=6)
datcem6 <- match.data(match6.cem)
## Show the effect of adding another cutpoint
match5.cem
match6.cem
```


```{r 9_27, echo=FALSE, w=6.5, h=2.7, warning=FALSE, quietly=FALSE, fig.show="hold", results="hide", out.width="100%"}
plt1 <- cobalt::love.plot(match5.cem, position="none", star.char="†",
          stars='raw', grid=T, title="A: CEM with five cutpoints") +
          theme(plot.title = element_text(hjust=0, vjust=0.5, size=11),
          plot.margin=unit(c(9,9,0,9), 'pt'))
plt2 <- cobalt::love.plot(match6.cem, position="none", star.char="†",
          stars='raw', grid=T, title="B: CEM with six cutpoints") +
          theme(plot.title = element_text(hjust=0, vjust=0.5, size=11),
          plot.margin=unit(c(9,9,0,9), 'pt'))
grid.arrange(plt1, plt2, ncol=2) + theme(plot.title = element_text(size=11))
```


```{r I7_5c, warning=FALSE}
library(sandwich)
datcem5.lm <- lm(re78log ~ trt, data = datcem5, weights = weights)
## The function vcovHC() provides cluster robust standard errors
lmtest::coeftest(datcem5.lm, vcov. = vcovHC)
## Estimate treatment effect on number with some earnings:
datcem6.glm <- glm(I(re78>0) ~ trt, data = datcem6, weights = weights,
family=binomial)
lmtest::coeftest(datcem6.glm, vcov. = vcovHC)
```


### Section 9.8: Multiple imputation
```{r I9a}
suppressPackageStartupMessages(library(mice))
Boys <- with(subset(mice::boys, age>=9), 
             data.frame(age=age, loghgt=log(hgt), logbmi=log(bmi), loghc=log(hc)))
(Pattern <- md.pattern(Boys, plot=F))
```

```{r I9b}
set.seed(31)       # Set to reproduce result shown
PatternB <- rbind(Pattern[-c(1,nrow(Pattern)), -ncol(Pattern)],
             c(0,1,1,1), c(0,1,0,0), c(0,0,1,0))
boys <- rbind(ic(Boys), 
              ampute(cc(Boys), pattern=PatternB, freq=c(.3,.15,.15,.2,.1,.1), 
                     prop=0.75)$amp)
md.pattern(boys, plot=FALSE)
```

```{r I8c}
set.seed(17)       # Set to reproduce result shown
out <- capture.output(        # Evaluate; send screen output to text string
  boys.mids <- mice(boys, method='pmm', m=8)  )
impDFs <- complete(boys.mids, action='all')   # Returns a list of m=8 dataframes
## Average over imputed dataframes (use for exploratory purposes only)
impArray <- sapply(impDFs, function(x)as.matrix(x), simplify='array')
boysAv <- as.data.frame(apply(impArray, 1:2, mean))  
```


```{r 9_28, echo=FALSE, w=6.5, h=3.15, mfrow=c(1,2), out.width="90%"}
## Check that straight line terms in `age` and `loghgt` are enough
suppressPackageStartupMessages(library(mgcv))
boys.gam <- gam(logbmi~s(age) + s(loghgt), data=boysAv)
plot(boys.gam, resid=TRUE, cex=1.5, shift=mean(predict(boys.gam)), fg='gray')
```


```{r I8d}
fits <- with(boys.mids, lm(logbmi~age+loghgt))
pool.coef <- summary(pool(fits))  # Include in table below
```

```{r I9d}
## 2) Regression that leaves out rows with NAs
omitNArows.coef <- coef(summary(lm(logbmi~age+loghgt, data=boys)))
## 3) Regression fit to average over data frames after imputation
boysAv.coef <- coef(summary(lm(logbmi~age+loghgt, data=boysAv)))
## 4) Fit to original data, with 36 rows had missing data
Orig.coef <- coef(summary(lm(logbmi ~ age+loghgt, data=Boys)))
```

```{r I9e}
ctab <- cbind(summary(pool(fits))[,2:3], omitNArows.coef[,1:2], boysAv.coef[,1:2], 
      Orig.coef[,1:2])
tab <- setNames(cbind(ctab[,c(1,3,5,7)], ctab[,c(2,4,6,8)]),
                paste0(rep(c('Est','SE'), c(4,4)), rep(1:4, 2)))
round(tab,3)
```

#####           Time series cross-sectional data -- an example
```{r I9f}
airquality <- datasets::airquality
airq <- cbind(airquality[, 1:4], day=1:nrow(airquality))
  # 'day' (starting May 1) replaces columns 'Month' & 'Day')
## Replace `Ozone` with `rt4ozone`:
airq <- cbind(rt4ozone=airq$Ozone^0.25, airq[,-1])
```

```{r I9g, eval=FALSE}
## Generate the scatterplot matrix, now with `rt4ozone` replacing `Ozone`
smoothPars <- list(col.smooth='red', lty.smooth=2, spread=0)
car::spm(airq, cex.labels=1.2, regLine=FALSE, col='blue', 
         oma=c(1.95,3,4, 3), gap=.25, smooth=smoothPars)
```

```{r I9h, echo=T, results='hide'}
airq.imp <- mice(airq, m=20, print=FALSE)
  ## 20 imputations shows up issues of concern very clearly
```


```{r 9_29, w=9, h=5.25, mar=c(2.1,3.1, 0.6,0.6), echo=FALSE, results="hide", warning=F, out.width='100%'}
## Code for figure
out <- micemd::overimpute(airq.imp)
```


```{r 9_29, eval=F}
```

#####                        Some further points

### Section 9.9: Further reading
#####             Data with more variables than observations
#####                          Causal inference
#####                        Multiple imputation

### Section 9.10: Exercises

9.3
```{r I7a, eval=xtras}
library(DAAG)
oz::oz(sections=c(3:5, 11:16))
names(possumsites)[1:2] <- c("long", "lat")
with(possumsites, {
points(long, lat);
text(long, lat, row.names(possumsites), pos=c(2,4,2,2,4,2,2))
})
```
9.7
```{r I7b}
data(wine, package='gclus')
mat <- with(wine, 
  round(1-cor(cbind(Alcohol, Malic, Magnesium, Phenols, Flavanoids)),2))
colnames(mat) <- rownames(mat) <- 1:5
print(mat)
```
9.9a
```{r I7c, eval=xtras}
`confusion` <-
function(actual, predicted, digits=4){
  tab <- table(actual, predicted)
  confuse <- apply(tab, 1, function(x)x/sum(x))
  print(round(confuse, digits))
  acc <- sum(tab[row(tab)==col(tab)])/sum(tab)
  invisible(print(c("Overall accuracy" = round(acc,digits))))
}
data(Vehicle, package="mlbench")
lhat <- MASS::lda(Class ~ ., data=Vehicle, CV=TRUE)$class
qhat <- MASS::qda(Class ~ ., data=Vehicle, CV=TRUE)$class
DAAG::confusion(Vehicle$Class, lhat)
DAAG::confusion(Vehicle$Class, qhat)
randomForest::randomForest(Class ~ ., data=Vehicle, CV=TRUE)
```

```{r I7d, eval=xtras}
## 9.9c
Vehicle.lda <- MASS::lda(Class ~ ., data=Vehicle)
twoD <- predict(Vehicle.lda)$x
ggplot2::quickplot(twoD[,1], twoD[,2], color=Vehicle$Class,
                   geom=c("point","density2d"))
```
9.10
```{r I7e, eval=xtras}
library(ape); library(MASS)
library(DAAGbio)
primates.dna <- as.DNAbin(primateDNA)
primates.dist <- dist.dna(primates.dna, model="K80")
primates.cmd <- cmdscale(primates.dist)
eqscplot(primates.cmd)
rtleft <- c(4,2,4,2)[unclass(cut(primates.cmd[,1], breaks=4))]
text(primates.cmd, labels=row.names(primates.cmd), pos=rtleft)
```

```{r I7f, eval=xtras}
d <- dist(primates.cmd)
sum((d-primates.dist)^2)/sum(primates.dist^2)
```
9.11
```{r I7g, eval=xtras}
library(DAAG)
pacific.dist <- dist(x = as.matrix(rockArt[-c(47,54,60,63,92),28:641]), 
                     method = "binary")
sum(pacific.dist==1)/length(pacific.dist)
plot(density(pacific.dist, to = 1))
## Check that all columns have at least one distance < 1
symmat <- as.matrix(pacific.dist)
table(apply(symmat, 2, function(x) sum(x<1)))
pacific.cmd <- cmdscale(pacific.dist)
pacific.sam <- sammon(pacific.dist)
```
9.15
```{r eval=F}
Wine <- setNames(cbind(stack(wine, select=2:14), rep(wine[,-1], 13)),
                 c("value", "measure", "Class"))
bwplot(measure ~ value, data=Wine)
```

```{r eval=F}
wine.pr <- prcomp(wine[,-1], scale=TRUE)
round(wine.pr$sdev,2)
t(round(wine.pr$rotation[,1:2],2))
scores <- as.data.frame(cbind(predict(wine.pr), Class=wine[,1]))
xyplot(PC2 ~ PC1, groups=Class, data=scores, aspect='iso', 
       par.settings=simpleTheme(pch=16), auto.key=list(columns=3))
```

```{r eval=F}
library(MASS)
wine.lda <- lda(Class ~ ., data=wine)
wineCV.lda <- lda(Class ~ ., data=wine, CV=T)
t(round(wine.lda$scaling,2))
tab <- table(wine$Class, wineCV.lda$class, 
             dnn=c('Actual', 'Predicted'))   
tab
setNames(round(1-sum(diag(tab))/sum(tab),4), "CV error rate")
scores <- as.data.frame(cbind(predict(wine.lda)$x, Class=wine[,1]))
xyplot(LD2 ~ LD1, groups=Class, data=scores, aspect='iso', 
       par.settings=simpleTheme(pch=16), auto.key=list(columns=3))
```

```{r eval=F}
wine$Class <- factor(Wine$Class)
wine.rf <- randomForest(x=wine[,-1], y=wine$Class)
```

```{r, eval=T}
if(file.exists("/Users/johnm1/pkgs/PGRcode/inst/doc/")){
code <- knitr::knit_code$get()
txt <- paste0("\n## ", names(code),"\n", sapply(code, paste, collapse='\n'))
writeLines(txt, con="/Users/johnm1/pkgs/PGRcode/inst/doc/ch9.R")
}
```
