---
title: "Create static html files for posting to GitHub"
author: "John Maindonald"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Create static html files}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r echo=F}
options(rmarkdown.html_vignette.check_title = FALSE)
```

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette is primarily intended as a reminder for the maintainer,
of steps that may be needed when changes are made, in particular
when a vignette is changed or added.  A commit and push to GitHub 
will be needed for changes to take effect. 

Details of using pkgdown to set up the web page can be found at
https://pkgdown.r-lib.org/articles/pkgdown.html#configuration

```{r setupCode, eval=F}
usethis::use_pkgdown()
pkgdown::build_site()
usethis::use_pkgdown_github_pages()
```

Rebuild just the home page and the CSS by running `build_home_index(); init_site()`. 

## Setup so that download from GitHub makes vignettes available

```{r exampleCode, eval=F}
tools::buildVignettes(dir = ".", tangle=FALSE)
if(!dir.exists('inst'))dir.create("~/pkgs/PGRcode/inst")   
if(!dir.exists('inst/doc'))dir.create("~/pkgs/PGRcode/inst/doc") 
## On installation, files in 'inst/doc/' are copied to 'doc/'
file.copy(dir("vignettes", full.names=TRUE), "inst/doc", overwrite=TRUE)
## For an individual file, go e.g., for GITready.Rmd
## tools::buildVignette('~/pkgs/PGRcode/vignettes/GITready.Rmd',
##                      dir =  "inst/doc", tangle=F)
```
