---
title: "PGRcode"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{PGRcode}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
suppressPackageStartupMessages(library(PGRcode))
```

This website is designed  to make available code and other
supplementary materials for a new Cambridge University Press text,
due to appear in print in May or June 2024.

"A Practical Guide to Data Analysis Using R -- An Example-Based Approach",  
by John H Maindonald, W John Braun, and Jeffrey L Andrews.   

Chapter and appendix headings are:

* Chapter 1: Learning from data, and tools for the task
* Chapter 2: Generalizing from models
* Chapter 3: Multiple linear regression
* Chapter 4: Exploiting the linear model framework
* Chapter 5: Generalized linear models and survival analysis
* Chapter 6: Time series models
* Chapter 7: Multilevel models, and repeated measures
* Chapter 8: Tree-based Classification and Regression
* Chapter 9: Multivariate data exploration and discrimination
* Appendix A: The R System – A Brief Overview

Major changes from the earlier text are 

* There is much more limited tutorial content on R syntax and semantics 
(readers are encouraged to refer to the wealth of online material.)
* There is more critical attention to $p$-values
* The use of Bayes Factors, contrasting them with $p$-values for the
different and complementary perspective that they provide.
* Ihere is greater attention to effect size issues.
* There is extended commentary on the insights that
replicability/reproducability studies provide.
* Simulation and resampling approaches get more extended use -- as
sources of insight, as devices for building intuition, and as mechanisms
for obtaining sampling distributions when theoretical results are not
available.
* There is new content on quantile regression with automatic
choice of smoothing parameter, and on fitting monotonic increasing or
decreasing response curves as specific forms as shape constrained
additive response.
* The treatment of models that allow for extra-binomial or
extra-Poisson variation has been substantially extended.
* Exponential time series (ETS) get greater attention, especially for
their use in forecasting.  Modeling of seasonal terms now gets
attention.
* A new section treats hierarchical and other forms of clustering.
* The treatment of causal inference from observational data has been
greatly extended.
* There is some limited attention to the use of multiple imputation to
fill in missing values in data where some observations are incomplete.

More detailed commentary is provided in "A Practical Guide -- Style and Content".
