---
title: "Use of Code Files"
author: "John Maindonald"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Packages

Should a package not be available on CRAN, it will in many instances
be possible to obtain the latest development version from a GitHub site.

# Alternative outputs and themes for the R markdown (`.Rmd`) files

Primarily, it is expected that users will run individual code chunks
one or more at a time as they work through the text.  In this mode
of working, needed R packages can be installed as required, as 
calculations proceed.

# Options and issues for processing a complete R Markdown file 

An alternatives is to process the whole file at one time, either by supplying the file name as argument to `knitr()`, or by clicking ‘knit’ in Rmarkdown with the file open.  Set the output option in the YAML header at the start of the file as required, to create HTML or PDF or Word output.  Additionally, the Markdown file that is created as an intermediary can be saved, if required.  An example of YAML settings is:

```
---
title: "Use of Code Files"
author: "John Maindonald"
date: "`r Sys.Date()`"
output: 
  html_document
---
```

Available alternatives for the final three lines include:

```
output:
  word_document: default
  html_document: default
  pdf_document: default
---
```

```
output:
  html_document:
    keep_md: true
---
```

The following is used in the Rmd files included here:

```
output:
  prettydoc::html_pretty:
    theme: cayman
    highlight: github
---
```

```{r}
library(knitr)
opts_chunk[['set']](cache.path='cache-', out.width="80%", fig.width=6, fig.height=6, fig.align='center', size="small", ps=10, strip.white = TRUE,
                    comment=NA, width=70, tidy.opts = list(replace.assign=FALSE))
```

```{r CodeControl, echo=FALSE}
## xtras=FALSE
xtras <- TRUE
library(knitr)
## opts_chunk[['set']](results="asis")
opts_chunk[['set']](eval=T)
## opts_chunk[['set']](eval=F)
```

## Code markup

The following shows an approach to placing graphs side by side
```{r, fig.width=4, fig.height=4,fig.show="hold", fig.align="default", out.width="48.5%"}
z <- seq(-4,4,length=101)
plot(z, dnorm(z), type="l", ylab="Normal density")
plot(z, dt(z, df=5), type="l", ylab="t-statistic density, 5 df")
```

## Using the `Hmisc::knitrSet()` function

This function creates a setup where 

* Space around the graph can be adjusted by setting the chunk 
options `bot`, `left`, `top`, and `rt`;
* Commonly used base graphics parameters can be supplied as chunk options.
These include `lwd`, `mgp`, `las`, `tcl`, `axes`, `xpd`, `mfg`.
* Default settings can be supplied for a number of chunk options.

See `?Hmisc::knitrSet` for details.  With R markdown files, the setting
`fig.align="default"` is required.  Either do not change from the
default globally, or accompany `fig.show="hold"` whenever used with
`fig.align="default"`, thus:

```{r, fig.width=4, fig.height=4,fig.show="hold", fig.align="default", out.width="48.5%"}
Hmisc::knitrSet(basename="tmp", lang='markdown', fig.path="figs/g", w=6, h=6)
z <- seq(-4,4,length=101)
plot(z, dnorm(z), type="l", ylab="Normal density")
plot(z, dt(z, df=5), type="l", ylab="t-statistic density, 5 df")
```

