---
title: "Chapter 2: Generalizing from models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Ch2: Generalizing . . .}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r echo=F}
options(rmarkdown.html_vignette.check_title = FALSE)
```

On options for working with the code see the vignettes
[Ch1-Learning](PGRcode/Ch1-Learning.html) and
[UsingCode](PGRcode/UsingCode.html).

```{r CodeControl, echo=FALSE}
options(rmarkdown.html_vignette.check_title = FALSE)
xtras <- F
library(knitr)
## opts_chunk[['set']](results="asis")
opts_chunk[['set']](eval=F)
```

##### Packages required (plus any dependencies)
DAAG MASS qra investr HistData BHH2 xtable BayesFactor boot zoo boot 
MCMCpack,

Additionally, knitr and Hmisc are required in order to process the Rmd source file. 


```{r setup, cache=FALSE, echo=FALSE}
Hmisc::knitrSet(basename="models", lang='markdown', fig.path="figs/g", w=7, h=7)
oldopt <- options(digits=4, formatR.arrow=FALSE, width=70, scipen=999)
library(knitr)
## knitr::render_listings()
opts_chunk[['set']](cache.path='cache-', out.width="80%", fig.align="center", 
                    fig.show='hold', size="small", ps=10, strip.white = TRUE,
                    comment=NA, width=70, tidy.opts = list(replace.assign=FALSE))
```

### Section 2.1 Model assumptions
#### Subsection 2.1.1: Inferences are never assumption free
#### Subsection 2.1.2: Has account been taken of all relevant effects?
```{r B1_2a}
## Tabulate by Admit and Gender
byGender <- 100*prop.table(margin.table(UCBAdmissions, margin=1:2), margin=2)
round(byGender,1)
```

```{r B1_2b}
## Admission rates, by department
pcAdmit <- 100*prop.table(UCBAdmissions, margin=2:3)["Admitted", , ]
round(pcAdmit,1)
```

```{r 2_1, echo=FALSE, w=4.5, h=2.25, left=-1, rt=1, top=1.5, mgp=c(1.5,0.5,0), ps=7, out.width="85%"}
applied <- margin.table(UCBAdmissions, margin=2:3)
pcAdmit <- 100*prop.table(UCBAdmissions, margin=2:3)["Admitted", , ]
      byGender <- 100*prop.table(margin.table(UCBAdmissions,
      margin=1:2), margin=2)
dimnam <- dimnames(UCBAdmissions)
mfStats <- data.frame(Admit=c(pcAdmit[1,],pcAdmit[2,]),
  Applicants=c(applied[1,], applied[2,]),
  mf=factor(rep(dimnam[['Gender']],c(6,6)),
  levels=dimnam[['Gender']]), Department=rep(dimnam[["Dept"]],2))
xlim <- c(0, max(mfStats$Admit)*1.025)
ylim <- c(0, max(mfStats$Applicants)*1.075)
plot(Applicants~Admit, data=mfStats, type="h",lwd=2, xlim=xlim, ylim=ylim,
     fg="gray", cex.lab=1.2, col=rep(c("blue","red"),rep(6,2)),
     xlab="UCB Admission rates (%), 1973", ylab="Number of applicants")
pcA <- rbind(pcAdmit[1,], apply(pcAdmit,2, mean)+2, pcAdmit[2,], rep(NA,6))
pcA[2,3] <- pcA[2,3]+1
appA <- rbind(applied[1,], apply(applied,2, mean)+80,
              applied[2,], rep(NA,6))
deptNam <- dimnam[[3]]
for(j in 1: ncol(appA)) lines(pcA[,j], appA[,j], col="gray", lwd=0.8)
points(pcA[2,],appA[2,], pch=16, cex=1.1, col="white")
text(pcA[2,],appA[2,],deptNam, cex=0.85)
##
par(xpd=TRUE)
text(byGender[1,1:2], rep(par()$usr[4],2)+0.5*strheight("^"),
     labels=c("^","^"), col=c("blue","red"),cex=1.2,srt=180)
text(byGender[1,], par()$usr[4]+1.4*strheight("A"),
     labels=paste(round(byGender[1,],1)),cex=0.85)
text(byGender[1,1:2]+c(-3.5,3.5), rep(par()$usr[4],2)+2.65*strheight("A"),
     labels=c("All males","All females"), pos=c(4,2), cex=1.2)
par(xpd=FALSE)
abline(h=200*(0:4),col="lightgray",lty="dotted")
abline(v=20*(0:4),col="lightgray",lty="dotted")
legend("topleft", col=c('blue','red'),lty=c(1,1), lwd=0.75, cex=0.9,
       y.intersp=0.65, legend=c("Males","Females"),bty="n")
```

```{r 2_1, eval=F}
```

```{r B1_2d}
## Calculate totals, by department, of males & females applying
margin.table(UCBAdmissions, margin=2:3)
```

#### Subsection 2.1.3: The limitations of models
#### Subsection 2.1.4: Use the methodology that best suits the task in hand?

### Section 2.2: t-statistics, binomial proportions, and correlations
#### Subsection 2.2.1: One- and two-sample t-tests
#### Subsection 2.2.2: A two-sample comparison

```{r B2_2a}
stats2 <- sapply(DAAG::two65,
                 function(x) c(av=mean(x), sd=sd(x), n=length(x)))
pooledsd <- sqrt( sum(stats2['n',]*stats2['sd',]^2)/sum(stats2['n',]-1) )
stats2 <- setNames(c(as.vector(stats2), pooledsd),
                   c('av1','sd1','n1','av2','sd2','n2','pooledsd'))
print(stats2, digits=4)
```

```{r B2_2b}
with(DAAG::two65, t.test(heated, ambient, var.equal=TRUE))
```

#####                      When is pairing helpful?

```{r 2_2, echo=FALSE, w=6, h=2.75, top=1, rt=2, las=0, tcl=-0.4, results='hide', ps=8, mfrow=c(1,2), out.width="85%"}
titl <- paste("Second versus first member, for each pair.  The first",
"\npanel is for the elastic band data. The second (from",
"\nDarwin) is for plants of the species Reseda lutea")
oldpar <- par(pty="s")
on.exit(par(oldpar))
DAAG::onesamp(dset = DAAG::pair65, x = "ambient", y = "heated",
  xlab = "Amount of stretch (ambient)",
  ylab = "Amount of stretch (heated)", fg='gray')
## Data set mignonette holds the Darwin (1877) data on Reseda lutea.
## Data were in 5 pots, holding 5,5,5,5,4 pairs of plants respectively.
DAAG::onesamp(dset = DAAG::mignonette, x = "self", y = "cross",
  xlab = "Height of self-fertilised plant", ylab =
  "Height of cross-fertilised plant", dubious = 0, cex=0.7, fg='gray')
```

```{r 2_2, eval=F}
```

#####            What if the standard deviations are unequal?
#### Subsection 2.2.3: The normal approximation to the binomial
#### Subsection 2.2.4: The Pearson or product--moment correlation

```{r B2_4a}
## Pearson correlation between `body` and `brain`: Animals
Animals <- MASS::Animals
rho <- with(Animals, cor(body, brain))
## Pearson correlation, after log transformation
rhoLogged <- with(log(Animals), cor(body, brain))
## Spearman rank correlation
rhoSpearman <- with(Animals, cor(body, brain, method="spearman"))
c(Pearson=round(rho,2), " Pearson:log values"=round(rhoLogged,2),
  Spearman=round(rhoSpearman,2))
```

### Section 2.3 Extra-binomial and extra-Poisson variation

```{r B3_1a}
maleDF <- data.frame(number=0:12, freq=unname(qra::malesINfirst12[["freq"]]))
N <- sum(maleDF$freq)
pihat <- with(maleDF, weighted.mean(number, freq))/12
probBin <- dbinom(0:12, size=12, prob=pihat)
rbind(Frequency=setNames(maleDF$freq, nm=0:12),
      binomialFit=setNames(probBin*N, nm=0:12),
      rawResiduals = maleDF$freq-probBin*N,
      SDbinomial=sqrt(probBin*(1-probBin)*N)) |>
  formatC(digits=2, format="fg") |> print(digits=2, quote=F, right=T)
```

```{r B3_1b, echo=F}
## Fit binomial and betabinomial distributions.
suppressPackageStartupMessages(library(gamlss))
doBI <- gamlss(cbind(number, 12-number)~1, weights=freq,
               family=BI, data=maleDF, trace=FALSE)
doBB <- gamlss(cbind(number, 12-number)~1, weights=freq,
                     family=BB, data=maleDF, trace=FALSE)
```

```{r 2_3, w=9, h=6, ps=10, top=2, left=-2, bot=0.5, mfrow=c(2,3), echo=FALSE,mar=c(3.1,2.6,3.1,0.6), mgp=c(2,.5,0), out.width="100%"}
set.seed(29)
rqres.plot(doBI, plot.type='all', type="QQ", main=""); box(col='white')
mtext(side=3, line=0.5, "A: Binomial model, Q-Q", adj=0, cex=1.25)
rqres.plot(doBI, plot.type='all', type="wp", main=""); box(col='white')
## Plots C, D, E, F: Set object name; set`type="wp" (C, E, F), or`"QQ"` (D)
mtext(side=3, line=0.5, "B: Binomial, worm plot 1", adj=-0.05, cex=1.25)
rqres.plot(doBI, plot.type='all', type="wp", main=""); box(col='white')
mtext(side=3, line=0.5, "C: Binomial, worm plot 2", adj=-0.05, cex=1.25)
rqres.plot(doBB, plot.type='all', type="QQ", main="", ylab=''); box(col='white')
mtext(side=3, line=0.5, "D: BB model, Q-Q", adj=0, cex=1.25)
rqres.plot(doBB, plot.type='all', type="wp", main="", ylab=''); box(col='white')
mtext(side=3, line=0.5, "E: BB, worm plot 1", adj=0, cex=1.25)
rqres.plot(doBB, plot.type='all', type="wp", main="", ylab=''); box(col='white')
mtext(side=3, line=0.5, "F: BB, worm plot 2", adj=0, cex=1.25)
```

```{r 2_3, eval=FALSE}
```

```{r B3_1d}
aicStat <- AIC(doBI, doBB)
rownames(aicStat) <-
  c(doBI="Binomial", doBB="Betabinomial")[rownames(aicStat)]
aicStat$dAIC <- with(aicStat, round(AIC-AIC[1],1))
aicStat
```

```{r B3_1e}
## Numbers of accidents in three months, with Poisson fit
machinists <- data.frame(number=0:8, freq=c(296, 74, 26, 8, 4, 4, 1, 0, 1))
N <- sum(machinists[['freq']])
lambda <- with(machinists, weighted.mean(number, freq))
fitPoisson <- dpois(0:8, lambda)*sum(machinists[['freq']])
rbind(Frequency=with(machinists, setNames(freq, number)),
      poissonFit=fitPoisson) |>
  formatC(digits=2, format="fg") |> print(quote=F, digits=2, right=T)
```

```{r B3_1f, echo=FALSE}
doPO <- gamlss(number~1, weights=freq,
                family=PO, data=machinists, trace=FALSE)
doNBI <- gamlss(number~1, weights=freq,
                  family=NBI, data=machinists, trace=FALSE)
```

```{r 2_4, w=9, h=3, ps=10, left=-2, top=2, bot=0.5, mfrow=c(1,3), echo=FALSE,mar=c(3.1,2.6,3.1,0.6), mgp=c(2,.5,0), out.width="100%"}
set.seed(23)
rqres.plot(doPO, plot.type='all', type="QQ", main=""); box(col='white')
## Repeat, changing the argument, for remaining plots
mtext(side=3, line=0.5, "A: Poisson, Q-Q plot", adj=0, cex=1.25)
rqres.plot(doPO, plot.type='all', type="wp", main="", ylab=''); box(col='white')
mtext(side=3, line=0.5, "B: Poisson, worm plot", adj=0, cex=1.25)
rqres.plot(doNBI, plot.type='all', type="wp", main="", ylab='')
mtext(side=3, line=0.5, "C: NBI, worm plot", adj=0, cex=1.25); box(col='white')
```

```{r 2_4, eval=F}
```

### Subsection 2.3.2: *Technical details -- extra-binomial or extra-Poisson variation

```{r B3_1i}
sigma <- exp(coef(doBB, "sigma"))
cat("Phi =", (1+12*sigma)/(1+sigma))
```

```{r B3_1j}
mu <- exp(coef(doNBI, "mu"))
sigma <- exp(coef(doNBI, "sigma"))
cat("Phi =", (1+sigma*mu))
```

### Section 2.4 Contingency tables

```{r B4a}
## 'Untreated' rows (no training) from psid3, 'treated' rows from nswdemo
nswpsid3 <- rbind(DAAG::psid3, subset(DAAG::nswdemo, trt==1))
degTAB <- with(nswpsid3, table(trt,nodeg))
# Code 'Yes' if completed high school; 'No' if dropout
dimnames(degTAB) <- list(trt=c("PSID3_males","NSW_male_trainees"),
                         deg =c("Yes","No"))
degTAB
```

```{r B4b}
# To agree with hand calculation below, specify correct=FALSE
chisq.test(degTAB, correct=FALSE)
```

#####               The mechanics of the chi-squared test
#####        An example where a chi-squared test may not be valid

```{r B4c}
## Engine man data
engineman <- matrix(c(5,3,17,85), 2,2)
chisq.test(engineman)
```

#####                 Rare and endangered plant species

```{r B4d}
fisher.test(engineman)
```

```{r B4e}
## Enter the data thus:
rareplants <- matrix(c(37,190,94,23, 59,23,10,141, 28,15,58,16), ncol=3,
  byrow=TRUE, dimnames=list(c("CC","CR","RC","RR"), c("D","W","WD")))
```

```{r B4f}
(x2 <- chisq.test(rareplants))
```

#####  Examination of departures from a consistent overall row pattern

```{r B4h, echo=-1}
options(digits=3)
## Expected values
x2$expected
```

```{r B4i}
options(digits=2)
## Standardized residuals
residuals(x2)
```

#####                       Interpretation issues

### Section 2.5 Issues for Regression with a single explanatory variable
#### Subsection 2.5.1: Iron slag example --- check residuals with care!

```{r 2_5, echo=FALSE,  w=5.5, h=3.35, cex.lab=0.9,  left=-2, top=1,  rt=0.5, mgp=c(1.6,0.5,0), echo=FALSE, ps=10, mfrow=c(2,3), out.width="100%"}
leg <- c("A: Fitted line", "B: Residuals from line", "C: Variance check")
ord <- order(DAAG::ironslag[["magnetic"]])
ironslag <- DAAG::ironslag[ord,]
slagAlpha.lm <- lm(chemical~magnetic, data=ironslag)
resval <- residuals(slagAlpha.lm)
fitchem <- fitted(slagAlpha.lm)
sqrtabs2 <- sqrt(abs(resval))
plot(chemical~magnetic, xlab = "Magnetic", ylab = "Chemical",
     pch = 1, data=ironslag, fg="gray")
lines(fitchem~ironslag[["magnetic"]])
mtext(side = 3, line = 0.25, leg[1], adj=-0.1, cex=0.925)
scatter.smooth(resval~ironslag[["magnetic"]], lpars=list(col="red"), span=0.8,
               xlab = "Magnetic", ylab = "Residual", fg="gray")
mtext(side = 3, line = 0.25, leg[2], adj = -0.1, cex=0.925)
scatter.smooth(sqrtabs2 ~ fitchem, lpars=list(col="red"), span=0.8,
xlab = "Predicted chemical", fg="gray",
ylab = expression(sqrt(abs(residual))))
mtext(side = 3, line = 0.25, leg[3], adj = -0.1, cex=0.8)
## Diagnostics from fit using loess()
leg2 <- c("D: Smooth, using loess()",
          "E: Residuals from smooth",
          "F: Variance check")
slag.loess <- loess(chemical~magnetic, data=ironslag, span=0.8)
resval2 <- slag.loess[["residuals"]]
fitchem2 <- slag.loess[["fitted"]]
sqrtabs2 <- sqrt(abs(resval2))
plot(chemical~magnetic, xlab = "Magnetic", ylab = "Chemical",
pch = 1, data=ironslag, fg="gray")
lines(fitchem2 ~ ironslag[["magnetic"]], col="red")
mtext(side = 3, line = 0.25, leg2[1], adj=-0.1, cex=0.925)
scatter.smooth(resval2~ironslag[["magnetic"]], span=0.8,
lpars=list(col="red"),
xlab = "Magnetic", ylab = "Residual", fg="gray")
mtext(side = 3, line = 0.25, leg2[2], adj = -0.1, cex=0.925)
scatter.smooth(sqrtabs2 ~ fitchem2, lpars=list(col="red"),
span=0.8, xlab = "Predicted chemical", fg="gray",
ylab = expression(sqrt(abs(residual))))
mtext(side = 3, line = 0.25, leg2[3], adj = -0.1, cex=0.925)
```

```{r 2_5, eval=F}
```

#### Subsection 2.5.2: The analysis of variance table

```{r B5_2a}
roller.lm <- lm(depression ~ weight, data=DAAG::roller)
anova(roller.lm)
```

#### Subsection 2.5.3: Outliers, influence, and robust regression

```{r 2_6, w=2.75, h=2.4, left=-1, lwd=0.75, ps=10, tcl=-0.25, lwd=0.75, echo=FALSE, fig.pos='h', out.width="42%"}
softbacks <- DAAG::softbacks
x <- softbacks[,"volume"]
y <- softbacks[,"weight"]
u <- lm(y ~ x)
yhat <- predict(u)
res <- resid(u)
r <- with(softbacks, cor(x, y))
xlim <- with(softbacks, range(volume))
xlim[2] <- xlim[2]+diff(xlim)*0.08
plot(y ~ x, xlab = "Volume (cc)", xlim=xlim,
data=softbacks, ylab = "Weight (g)", pch = 4,
ylim = range(c(y, yhat)), cex.lab=0.9, fg="gray")
abline(u$coef[1], u$coef[2], lty = 1)
bottomright <- par()$usr[c(2, 3)]
chw <- par()$cxy[1]
chh <- par()$cxy[2]
z <- summary(u)$coef
btxt <- c(paste("a =", format(round(z[1, 1], 1)),
"  SE =", format(round(z[1, 2], 1))),
paste("b =", format(round(z[2, 1], 2)),
"  SE =", format(round(z[2, 2], 2))))
legend(bottomright[1],  bottomright[2],
legend=btxt, xjust=1, yjust=0, cex=0.8, bty="n")
```

```{r 2_6, eval=F}
```

```{r B5_3a}
softbacks.lm <- lm(weight ~ volume, data=DAAG::softbacks)
print(coef(summary(softbacks.lm)), digits=3)
```

```{r 2_7, w=7.25, h=1.65, mgp=c(1.95,0.5,0), top=1, left=-0.5, ps=10, mfrow=c(1,4), echo=FALSE, out.width="100%"}
plot(softbacks.lm, fg="gray",
caption = c("A: Residuals vs Fitted", "B: Normal Q-Q",
"C: Scale-Location", "", "D: Resids vs Leverage"))
```

```{r 2_7, eval=F}
```

#####                         Robust regression
#### Subsection 2.5.4: Standard errors and confidence intervals
#####            Confidence intervals and tests for the slope

```{r B5_4a}
SEb <- coef(summary(roller.lm))[2, 2]
coef(roller.lm)[2] + qt(c(0.025,.975), 8)*SEb
```

#####         SEs and confidence intervals for predicted values

```{r B5_4b}
## Code to obtain fitted values and standard errors (SE, then SE.OBS)
fit.with.se <- predict(roller.lm, se.fit=TRUE)
fit.with.se$se.fit                                            # SE
sqrt(fit.with.se[["se.fit"]]^2+fit.with.se$residual.scale^2)  # SE.OBS
```

```{r B5_4c}
predict(roller.lm, interval="confidence", level=0.95)
predict(roller.lm, interval="prediction", level=0.95)  # CI for a new observation
```

```{r 2_8, w=6.0, h=2.85, left=-1, top=1.5, mfrow=c(1,2), ps=10, tcl=-0.25, lwd=0.75, echo=FALSE, warning=FALSE, out.width="80%"}
## Depression vs weight, with 95\% pointwise bounds for both
## the fitted line and predicted values
investr::plotFit(roller.lm, interval="both", col.conf="red", fg="gray")
mtext(side=3,line=0.75, "A: Lawn roller data", cex=1.2, adj=-0.25)
## Male child vs father height, Galton's data
galtonMales <- subset(HistData::GaltonFamilies, gender=="male")
galton.lm <- lm(childHeight~father, data=galtonMales)
investr::plotFit(galton.lm, interval="both", col.conf="red", hide=FALSE,
                 col=adjustcolor('black',alpha=0.5), fg="gray")
mtext(side=3,line=0.75, "B: Son vs father heights", cex=1.2, adj=-0.25)
```

```{r 2_8, eval=F}
```

```{r B5_4e, echo=FALSE}
round(coef(summary(galton.lm))[,-4],2)
```

#####             Implications for design

```{r 2_9, echo=FALSE, w=5, h=3, bot=2, lwd=0.75, ps=10, out.width="75%"}
panelci<-function(data,...)
{
nrows<-list(...)$nrows
ncols<-list(...)$ncols
if(ncols==1)axis(2, lwd=0, lwd.ticks=1)
if(ncols==1)axis(1, lwd=0, lwd.ticks=1) else
axis(3, lwd=0, lwd.ticks=1)
x<-data$stretch; y<-data$distance
u <- lm(y ~ x)
upred <- predict(u, interval="confidence")
ci <- data.frame(fit=upred[,"fit"],lower=upred[,"lwr"], upper=upred[,"upr"])
ord<-order(x)
lines(x[ord], ci[["fit"]][ord], lty=1, lwd=2)
lines(lowess(x[ord], ci[["upper"]][ord]), lty=2, lwd=2, col="grey")
lines(lowess(x[ord], ci[["lower"]][ord]), lty=2, lwd=2, col="grey")
}
elastic1 <- DAAG::elastic1
elastic2 <- DAAG::elastic2
xy<-rbind(elastic2,elastic1)
nam <- c("Range of stretch 30-65 mm","Range of stretch 42-54 mm")
trial<-rep(nam, c(dim(elastic2)[1],dim(elastic1)[1]))
xlim<-range(elastic2$stretch)
ylim<-range(elastic2$distance)
xy<-split(xy,trial)
xy<-lapply(1:length(xy),function(i){c(as.list(xy[[i]]), list(xlim=xlim,
ylim=ylim))})
names(xy) <- nam
DAAG::panelplot(xy,panel=panelci,totrows=1,totcols=2,
                par.strip.text=list(cex=.9), oma=c(4,4,2.5,2), fg='gray')
mtext(side = 2, line = 3.35, "Distance moved (cm)", cex=1.1, las=0)
mtext(side=1,line=3,"Amount of stretch (mm)", cex=1.1)
```

```{r 2_9, eval=F}
```

```{r 2_10, echo=FALSE, w=5.5, h=2.7, left=-1, top=1.5, cex.lab=0.75, ps=10, mfrow=c(1,2), out.width="85%"}
## There are two regression lines!
pair65 <- DAAG::pair65
bothregs <- function(x=pair65[, "ambient"], y=pair65[, "heated"],
  xlab="Stretch (band at ambient)", ylab = "Stretch (heated band)", pch=16){
    plot(y ~ x, xlab = xlab, ylab = ylab, pch = pch, fg="gray")
    topleft <- par()$usr[c(1, 4)] + c(0.5, -0.5) * par()$cxy
    text(topleft[1], topleft[2], paste("r =", round(cor(x, y), 2)), adj = 0)
    u1 <- lm(y ~ x)
    abline(u1$coef[1], u1$coef[2])
    u2 <- lm(x ~ y)
    abline( - coef(u2)[1]/coef(u2)[2], 1/coef(u2)[2], lty = 2)
}
bothregs()
mtext(side = 3, line = 0.5, "A", adj = 0)
bothregs(x=trees[, "Girth"], y=trees[, "Height"],
         xlab="Girth (in)", ylab <- "Height (ft)", pch=16)
mtext(side = 3, line = 0.5, "B", adj = 0)
```

#### Subsection 2.5.5: There are two regression lines!

```{r 2_10, eval=F}
```

#####                An alternative to a regression line
#### Subsection 2.5.6: Logarithmic and Power Transformations

```{r 2_11, w=6, h=3.6, left=-1, top=1.5, ps=10, tcl=-0.25, mfrow=c(2,3), echo=FALSE, out.width="80%"}
## Logarithmic and Power Transformations
DAAG::powerplot(expr="sqrt(x)", xlab="")
DAAG::powerplot(expr="x^0.25", xlab="", ylab="")
DAAG::powerplot(expr="log(x)", xlab="", ylab="")
DAAG::powerplot(expr="x^2")
DAAG::powerplot(expr="x^4", ylab="")
DAAG::powerplot(expr="exp(x)", ylab="")
```

```{r 2_11, eval=F}
```

#####  General power transformations --- Box-Cox and Yeo-Johnson
#### Subsection 2.5.7: General forms of nonlinear response
#### Subsection 2.5.8: Size and shape data -- allometric growth

```{r 2_12, w=3.25, h=2.7, top=1.25, rt=1, ps=10, tcl=-0.25, echo=FALSE, out.width="42%"}
## Heart weight versus body weight, for 30 Cape fur seals.
g2.12 <- function()
{
cfseal <- DAAG::cfseal
x <- log(cfseal[,"weight"])
y <- log(cfseal[, "heart"])
ylim <- log(c(82.5,1100))
xlim <- log(c(17,180))
ylab <- "Heart weight (g, log scale)"
xlab <- "Body weight (kg, log scale)"
xtik <- c(20,40,80,160)
ytik <- c(100,200,400,800)
plot(x, y, xlab = xlab, ylab = ylab, axes = F, xlim =
xlim, ylim = ylim, pch = 16, cex=0.85, fg="gray", cex.lab=1.1)
axis(1, at = log(xtik), labels = paste(xtik), lwd=0, lwd.ticks=1)
axis(2, at = log(ytik), labels = paste(ytik), lwd=0, lwd.ticks=1)
box(col="gray")
form1 <- formula(y ~ x)
u <- lm(form1, data = cfseal)
abline(u$coef[1], u$coef[2])
usum <- summary(u)$coef
options(digits=3)
print(usum)
cwh <- par()$cxy
eqn <- paste("log y =", round(usum[1, 1], 2), " [",
round(usum[1, 2], 2), "] +", round(usum[2, 1], 3),
" [", round(usum[2, 2], 3), "] log x")
mtext(side=3, line=1.15, eqn, adj = 0.4, cex = 0.8)
mtext(side=3, line=0.25, "(Values in square brackets are SEs)", adj = 0.4, cex = 0.8)
}
g2.12()
```

```{r 2_12, eval=F}
```

#####                  The  allometric growth equation

```{r B5_4f}
options(scipen=4)
cfseal.lm <- lm(log(heart) ~ log(weight), data=DAAG::cfseal)
print(coef(summary(cfseal.lm)), digits=4)
```

### Section 2.6 Empirical assessment of predictive accuracy
#### Subsection 2.6.1: The training/test approach, and cross-validation
#####               Cross-validation -- a tutorial example

```{r 2_13, w=3.0, h=2.65, echo=FALSE, out.width="42%"}
houseprices <- DAAG::houseprices
df <- DAAG::CVlm(houseprices, form.lm = formula(sale.price ~ area),m=3,printit=F,plotit=FALSE)
panelfun <- function(x,y,subscripts,groups, ...){
  lattice::panel.superpose(x,y,subscripts,groups, ...)
  lattice::panel.superpose(x,df[["cvpred"]],subscripts,groups,type="b", cex=0.5, ...)
}
gph <- lattice::xyplot(sale.price ~ area, groups=fold, data=df, pch=1:3, panel=panelfun)
parset <- DAAG::DAAGtheme(color=T, lty=1:3, pch=1:3, lwd=2)
keylist <- list(lines=TRUE, columns=3, between.columns=1.5, between=1, cex=0.85)
update(gph, par.settings=parset, auto.key=keylist)
```

```{r 2_13, eval=F}
```

```{r B6_1b}
set.seed(29)        # Generate results shown
rand <- sample(rep(1:3, length=15))
## sample() randomly permutes the vector of values 1:3
for(i in 1:3) cat(paste0(i,":"), (1:15)[rand == i],"\n")
```

```{r B6_1c}
houseprices <- DAAG::houseprices
row.names(houseprices) <- (1:nrow(houseprices))
DAAG::CVlm(houseprices, form.lm = formula(sale.price ~ area), plotit=FALSE)
```

```{r B6_1d}
## Estimate of sigma^2 from regression output
houseprices <- DAAG::houseprices
houseprices.lm <- lm(sale.price ~ area, houseprices)
summary(houseprices.lm)[["sigma"]]^2
```

#### Subsection 2.6.2: Bootstrapping in regression

```{r B6_2a}
houseprices <- DAAG::houseprices
houseprices.lm <- lm(sale.price ~ area, houseprices)
print(coef(summary(houseprices.lm)),digits=2)
```

```{r B6_2b}
houseprices.fn <-
  function (houseprices, index,
            statfun=function(obj)coef(obj)[2]){
            house.resample <- houseprices[index, ]
            house.lm <- lm(sale.price ~ area, data=house.resample)
            statfun(house.lm)    # slope estimate for resampled data
            }
```

```{r B6_2c, warning=FALSE, message=FALSE}
set.seed(1028)     # use to replicate the exact results below
library(boot)      # ensure that the boot package is loaded
## requires the data frame houseprices (DAAG)
(houseprices.boot <- boot(houseprices, R=999, statistic=houseprices.fn))
```

```{r B6_2d}
statfun1200 <- function(obj)predict(obj, newdata=data.frame(area=1200))
price1200.boot <- boot(houseprices, R=999, statistic=houseprices.fn,
statfun=statfun1200)
boot.ci(price1200.boot, type="perc") # "basic" is an alternative to "perc"
```

```{r 2_14, echo=FALSE, w=6, h=2.65, top=1, cex.lab=0.9, ps=10, mfrow=c(1,2), echo=FALSE, lwd=0.75, out.width="90%"}
set.seed(1111)
library(boot)
par(las=0)
houseprices2.fn<-function (houseprices,index){
house.resample<-houseprices[index,]
house.lm<-lm(sale.price~area,data=house.resample)
houseprices$sale.price-predict(house.lm,houseprices)
# resampled prediction errors
}
houseprices <- DAAG::houseprices
n<-nrow(houseprices)
R <- 199    ## Will obtain 199 estimates of prediction error
houseprices.lm<-lm(sale.price~area,data=houseprices)
houseprices2.boot<-boot(houseprices, R=R, statistic=houseprices2.fn)
house.fac<-factor(rep(1:n,rep(R,n)))
plot(house.fac,as.vector(houseprices2.boot$t),
     ylab="", xlab="House", fg="gray")
mtext(side=2, line=2, "Prediction Errors")
mtext(side = 3, line = 0.5, "A", adj = 0)
boot.se <- apply(houseprices2.boot$t,2,sd)
model.se <- predict.lm(houseprices.lm,se.fit=T)$se.fit
plot(boot.se/model.se, ylab="", xlab="House",pch=16, fg="gray")
mtext(side=2, line=2.0, "Ratio of SEs\nBootstrap to Model-Based", cex=0.9)
mtext(side = 3, line = 0.5, "B", adj = 0)
abline(1,0)
```

```{r 2_14, eval=F}
```

#####                             Commentary

### Section 2.7 One- and two-way comparisons
#### Subsection 2.7.1: One-way comparisons
```{r B7_1a}
tomato <- data.frame(Weight = c(1.5, 1.9, 1.3, 1.5, 2.4, 1.5,   # Water
                                1.5, 1.2, 1.2, 2.1, 2.9, 1.6,   # Nutrient
                                1.9, 1.6, 0.8, 1.15, 0.9, 1.6), # Nutrient+24D
  trt = factor(rep(c("Water", "Nutrient", "Nutrient+24D"), c(6, 6, 6))))
## Make `Water` the first level of trt.  In aov or lm calculations, it is
## then taken as the baseline or reference level.
tomato$trt <- relevel(tomato$trt, ref="Water")
```

```{r 2_15a, w=3.5, h=2, echo=FALSE, lwd=0.5, out.width="55%", message=FALSE}
## A: Weights of tomato plants (g)
library(lattice, quietly=TRUE)
gph <- stripplot(trt~Weight, aspect=0.35, scale=list(tck=0.6), data=tomato)
update(gph, scales=list(tck=0.4), cex=0.9, col="black", xlab="",
       main=list('A: Weights of tomato plants (g)', y=0, cex=1.1))
```

```{r 2_15a, eval=F}
```

```{r 2_15b, w=9, h=3.5, ps=9, echo=FALSE, out.width="100%"}
## B: Summarize comparison between LSD and Tukey's HSD graphically
tomato.aov <- aov(Weight ~ trt, data=tomato)
DAAG::onewayPlot(obj=tomato.aov)
title(main="B: LSD, compared with Tukey HSD", adj=0.1, outer=T,
      line=-1.0, font.main=1, cex.main=1.25)
```

```{r 2_15b, eval=F}
```

```{r B7_1d, eval=xtras}
BHH2::anovaPlot(tomato.aov)
```

#####                   The analysis of variance table

```{r B7_1e}
## Do analysis of variance calculations
anova(tomato.aov)
```

#####                  Other multiple comparison tests
#### Subsection 2.7.2: Regression versus qualitative comparisons -- issues of power

```{r 2_16, w=2.85, h=2.65, left=-1, mgp=c(2,0.5,0), echo=FALSE, out.width="42%"}
gph <- DAAG::simulateLinear(alpha=0.6, seed=17, aspect='iso')
update(gph, par.settings=DAAG::DAAGtheme(color=FALSE, alpha=0.4))
```

```{r 2_16, eval=F}
```

#### Subsection 2.7.3: *Severe multiplicity --- the false discovery rate
#####          *Microarrays and alternatives --- technical note
#####                   The false discovery rate (FDR)

```{r B7_3a}
coralPval <- DAAG::coralPval
pcrit <- c(0.05, 0.02, 0.01, 0.001)
under <- sapply(pcrit, function(x)sum(coralPval<=x))
```

```{r B7_3b}
expected <- pcrit*length(coralPval)
```

```{r B7_3c}
fdrtab <- data.frame(Threshold=pcrit, Expected=expected,
Discoveries=under, FDR=round(expected/under, 4))
print(xtable::xtable(fdrtab), include.rownames=FALSE, hline.after=FALSE)
```

```{r B7_3d}
fdr <- p.adjust(coralPval, method="BH")
```

```{r B7_3e}
fdrcrit <- c(0.05, 0.04, 0.02, 0.01)
under <- sapply(fdrcrit, function(x)sum(coralPval<=x))
setNames(under, paste(fdrcrit))
```

```{r B7_3f, echo=FALSE}
p45 <- (fdrcrit[1]*under[1]-fdrcrit[2]*under[2])/(under[1]-under[2])
p12 <- (fdrcrit[3]*under[3]-fdrcrit[4]*under[4])/(under[3]-under[4])
```

#### Subsection 2.7.4: Data with a two-way structure, i.e., two factors

```{r 2_18, echo=FALSE, w=7.5, h=3.4, left=-1, top=1.5, tcl=-0.2, cex=0.9, echo=FALSE, ps=9, out.width="100%"}
par(fig=c(0.525,1,0,1), mgp=c(1.5,0.4,0))
lev <- c("F10", "NH4Cl", "NH4NO3", "F10 +ANU843",
         "NH4Cl +ANU843", "NH4NO3 +ANU843")
rice <- within(DAAG::rice, trt <- factor(trt, levels=lev))
with(rice, interaction.plot(fert, variety, ShootDryMass, fg="gray",
     legend = FALSE, xlab="Fertiliser", cex.lab=0.95, mex=0.65))
xleg <- par()$usr[2]
yleg <- par()$usr[4] - 0.72 * diff(par()$usr[3:4])
leginfo <- legend(xleg, yleg, bty = "n", legend = levels(rice$variety),
                  col = 1, lty = 2:1, lwd=1, xjust = 1, cex = 0.8,
                  y.intersp=0.8)$rect
text(leginfo$left + 0.5 * leginfo$w, leginfo$top, "  variety",
      adj = 0.5, cex = 0.8)
mtext(side=3, line=0.65, cex=0.9, adj=-0.15, "B")
gph <- dotplot(trt ~ ShootDryMass, pch=1, cex=0.9, las=2,
               xlab="Shoot dry mass (g)", data=rice,
               panel=function(x,y,...){panel.dotplot(x,y,...)
                 av <- sapply(split(x,y),mean);
                 ypos <- unique(y)
                 lpoints(ypos~av, pch=3, col="gray40", cex=1.25)},
               main=list("A", cex=0.88, just="left", x=0.1, y=-0.7, font=1))
pars <-  DAAG::DAAGtheme(fontsize=list(text=9, points=6), color=FALSE)
print(update(gph, scales=list(tck=0.5), par.settings=pars, aspect=0.9),
      position=c(-0.065,0.0,0.6,1), newpage=FALSE)
```

```{r 2_18, eval=F}
```

#### Subsection 2.7.5: Presentation issues

### Section 2.8 Data with a nested variation structure
#### Subsection 2.8.1: Degrees of freedom considerations
#### Subsection 2.8.2: General multi-way analysis of variance designs

### Section 2.9 Bayesian estimation -- further commentary and approaches
#### Subsection 2.9.1: Bayesian estimation with normal priors and likelihood
#### Subsection 2.9.2: Further comments on Bayes Factors
#####             The Sellke calibration upper limit
#####       A note on the Bayesian Information Criterion

```{r B9_2a}
pval <- c(.05,.01,.001); np <- length(pval)
Nval <- c(4,6,10,20,40,80,160); nlen <- length(Nval)
## Difference in BIC statistics, interpreted as Bayes factor
t2BFbic <- function(p,N){t <- qt(p/2, df=N-1, lower.tail=FALSE)
                         exp((N*log(1+t^2/(N-1))-log(N))/2)}
bicVal <- outer(pval, Nval, t2BFbic)
## Bayes factor, calculated using BayesFactor::ttest.tstat()
t2BF <- function(p, N){t <- qt(p/2, df=N-1, lower.tail=FALSE)
          BayesFactor::ttest.tstat(t=t, n1=N, simple=TRUE, rscale = "medium")}
BFval <- matrix(nrow=np, ncol=nlen)
for(i in 1:np)for(j in 1:nlen) BFval[i,j] <- t2BF(pval[i], Nval[j])
cfVal <- rbind(BFval, bicVal)[c(1,4,2,5,3,6),]
dimnames(cfVal) <- list(
  paste(rep(pval,rep(2,np)), rep(c("- from ttest.tstat", "- from BIC"),np)),
        paste0(c("n=",rep("",nlen-1)),Nval))
round(cfVal,1)
```

#### Subsection 2.9.3: Bayesian regression estimation using the MCMCpack package

```{r B9_3a, message=FALSE, warning=FALSE}
suppressPackageStartupMessages(library(MCMCpack))
roller.mcmc <- MCMCregress(depression ~ weight, data=DAAG::roller)
summary(roller.mcmc)
```

```{r 2_19, w=6, h=4.0, left=-1, top=1, ps=10, tcl=-0.25, lwd=0.75, mar=c(4.1,4.1,3.1,2), mfrow=c(2,3), echo=FALSE, out.width="100%"}
mat <- matrix(c(1:6), byrow=TRUE, ncol=2)
layout(mat, widths=rep(c(2,1.1),3), heights=rep(0.9,8))
  # NB: widths & heights are relative
plot(roller.mcmc, auto.layout=FALSE, ask=FALSE, col="gray", fg="gray")
```

```{r 2_19, eval=F}
```

### Section 2.10: Recap
#####                 Regress $y$ on $x$, or $x$ on~$y$?

### Section 2.11: Further reading

### Exercises (2.12)

```{r B12_22}
## UCBAdmissions is in the datasets package
## For each combination of margins 1 and 2, calculate the sum
UCBtotal <- apply(UCBAdmissions, c(1,2), sum)
```

```{r B12_2b}
apply(UCBAdmissions, 3, function(x)(x[1,1]*x[2,2])/(x[1,2]*x[2,1]))
```

```{r B12_3AB}
tabA <- array(c(30,30,10,10,15,5,30,10), dim=c(2,2,2))
tabB <- array(c(30,30,20,10,10,5,20,25), dim=c(2,2,2))
```

```{r B12_5}
z.transform <- function(r) .5*log((1+r)/(1-r))
z.inverse <- function(z) (exp(2*z)-1)/(exp(2*z)+1)
  possum.fun <- function(data, indices) {
    chest <- data$chest[indices]
    belly <- data$belly[indices]
    z.transform(cor(belly, chest))}
possum.boot <- boot::boot(DAAG::possum, possum.fun, R=999)
z.inverse(boot.ci(possum.boot, type="perc")$percent[4:5])
 # The 4th and 5th elements of the percent list element
 # hold the interval endpoints. See ?boot.ci
```

```{r B12_11, fig.width=5, fig.height=4, out.width="40%"}
with(pressure, MASS::boxcox(pressure ~ I(1/(temperature+273))))
```

```{r B12_14}
"funRel" <-
function(x=leafshape$logpet, y=leafshape$loglen, scale=c(1,1)){
  ## Find principal components rotation; see Subsection 9.1.2
  ## Here (unlike 9.1.2) the interest is in the final component
  xy.prc <- prcomp(cbind(x,y), scale=scale)
  b <- xy.prc$rotation[,2]/scale
  c(bxy = -b[1]/b[2])     # slope of functional equation line
}
## Try the following:
leafshape <- DAAG::leafshape
funRel(scale=c(1,1))    # Take x and y errors as equally important
  # Note that all lines pass through (mean(x), mean(y))
```

```{r B12_15}
P <- rbind(
    c(1 , 0 , 0 , 0 , 0 , 0),
    c(.5, 0 , .5, 0 , 0 , 0),
    c(0 , .5, 0 , .5, 0 , 0),
    c(0 , 0 , .5, 0 , .5, 0),
    c(0 , 0 , 0 , .5, 0 , .5),
    c(0 , 0 , 0 , 0 , 0 , 1))
dimnames(P) <- list(0:5,0:5)
P
```

```{r B12_15x}
Markov <- function(N=15, initial.value=1, transition=P, stopval=NULL)
  {X <- numeric(N)
   X[1] <- initial.value + 1  # States 0:(n-1); subscripts 1:n
   n <- nrow(transition)
   for (i in 2:N){
    X[i] <- sample(1:n, size=1, prob=transition[X[i-1], ])
    if(length(stopval)>0)if(X[i] %in% (stopval+1)){X <- X[1:i]; break}}
  X - 1
}
 # Set `stopval=c(0,5)` to stop when  the player's fortune is $0 or $5
```

```{r B12_16}
Pb <- rbind(
  Sun = c(Sun=0.6, Cloud=0.2, Rain=0.2),
  Cloud= c(0.2, 0.4, 0.4),
  Rain= c(0.4, 0.3, 0.3))
Pb
```

```{r B12_16b}
plotmarkov <-
  function(n=1000, width=101, start=0, transition=Pb, npanels=5){
    xc2 <- Markov(n, initial.value=start, transition)
    mav0 <- zoo::rollmean(as.integer(xc2==0), k=width)
    mav1 <- zoo::rollmean(as.integer(xc2==1), k=width)
    npanel <- cut(1:length(mav0), breaks=seq(from=1, to=length(mav0),
                  length=npanels+1), include.lowest=TRUE)
    df <- data.frame(av0=mav0, av1=mav1, x=1:length(mav0), gp=npanel)
    print(xyplot(av0+av1 ~ x | gp, data=df, layout=c(1,npanels), type="l",
          par.strip.text=list(cex=0.65), auto.key=list(columns=2),
          scales=list(x=list(relation="free"))))
}
```

```{r eval=T}
code <- knitr::knit_code$get()
txt <- paste0("\n## ", names(code),"\n", sapply(code, paste, collapse='\n'))
writeLines(txt, con="../inst/doc/ch2.R")
```
